---
title: "Legal hypergraphs: the paper workflow in honets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Legal hypergraphs: the paper workflow in honets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(honets)
```

Coupette, Hartung and Katz (2024) model legal relations without reducing a
tribunal, citation block, or other group event to unrelated dyads. This
vignette maps the paper's computational workflow onto honets. Hypergraph
construction and mathematics remain native to honets; once a projection is
an ordinary graph, cograph supplies graph algorithms. Dynet is not a package
dependency.

## Temporal tribunals

Wide, fixed-cardinality records can be read directly. An interval snapshot
contains tribunals active at the requested date; a cumulative snapshot keeps
every tribunal constituted by then.

```{r temporal}
tribunals <- data.frame(
  case = c("A", "B", "C"),
  president = c("p1", "p2", "p1"),
  arbitrator_1 = c("a1", "a1", "a4"),
  arbitrator_2 = c("a2", "a3", "a5"),
  constituted = c(1, 2, 4), concluded = c(4, 3, 6)
)

thg <- temporal_hypergraph(
  tribunals,
  member = c("president", "arbitrator_1", "arbitrator_2"),
  edge = "case", start = "constituted", end = "concluded",
  evolution = "interval"
)

active <- hypergraph_snapshot(thg, at = 2)
c(active = active$n_hyperedges,
  cumulative = hypergraph_snapshot(thg, at = 2,
                                   mode = "cumulative")$n_hyperedges)
```

`hypergraph_snapshots()` applies the same contract over all event times, and
`multiedges = FALSE` collapses repeated member sets while retaining their
multiplicity.

## Hyperedges, paths, and node centrality

The paper studies both vertices and hyperedges. `hypergraph_edges()` reports
cardinality, weight, incident hyperedges, and vertex neighbourhoods.
`hypergraph_edge_centrality()` constructs an unweighted s-line graph and
computes ordinary shortest-path centralities on its hyperedge-vertices.

```{r centrality}
hypergraph_edges(active)
hypergraph_edge_centrality(active, s = c(1, 2))
hypergraph_centrality(active, type = "subhypergraph")
```

## Association projections

For a hyperedge of size |e|, the association projection gives every
member pair weight `1 / (|e| - 1)`. Thus every edge contributes one unit of
incident weight around each member. Repeated hyperedges may count repeatedly
or be collapsed, reproducing the paper's multi- and binary-hypergraph
representations.

```{r projection}
association <- hypergraph_project(
  active, method = "association", duplicate_edges = "count",
  what = "matrix"
)
association
```

For citation-block hypergraphs, `self_association = TRUE` adds the source
decision to its targets. Sources can be stored by `temporal_hypergraph(source
= ...)` or supplied through `edge_source =`.

## Communities and their stability

`hypergraph_communities()` runs Infomap 50 times with 100 trials per seed by
default, builds AMI, ARI, and NMI similarity matrices, and chooses the run
with maximum summed AMI as the medoid. Small values are used below only to
keep the vignette quick.

```{r communities}
if (requireNamespace("igraph", quietly = TRUE)) {
  fit <- hypergraph_communities(active, n_runs = 3, trials = 3,
                                seeds = 1:3)
  as.data.frame(fit)
  hypergraph_community_quality(active, fit)
  as.data.frame(fit, what = "ami")
}
```

Independent partitions can be compared directly with
`hypergraph_agreement(method = c("ari", "ami", "nmi"))`.

## Y/T/O motifs and the configuration null

In a simple 3-uniform hypergraph, four vertices can induce two (Y), three
(T), or all four (O) possible triples. The null uses the same detailed
pairwise-edge reshuffle as HypergraphX 1.5 and preserves degrees and edge
cardinalities before duplicate collapse.

```{r motifs}
motif_data <- data.frame(
  member = c(1, 2, 3, 1, 2, 4, 1, 3, 4),
  event = rep(c("e1", "e2", "e3"), each = 3)
)
motif_hg <- group_hypergraph(motif_data, "member", "event")
hypergraph_motifs(motif_hg, what = "counts")
hypergraph_motifs(motif_hg, n = 19, seed = 1)
```

The implementation was checked against the authors' Zenodo archive
(10.5281/zenodo.8081507). Reconstructing their 441-node, 742-case ICSID
aggregate reproduces its observed motif census exactly: Y = 478, T = 7, and
O = 0.
