---
title: "Hypergraph analysis of a text corpus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hypergraph analysis of a text corpus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(honets)
```

A corpus of abstracts shares vocabulary unevenly: a word like *telescope*
binds a few documents tightly, while *challenges* touches half the corpus
weakly. A weighted hypergraph represents that structure directly — documents
are vertices and every word is a hyperedge over the documents it occurs in,
weighted by tf-idf. Two questions drive this vignette, asked of the bundled
`covid_abstracts` corpus (165 COVID-19 education-research abstracts,
2020–2024): do the abstracts organize into coherent groups purely through
shared vocabulary, and can a handful of labeled abstracts classify the rest?

The methods are Zhou, Huang & Schölkopf's (2006) hypergraph Laplacian and
Hayashi et al.'s (2020) random-walk Laplacian with edge-dependent vertex
weights, the spectral engines of honets' hypergraph family (HyperNetX-parity
verified), applied to a text hypergraph built by `text_hypergraph()`.

## From corpus to hypergraph

`text_hypergraph()` tokenizes deterministically in base R, counts
document–word occurrences, and hands the weighted incidence structure to
`group_hypergraph()`. Function words and a few
corpus-boilerplate terms are excluded, and words in fewer than three
abstracts are dropped — a word seen once cannot connect anything.

```{r construct}
stops <- c(
  stop_words_en(),
  "using", "used", "use", "based", "results", "study", "research",
  "paper", "article", "findings", "data"
)

hg <- text_hypergraph(
  covid_abstracts,
  column = "abstract",
  id = "doc",
  weight = "tfidf",
  stop_words = stops,
  min_count = 3L
)
hg
```

The corpus becomes 165 vertices connected by 1,453 word-hyperedges. The
median word touches 4 abstracts; the largest hyperedge spans 140 of the 165 —
a near-universal term that tf-idf weighting correspondingly down-weights.

```{r summary}
hg_measures(hg, what = "summary")
```

A pairwise participation of 0.996 says nearly every pair of abstracts shares
at least one word: the hypergraph is one dense component, so spectral methods
apply to the whole corpus at once.

## Which words hold the corpus together?

Reversing the orientation makes words the vertices — two words are close
when they co-occur in many documents — and centrality then ranks the
corpus's organizing vocabulary.

```{r words}
word_hg <- text_hypergraph(
  covid_abstracts,
  column = "abstract",
  id = "doc",
  nodes = "word",
  weight = "tfidf",
  stop_words = stops,
  min_count = 3L
)
hg_centrality(word_hg, type = "clique", sort_by = "clique", n = 10)
```

The ranking is the corpus in miniature: *covid*, *pandemic*, *education*,
*students*, *learning*, *online*, *teaching*. This is remote-education
research, and the centrality ranking recovers that without reading a single
abstract.

## Do the abstracts cluster?

`hg_cluster()` with `type = "random_walk"` is the Hayashi et al. (2020)
pipeline: the tf-idf weights become edge-dependent vertex weights of a random
walk, whose normalized Laplacian is embedded and clustered. The number of
clusters is explicit by design; four is the working choice here and is
checked below.

```{r cluster}
clusters <- hg_cluster(hg, k = 4, type = "random_walk", seed = 1)
aggregate(node ~ cluster, data = clusters, FUN = length)
```

```{r cluster-years}
merged <- merge(clusters, covid_abstracts, by.x = "node", by.y = "doc")
aggregate(year ~ cluster, data = merged, FUN = mean)
```

The partition is 16/44/29/76, and every cluster's mean publication year sits
between 2021.1 and 2021.7: the clusters are thematic, not temporal. The
vocabulary of this literature separates by topic while staying stable across
the pandemic years.

Is `k = 4` doing real work? Refitting at `k = 3` answers the sensitivity
question:

```{r sensitivity}
clusters_3 <- hg_cluster(hg, k = 3, type = "random_walk", seed = 1)
hg_agreement(clusters, clusters_3, what = "table")
```

The two compact themes survive the change almost intact (15 of 16, and all
29, stay together), while the largest cluster redistributes. The small
clusters are stable structure worth interpreting; the large one is the
corpus's residual mass, and conclusions should not lean on its boundary.

## Classifying from six labels

Transductive label spreading (Zhou et al. 2006) takes labels for a few
vertices and propagates them over the hyperedges. Labeling the three
alphabetically first abstracts of 2020 as `"early"` and of 2024 as `"late"`
asks whether pandemic-era vocabulary dates an abstract:

```{r classify}
labels <- c(
  "2-s2.0-85085897904" = "early",
  "2-s2.0-85087124472" = "early",
  "2-s2.0-85087646998" = "early",
  "2-s2.0-85107489479" = "late",
  "2-s2.0-85112557152" = "late",
  "2-s2.0-85113355756" = "late"
)
predictions <- hg_classify(hg, labels = labels)
predicted_years <- merge(predictions, covid_abstracts,
                         by.x = "node", by.y = "doc")
aggregate(year ~ predicted, data = predicted_years, FUN = mean)
```

```{r classify-check}
hg_agreement(predictions, covid_abstracts,
             node = c("node", "doc"), label = c("predicted", "year"),
             what = "table")
```

The classifier is conservative and correct where it commits: every abstract
it calls `"early"` is genuinely from 2020, but it finds only three of them —
the rest of the corpus, 2020 included, reads as `"late"`. The decisive
reading: this literature's vocabulary converged almost immediately, so
publication year is weakly encoded in word choice. Label spreading here is
trustworthy for what it asserts and honest about how little the structure
supports.

## Ranking, dualizing, and testing the structure

Three verbs complete the analysis layer. `hg_pagerank()` ranks vertices by
the stationary importance of the Chitra & Raphael (2019) random walk with
edge-dependent vertex weights -- the walk that actually uses the tf-idf
weights (verified against the HyperNetX reference implementation at
machine precision):

```{r pagerank}
hg_pagerank(hg, sort_by = "pagerank", n = 3)
```

The top-ranked abstracts are the corpus's connective tissue -- documents
whose vocabulary reaches everywhere. A `personalized =` vector turns the
same verb into "importance from the perspective of these documents".
`dual_hypergraph()` swaps the vertex and hyperedge roles of a fitted
object, so the word-level view never requires re-tokenizing:

```{r dual}
dual_hypergraph(hg)
```

Finally, `hg_null_test()` asks whether observed structure exceeds what the
degree sequences alone imply: it rewires the binary membership by
degree-preserving checkerboard swaps (Gotelli 2000) and compares. On the
2020 abstracts:

```{r null-test}
abstracts_2020 <- text_hypergraph(
  subset(covid_abstracts, year == 2020),
  column = "abstract",
  id = "doc",
  weight = "tfidf",
  stop_words = stops,
  min_count = 3L
)
hg_null_test(
  abstracts_2020,
  statistic = c("pairwise_participation", "avg_jaccard"),
  n = 99,
  seed = 1
)
```

Read the two rows together: near-universal pairwise participation (0.988)
is *not* evidence of structure -- the null reproduces it (p = 0.36),
because vocabulary-rich abstracts connect to everything under any
degree-preserving rewiring. The mean edge Jaccard *is* significant
(z = 2.48, p = 0.01): words repeat each other's document combinations more
than their frequencies require, which is exactly the topical co-occurrence
the clustering exploits. The decisive practice: never report a structural
statistic without its null interval.

## Using an existing quanteda or tidytext pipeline

`text_hypergraph()` needs no NLP dependency, but any long document–word
table is already the incidence structure, so existing pipelines plug into
the same engines directly:

```{r bridge, eval = FALSE}
# quanteda: tidytext::tidy() turns a dfm into a long table with columns
# document, term, count (verified against quanteda 4.x / tidytext 0.4.x)
group_hypergraph(
  tidytext::tidy(my_dfm),
  actor = "term", group = "document", weight = "count"
)

# tidytext: word counts are already long (document, word, n)
group_hypergraph(
  my_word_counts,
  actor = "word", group = "document", weight = "n"
)
```

## When to use which

- **`hg_cluster()`** — no labels, discover thematic groups; prefer
  `type = "random_walk"` whenever the weights are tf-idf (edge-dependent
  vertex weights are exactly its assumption).
- **`hg_classify()`** — a few trusted labels, classify the rest; read
  `margin` as the confidence and expect conservative behavior when the
  vocabulary signal is weak.
- **`hg_centrality()`** — rank organizing vocabulary (words as nodes) or
  hub documents (documents as nodes).
- **`hg_measures()`** — the structural facts (sizes, overlaps, density)
  that say whether spectral analysis is even applicable; check
  connectivity-adjacent quantities before clustering.

## References

Hayashi, K., Aksoy, S. G., Park, C. H., & Park, H. (2020). Hypergraph random
walks, Laplacians, and clustering. *CIKM 2020*.
doi:10.1145/3340531.3412034.

Zhou, D., Huang, J., & Schölkopf, B. (2006). Learning with hypergraphs:
Clustering, classification, and embedding. *NeurIPS 2006*.
