---
title: "Windowed and embedding hypergraphs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Windowed and embedding hypergraphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(honets)
```

The bag-of-words hypergraph of `vignette("text-hypergraphs")` ignores two
things a corpus knows: the *order* of words within a document and the
*position* of documents in an embedding space. Two further constructions use
them. The question for each is the same: does the added structure change a
conclusion the bag construction would have reached?

## Windowed hyperedges: word order as structure

`construction = "window"` slides a token window over each document (never
crossing document boundaries); each distinct window content — the set of
words co-occurring within `window` tokens of each other — is one hyperedge,
weighted by its window count. This is the sequential construction of
Ding et al. (2020, Sec. 3.2), extended with weights — the text front end of
`window_hypergraph()`, with which it shares its incidence matrix (a shipped
package test).

The 2020 abstracts make a compact testbed:

```{r window}
stops <- c(
  stop_words_en(),
  "using", "used", "use", "based", "results", "study", "research",
  "paper", "article", "findings", "data"
)
early <- subset(covid_abstracts, year == 2020)

windowed <- text_hypergraph(
  early,
  column = "abstract",
  id = "doc",
  construction = "window",
  window = 3,
  stop_words = stops,
  min_count = 3L
)
windowed
```

```{r window-centrality}
hg_centrality(windowed, type = "clique", sort_by = "clique", n = 10)
```

Compare the same corpus as a bag-of-words hypergraph over words:

```{r bag-centrality}
bag_words <- text_hypergraph(
  early,
  column = "abstract",
  id = "doc",
  nodes = "word",
  weight = "tfidf",
  stop_words = stops,
  min_count = 3L
)
hg_centrality(bag_words, type = "clique", sort_by = "clique", n = 10)
```

The rankings agree on the theme words — but the windowed construction puts
**distance** third, and the bag construction does not rank it in the top ten
at all. Adjacency is doing real work: *distance* matters in this corpus
almost exclusively inside the phrases *distance learning* and *distance
education*, which a window sees and a bag cannot. The window size is an
arbitrary choice, so it gets a sensitivity check:

```{r window-sensitivity}
windowed_4 <- text_hypergraph(
  early,
  column = "abstract",
  id = "doc",
  construction = "window",
  window = 4,
  stop_words = stops,
  min_count = 3L
)
hg_centrality(windowed_4, type = "clique", sort_by = "clique", n = 10)
```

At `window = 4`, *distance* rises to second. The conclusion — adjacency
promotes phrase vocabulary the bag misses — survives the window choice.

## kNN embedding hyperedges: semantic space as structure

`construction = "knn"` places each document in an embedding space and forms
one hyperedge per document: itself plus its `k` nearest neighbors, weighted
by cosine similarity. The bundled `covid_embeddings` matrix holds sbert
embeddings of the `covid_abstracts` corpus (pinned `all-MiniLM-L6-v2`,
L2-normalized, built by `data-raw/covid_embeddings.R`), so the pipeline runs
offline; with the sbert package installed, omitting `embeddings` encodes the
text directly.

```{r knn}
knn_hg <- text_hypergraph(
  covid_abstracts,
  column = "abstract",
  id = "doc",
  construction = "knn",
  k = 10,
  embeddings = covid_embeddings
)
knn_hg
```

```{r knn-cluster}
knn_clusters <- hg_cluster(knn_hg, k = 4, type = "random_walk", seed = 1)
aggregate(node ~ cluster, data = knn_clusters, FUN = length)
```

The embedding hypergraph partitions the corpus into balanced groups
(41/41/37/46), where the bag-of-words clustering of
`vignette("text-hypergraphs")` produced 16/44/29/76 — one dominant residual
cluster. Semantic neighborhoods spread structure more evenly than shared
vocabulary does. The two views agree only partially:

```{r knn-vs-bag}
bag <- text_hypergraph(
  covid_abstracts,
  column = "abstract",
  id = "doc",
  weight = "tfidf",
  stop_words = stops,
  min_count = 3L
)
bag_clusters <- hg_cluster(bag, k = 4, type = "random_walk", seed = 1)
hg_agreement(knn_clusters, bag_clusters, what = "table")
```

Two themes are found by both constructions (32 and 37 documents stay
together); the rest reshuffles. Neither view is "the" truth — they answer
different questions (shared vocabulary vs. semantic proximity) — but a
grouping stable across both is worth trusting.

`k` is the consequential arbitrary choice here, so it gets the sensitivity
check:

```{r knn-sensitivity}
knn_5_hg <- text_hypergraph(covid_abstracts, column = "abstract", id = "doc",
                            construction = "knn", k = 5,
                            embeddings = covid_embeddings)
knn_5 <- hg_cluster(knn_5_hg, k = 4, type = "random_walk", seed = 1)
knn_15_hg <- text_hypergraph(covid_abstracts, column = "abstract", id = "doc",
                             construction = "knn", k = 15,
                             embeddings = covid_embeddings)
knn_15 <- hg_cluster(knn_15_hg, k = 4, type = "random_walk", seed = 1)
hg_agreement(knn_clusters, knn_15)
hg_agreement(knn_clusters, knn_5)
```

The `aligned` column counts the documents that stay with the majority of
their cluster. Against `k = 15`, 143 of 165 documents keep their cluster's
majority alignment; against `k = 5`, 125 do. The partition is robust upward and
softer downward — small neighborhoods fragment one of the four groups.
Report conclusions at the k where they are stable, and say so.

One honesty note: everything in this section is conditional on the encoder.
`covid_embeddings` comes from one pinned model; a different encoder yields a
different geometry, and any claim that matters should be re-checked with a
second model (`text_hypergraph(construction = "knn", model = )` makes that a
one-argument change when sbert is installed).

## When to use which construction

- **`"bag"`** — the default; shared vocabulary, interpretable hyperedges
  (words or documents), tf-idf weighting, the full clustering/transduction
  story of `vignette("text-hypergraphs")`.
- **`"window"`** — when word order carries the signal: phrase vocabulary,
  collocations, sequence-like corpora. Vertices are words; use it with
  `hg_centrality()` and `hg_measures()`. Prefer modest corpora or
  `min_count` filtering — distinct windows multiply quickly.
- **`"knn"`** — when meaning matters more than vocabulary overlap:
  paraphrase-heavy or multilingual-adjacent corpora, or as the explicit,
  spectral alternative to UMAP+HDBSCAN pipelines. Vertices are documents;
  every hyperedge has `k + 1` members; the geometry is the encoder's.

## References

Ding, K., Wang, J., Li, J., Li, D., & Liu, H. (2020). Be more with less:
Hypergraph attention networks for inductive text classification.
*EMNLP 2020*. doi:10.18653/v1/2020.emnlp-main.399.

Hayashi, K., Aksoy, S. G., Park, C. H., & Park, H. (2020). Hypergraph random
walks, Laplacians, and clustering. *CIKM 2020*. doi:10.1145/3340531.3412034.
