---
title: "Mixed-membership topics of the COVID-19 education literature"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-membership topics of the COVID-19 education literature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7, fig.height = 5,
                      out.width = "100%")
library(hypernets)
```

A topic model describes a corpus by a small number of themes. Each theme, or
topic, is a probability distribution over the words of the vocabulary, and
each document is a mixture of topics in proportions of its own. An abstract on
the online teaching of medical students draws on a topic about online teaching
and on a topic about medical education, and it belongs to both in the
proportions its words indicate. A partition, in which every document belongs
to exactly one topic, is the special case of documents that are each about
one thing. The shares of a mixed-membership model show how close a corpus
comes to that case.

## The corpus

`covid_sample` holds a random sample of 1,000 abstracts of research on
COVID-19 and education published from 2020 to 2024. `clean_text()` removes
what a bibliographic export adds to an abstract, such as HTML, citation
numbers in brackets, URLs and copyright notices. The stop list is
`stop_words_en()` together with the words that name the subject of every
abstract, since a word that occurs in every document separates none of them.

```{r corpus}
abstracts <- clean_text(covid_sample, column = "abstract")
boilerplate <- c("covid", "coronavirus", "sars", "cov", "pandemic",
                 "disease", "study", "studies", "result", "results",
                 "conclusion", "conclusions", "background", "method",
                 "methods", "objective", "aim")
stops <- c(stop_words_en(), boilerplate)
```

## The hypergraph

Each abstract is a node, and each word that occurs at least five times in the
corpus is a hyperedge that binds the abstracts using it. The incidence of an
abstract and a word is the count of the word in the abstract. The incidence
matrix is stored dense (`sparse = FALSE`), which the graded memberships at the
end require.

```{r hypergraph}
abstracts_hg <- text_hypergraph(abstracts, column = "abstract", id = "doc",
                                stop_words = stops, min_count = 5,
                                min_chars = 2, sparse = FALSE)
abstracts_hg
```

```{r hypergraph-values, include = FALSE}
vocabulary <- hg_get(abstracts_hg, what = "vocabulary")
```

The hypergraph has `r abstracts_hg$n_nodes` abstracts and
`r format(nrow(vocabulary), big.mark = ",")` words.

## How many topics

A topic model has as many topics as it is given, and no statistic identifies
the correct number. Two properties of the topics guide the choice. Semantic
coherence measures how often the most probable words of a topic occur in the
same documents (Mimno et al. 2011). For every pair of top words it takes the
logarithm of the number of documents containing both, plus one, divided by
the number of documents containing the more probable word, and it sums these
values over the pairs. Values closer to zero are more coherent. Exclusivity
measures how far the top words of a topic belong to that topic rather than to
the others. FREX combines, for every top word, the rank of its probability
within the topic and the rank of its share of the probability it has across
all topics, and sums these over the top words (Bischof & Airoldi 2012).
Models with fewer topics have broader, more coherent topics, and models with
more topics have narrower, more exclusive ones. A number of topics is worth
considering when no other number gives both a higher coherence and a higher
exclusivity, and these numbers form the frontier of the two measures (Roberts
et al. 2014).

`hg_topic_search()` fits the topic model of the next section for 4, 8, 12, 16
and 20 topics, each from two random starts, and reports the mean coherence
and the mean exclusivity of the ten most probable words of the topics.

```{r topic-search}
topic_search <- hg_topic_search(abstracts_hg, k = seq(4, 20, by = 4),
                                nstart = 2, parallel = TRUE, n_cores = 2)
topic_search
```

```{r topic-search-plot}
plot(topic_search)
```

```{r topic-search-values, include = FALSE}
k4 <- subset(topic_search, k == 4)
k16 <- subset(topic_search, k == 16)
k20 <- subset(topic_search, k == 20)
on_frontier <- subset(topic_search, frontier)
frontier_k <- sub(", ([^,]*)$", " and \\1", toString(on_frontier$k))
```

Mean coherence falls from `r round(k4$coherence, 1)` with 4 topics to
`r round(k16$coherence, 1)` with 16 and `r round(k20$coherence, 1)` with 20,
and mean exclusivity rises from `r round(k4$exclusivity, 2)` to
`r round(k16$exclusivity, 2)` and `r round(k20$exclusivity, 2)`. The models
with `r frontier_k` topics lie on the frontier.
Sixteen topics is the number of the published structural topic model of this
literature (Saqr et al. 2023), and the model below uses it.

## A mixed-membership model of 16 topics

`hg_topics()` approximates the matrix of word counts, with one row per
abstract and one column per word, by the product of two non-negative
matrices. The first gives every abstract a weight on every topic, and the
second gives every topic a weight on every word. The fit minimises the
Kullback-Leibler divergence between the counts and their approximation by the
multiplicative updates of Lee and Seung (2001). With this divergence the
factorization is the maximum-likelihood fit of probabilistic latent semantic
analysis (Hofmann 1999; Gaussier & Goutte 2005). After normalisation, the
weights of an abstract are its shares of the topics, which sum to one, and
the weights of a topic are the probabilities of its words.

The fit depends on its random start. It is run from five starts, and the
start with the smallest divergence is kept. The agreement of a topic is the
average Jaccard coefficient of its first one, two, and up to ten most probable
words with those of the matching topic of another start, the topics of two
starts being paired by the Hungarian method (Kuhn 1955; Greene et al. 2014).
A topic whose agreement is close to one is found again from every start.

```{r topic-model}
topic_model <- hg_topics(abstracts_hg, k = 16, nstart = 5, parallel = TRUE,
                         n_cores = 2)
topic_model
```

```{r topic-model-values, include = FALSE}
topic_table <- hg_get(topic_model)
largest <- subset(topic_table, prevalence == max(prevalence))
smallest <- subset(topic_table, prevalence == min(prevalence))
```

The prevalence of a topic is its mean share over the abstracts, and its
documents column is the sum of those shares, the expected number of abstracts
it accounts for. The largest topic, led by `r largest$top_words`, accounts for
`r round(100 * largest$prevalence, 1)`% of the corpus, and the smallest, led by
`r smallest$top_words`, for `r round(100 * smallest$prevalence, 1)`%. The mean
agreement of the topics across the starts is
`r round(mean(topic_table$agreement), 2)`. Topics with a low agreement are one
of several divisions of their part of the vocabulary that the data support
about equally well.

```{r topic-model-plot, fig.height = 9}
plot(topic_model)
```

## How mixed the abstracts are

The largest share of an abstract identifies its main topic.

```{r main-topics}
main_topics <- hg_get(topic_model, what = "documents")
head(main_topics)
```

```{r main-topic-values, include = FALSE}
main_share <- main_topics$share
share_quartiles <- quantile(main_share, c(0.25, 0.5, 0.75))
```

In the median abstract the main topic has a share of
`r round(share_quartiles[2], 2)`, and half the abstracts lie between
`r round(share_quartiles[1], 2)` and `r round(share_quartiles[3], 2)`.
`r round(100 * mean(main_share >= 0.9), 1)`% of the abstracts have a main
topic with a share of 0.9 or more, and `r round(100 * mean(main_share < 0.5), 1)`%
have no topic with a share above one half. Most abstracts combine topics.

The shares of a single abstract show the mixture.

```{r example-shares}
shares <- hg_get(topic_model, what = "shares")
example_shares <- subset(shares, node == "2-s2.0-85101300775" & share >= 0.05)
example_shares
```

```{r example-text}
example_abstract <- subset(abstracts, doc == "2-s2.0-85101300775")
with(example_abstract, writeLines(strwrap(abstract, 78)))
```

```{r example-values, include = FALSE}
ranked <- example_shares[order(-example_shares$share), ]
words_of <- function(label) subset(topic_table, topic == label)$top_words
```

The abstract describes near-peer teaching in a medical course moved online.
It takes a share of `r round(ranked$share[1], 2)` of the topic led by
`r words_of(ranked$topic[1])` and `r round(ranked$share[2], 2)` of the topic
led by `r words_of(ranked$topic[2])`. A partition would assign it to one of
the two and lose the other half of what it is about.

## Topic quality

`hg_topic_quality()` scores every topic by the same coherence and
exclusivity. The size of a topic is its expected number of abstracts.

```{r model-quality}
model_quality <- hg_topic_quality(abstracts_hg, topics = topic_model)
model_quality
```

```{r model-quality-plot}
plot(model_quality)
```

## A partition of the abstracts

A partition is appropriate when the abstracts are each about one topic. It is
useful when they have to be drawn or compared as groups, and the hypergraph
can then be divided directly. `hg_cluster()` places the abstracts by the
eigenvectors of the 16 smallest eigenvalues of the hypergraph Laplacian of
Zhou et al. (2006) and divides them by k-means.

```{r clusters}
clusters <- hg_cluster(abstracts_hg, k = 16, seed = 1)
cluster_quality <- hg_topic_quality(abstracts_hg, clusters = clusters)
cluster_quality
```

```{r cluster-agreement}
hg_agreement(clusters, main_topics, label = c("cluster", "topic"),
             method = c("ari", "nmi"))
```

```{r cluster-values, include = FALSE}
cluster_agreement <- hg_agreement(clusters, main_topics,
                                  label = c("cluster", "topic"),
                                  method = c("ari", "nmi"))
```

The clusters have a mean coherence of
`r round(mean(cluster_quality$coherence), 1)` and a mean exclusivity of
`r round(mean(cluster_quality$exclusivity), 2)`, against
`r round(mean(model_quality$coherence), 1)` and
`r round(mean(model_quality$exclusivity), 2)` for the topics of the mixed
model, so the clusters are more coherent and less exclusive. The adjusted Rand
index between the clusters and the main topics is
`r round(cluster_agreement$ari, 2)`. A partition has to place every abstract
that lies between topics on one side, and the two divisions agree only in
part.

The clustering can itself be given graded memberships. The symmetric
non-negative factorization of the similarity between abstracts that the
hypergraph Laplacian defines (Kuang et al. 2012; Hayashi et al. 2020) gives
every abstract a non-negative weight on every cluster, normalised to sum to
one. These memberships describe how strongly the position of an abstract in
the hypergraph loads on each cluster, and they do not model its words.

```{r symnmf}
cluster_membership <- hg_cluster(abstracts_hg, k = 16, algorithm = "symnmf",
                                 seed = 1, nstart = 5, what = "membership")
subset(cluster_membership, node == "2-s2.0-85101300775" & membership >= 0.05)
```

```{r symnmf-values, include = FALSE}
cluster_main <- tapply(cluster_membership$membership,
                       cluster_membership$node, max)
```

The main cluster of an abstract has a membership of
`r round(median(cluster_main), 2)` in the median abstract, and
`r round(100 * mean(cluster_main >= 0.9), 1)`% of the abstracts have a main
cluster with a membership of 0.9 or more. The graded memberships of the
clustering also place most abstracts between clusters.

## References

Bischof, J. M., & Airoldi, E. M. (2012). Summarizing topical content with word
frequency and exclusivity. *Proceedings of the 29th International Conference
on Machine Learning*, 201-208.

Gaussier, E., & Goutte, C. (2005). Relation between PLSA and NMF and
implications. *Proceedings of SIGIR 2005*, 601-602.
https://doi.org/10.1145/1076034.1076148

Greene, D., O'Callaghan, D., & Cunningham, P. (2014). How many topics?
Stability analysis for topic models. *Machine Learning and Knowledge Discovery
in Databases*, 498-513. https://doi.org/10.1007/978-3-662-44848-9_32

Hayashi, K., Aksoy, S. G., Park, C. H., & Park, H. (2020). Hypergraph random
walks, Laplacians, and clustering. *Proceedings of CIKM 2020*, 495-504.
https://doi.org/10.1145/3340531.3412034

Hofmann, T. (1999). Probabilistic latent semantic indexing. *Proceedings of
SIGIR 1999*, 50-57. https://doi.org/10.1145/312624.312649

Kuang, D., Ding, C., & Park, H. (2012). Symmetric nonnegative matrix
factorization for graph clustering. *Proceedings of the 2012 SIAM
International Conference on Data Mining*, 106-117.
https://doi.org/10.1137/1.9781611972825.10

Kuhn, H. W. (1955). The Hungarian method for the assignment problem. *Naval
Research Logistics Quarterly*, 2, 83-97. https://doi.org/10.1002/nav.3800020109

Lee, D. D., & Seung, H. S. (2001). Algorithms for non-negative matrix
factorization. *Advances in Neural Information Processing Systems*, 13,
556-562.

Mimno, D., Wallach, H. M., Talley, E., Leenders, M., & McCallum, A. (2011).
Optimizing semantic coherence in topic models. *Proceedings of EMNLP 2011*,
262-272.

Roberts, M. E., Stewart, B. M., Tingley, D., Lucas, C., Leder-Luis, J.,
Gadarian, S. K., Albertson, B., & Rand, D. G. (2014). Structural topic models
for open-ended survey responses. *American Journal of Political Science*,
58(4), 1064-1082. https://doi.org/10.1111/ajps.12103

Saqr, M., Raspopovic Milic, M., Pancheva, K., Jovic, J., Peltekova, E. V., &
Conde, M. Á. (2023). A multimethod synthesis of Covid-19 education research:
The tightrope between covidization and meaningfulness. *Universal Access in
the Information Society*, 23(3), 1163-1176.
https://doi.org/10.1007/s10209-023-00989-w

Zhou, D., Huang, J., & Schölkopf, B. (2006). Learning with hypergraphs:
Clustering, classification, and embedding. *Advances in Neural Information
Processing Systems*, 19, 1601-1608.
