---
title: "The covid_abstracts corpus, end to end"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 5, cache = FALSE)
library(hypernets)
okabe <- c("#E69F00", "#56B4E9", "#009E73", "#CC79A7",
           "#0072B2", "#D55E00", "#F0E442", "#999999")
```

What is in the 165 COVID-era education abstracts bundled with
hypernets -- which documents and words carry the corpus, how many
topics it contains, what those topics are, and whether its structure is
distinguishable from chance? Every step is one package verb; every claim
in the prose restates a number computed in the chunk above it.

## The corpus

```{r}
data(covid_abstracts)
str(covid_abstracts)
xtabs(~ year, data = covid_abstracts)
```

165 abstracts on education during and after COVID-19, with a Scopus id
(`doc`), title, abstract text and year. Coverage is 40 abstracts per year
for 2020--2023 and 5 from 2024: a balanced four-year window with a small
tail.

```{r years}
barplot(xtabs(~ year, data = covid_abstracts), col = okabe[5],
        border = NA, xlab = "Publication year", ylab = "Abstracts",
        main = "Corpus coverage by year")
```

## From text to hypergraph

The corpus becomes a document--word hypergraph: each abstract is a vertex,
each word is a hyperedge joining every abstract that uses it, and the
incidence weight is smoothed tf-idf, so a word counts for more in an
abstract where it is frequent and rare elsewhere. English stop words are
removed -- with them left in, "the" and "of" dominate every downstream
table (verified; they swamp the keyword lists).

```{r}
hg <- text_hypergraph(covid_abstracts, column = "abstract", id = "doc",
                      weight = "tfidf", stop_words = stop_words_en())
hg
```

165 documents share 4,043 distinct words. Hyperedge sizes run from 1 (a
word used in a single abstract) to 140 (a word used in almost every one),
with median 1: most vocabulary is idiosyncratic, and a small core of
words does the connecting.

```{r structure}
edges <- hg_measures(hg, what = "edges")
nodes <- hg_measures(hg, what = "nodes")
hg_measures(hg, what = "summary")
```

```{r degree-plots, fig.height = 4}
par(mfrow = c(1, 2))
hist(edges$size, breaks = 50, col = okabe[5], border = "white",
     xlab = "Hyperedge size (documents per word)", main = "Words")
hist(nodes$hyperdegree, breaks = 20, col = okabe[1], border = "white",
     xlab = "Hyperdegree (distinct words per document)", main = "Documents")
par(mfrow = c(1, 1))
```

Density 0.021 with pairwise participation 0.997: almost every pair of
abstracts shares at least one word, so the corpus is one tightly connected
component -- the regime where spectral methods are well defined. Documents
use between roughly 50 and 200 distinct words each; the word-size
distribution is the long-tailed one typical of text.

## Which documents carry the corpus

The Chitra & Raphael (2019) hypergraph random walk with edge-dependent
vertex weights ranks documents by their stationary visiting probability:
a document ranks high when it shares weighty vocabulary with many other
well-connected documents.

```{r}
ranked <- hg_pagerank(hg, sort_by = "pagerank", n = 5)
subset(merge(ranked, covid_abstracts, by.x = "node", by.y = "doc",
             sort = FALSE),
       select = c(pagerank, year, title))
```

The top document (PageRank 0.0160, 2022) is "Study of Resilience in
Learning Environments During the Covid-19 Pandemic" -- a synthesis paper
whose vocabulary bridges the corpus. The rest of the top five are the
same kind of connector: online peer support, distance-learning
perceptions, home physical education, digital library services. PageRank
here reads as "most representative of the shared conversation", not "most
cited".

The tensor centralities (clique-expansion eigenvector, Z- and
H-eigenvector) rank by a different mechanism -- nonlinear eigenvectors of
the adjacency tensor rather than a random walk -- so agreement between
them and PageRank is a convergence check, not a repetition.

```{r}
central <- hg_centrality(hg, sort_by = "Z", n = 5)
subset(merge(central, covid_abstracts, by.x = "node", by.y = "doc",
             sort = FALSE),
       select = c(clique, Z, H, title))
```

The Z-eigenvector ranking puts the same resilience study first: two
unrelated notions of importance agree on the corpus anchor, which makes
the "most representative document" claim robust rather than
method-dependent.

A personalized walk restarts at a chosen document and ranks the rest by
proximity to it -- the document's neighborhood rather than the global
core.

```{r}
anchor <- head(hg_pagerank(hg, sort_by = "pagerank", n = 1), 1)$node
neighborhood <- hg_pagerank(hg, personalized = anchor,
                            sort_by = "pagerank", n = 6)
subset(merge(neighborhood, covid_abstracts, by.x = "node", by.y = "doc",
             sort = FALSE),
       select = c(pagerank, title))
```

Around the resilience anchor the walk gathers self-directed learning,
educational continuity in segregated schools, learning-from-home
profiles, and institutional trust -- the resilience conversation
specifically, not the corpus-wide online-teaching core.

## Which words carry the corpus

The dual hypergraph swaps roles -- words become vertices, documents become
hyperedges -- so the same walk ranks vocabulary.

```{r}
dual <- dual_hypergraph(hg)
top_words <- hg_pagerank(dual, sort_by = "pagerank", n = 12)
top_words
```

```{r word-rank, fig.height = 5}
barplot(rev(top_words$pagerank), names.arg = rev(top_words$node),
        horiz = TRUE, las = 1, col = okabe[3], border = NA,
        xlab = "PageRank in the dual hypergraph",
        main = "The corpus core vocabulary")
```

The core is exactly what the corpus is about: *education*, *learning*,
*students*, *online*, *teaching*, then *covid*, *teachers*, *pandemic*.
No surprises is the correct outcome here -- the walk recovers the shared
frame, and the interesting structure is what remains after it (the
topics below).

## How many topics

The Hayashi et al. (2020) random-walk Laplacian uses the tf-idf incidence
directly (edge-dependent vertex weights), which is the natural spectral
operator for weighted text hypergraphs. Its eigenvalue gaps indicate how
many clusters the spectrum supports.

```{r scree, fig.height = 4.5}
spectrum <- head(hg_cluster(hg, k = 2, seed = 1, type = "random_walk",
                            what = "eigenvalues"), 20)
plot(spectrum$index, spectrum$value, type = "b", pch = 19, col = okabe[5],
     xlab = "Eigenvalue index", ylab = "Laplacian eigenvalue",
     main = "Spectrum of the random-walk Laplacian (first 20)")
abline(v = c(2, 4), lty = c(2, 3), col = okabe[6])
head(spectrum, 10)
```

The gap after index 1 is the trivial connected-component gap. The largest
non-trivial gap sits after index 2 (0.022): the coarsest reliable split is
two clusters. Beyond it the spectrum rises smoothly to a mild shoulder
around 8--9. The spectrum therefore supports anything from a coarse 2-way
to a fine ~9-way partition; we analyze k = 4 as the finest resolution
whose keyword sets stay distinct (sensitivity below), not because the
eigengap alone demands it.

## The topics

```{r}
topics <- hg_cluster(hg, k = 4, seed = 1, type = "random_walk")
xtabs(~ cluster, data = topics)
keywords <- hg_keywords(hg, topics, n = 8)
keywords
```

```{r keyword-panel, fig.height = 8}
par(mfrow = c(2, 2), mar = c(4, 7, 3, 1))
panels <- split(keywords, keywords$cluster)
invisible(Map(\(block, panel_col) {
  barplot(rev(block$score), names.arg = rev(block$word), horiz = TRUE,
          las = 1, col = panel_col, border = NA,
          main = unique(block$cluster), xlab = "tf-idf mass in cluster")
}, panels, head(okabe, length(panels))))
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
```

Reading the keyword panels, the four topics are:

- **Cluster 2 (80 abstracts) -- schools and teachers.** *learning,
  education, teachers, school*: the general remote-schooling literature,
  the corpus's center of mass.
- **Cluster 1 (44) -- online higher and clinical education.** *students,
  online, clinical, dental*: professional and university programs moving
  their practical teaching online.
- **Cluster 3 (27) -- systems and resilience.** *system, social,
  resilience, challenges*: the policy and systems view of educational
  disruption.
- **Cluster 4 (14) -- science and public trust.** *science, trust,
  reality, harms*: science education and the public-understanding strand.

The `share` column separates heavy from distinctive vocabulary: a word
with share near 1 (e.g. *dental*, *resilience*, *trust* in their
clusters) belongs almost entirely to that topic.

```{r sensitivity}
hg_keywords(hg, hg_cluster(hg, k = 3, seed = 1, type = "random_walk"),
            n = 6, collapse = TRUE)
hg_keywords(hg, hg_cluster(hg, k = 2, seed = 1, type = "random_walk"),
            n = 6, collapse = TRUE)
hg_stability(hg, k = 4, type = "random_walk")
```

Sensitivity: at k = 3 the science-and-trust topic is already separate and
the two education clusters merge their boundary; at k = 2 the split is
online-teaching versus everything else. The four-topic reading survives
the resolution choice, and `hg_stability()` confirms the k = 4 partition
is identical across k-means seeds (adjusted Rand 1).

## The topic map

The clustering's own spectral embedding places every abstract in the
plane; plotting it colored and shaped by topic shows how separated the
topics actually are.

```{r topic-map, fig.height = 6}
emb <- hg_cluster(hg, k = 4, seed = 1, type = "random_walk",
                  what = "embedding")
styles <- data.frame(cluster = sort(unique(emb$cluster)),
                     color = head(okabe, 4), pch = c(19, 17, 15, 18))
emb_styled <- merge(emb, styles, by = "cluster")
plot(emb_styled$dim2, emb_styled$dim3, col = emb_styled$color,
     pch = emb_styled$pch,
     xlab = "Spectral dimension 2", ylab = "Spectral dimension 3",
     main = "Documents in the Laplacian eigenspace")
legend("topright", legend = styles$cluster, col = styles$color,
       pch = styles$pch, bty = "n")
```

The map matches the sizes: the large schools cluster forms the dense
center, the online/clinical cluster separates along one axis, and the
two small topics (resilience; science and trust) sit at the periphery --
distinct but connected, which is why the corpus stays one component.

## Word communities, and an honest negative

The `nodes = "word"` orientation makes words the vertices and documents
the hyperedges, so the same clustering machinery can look for word
communities directly.

```{r}
whg <- text_hypergraph(covid_abstracts, column = "abstract", id = "doc",
                       weight = "tfidf", stop_words = stop_words_en(),
                       nodes = "word", sparse = TRUE)
whg
word_emb <- hg_cluster(whg, k = 4, seed = 1, what = "embedding")
xtabs(~ cluster, data = word_emb)
strata <- hg_seeds(word_emb, n = 8)
aggregate(word ~ cluster,
          data = data.frame(word = names(strata), cluster = strata),
          paste, collapse = ", ")
```

This one earns a negative interpretation: with hyperedges of size 36 to
161 (documents are large word sets), the word clusters are frequency
strata -- one 2,987-word background cluster and three bands of common
vocabulary -- not topics. On this corpus the informative direction is
documents-as-vertices with `hg_keywords()` on top, and reporting that is
more useful than pretending both directions always work.

## Order matters: the windowed construction

The bag construction discards word order. The windowed construction
(HyperGAT-style sequential hyperedges) keeps it: every sliding window of
`window` consecutive tokens becomes a hyperedge, so a word's degree
counts its distinct local contexts.

```{r}
anchor_text <- subset(covid_abstracts, doc == anchor)
win <- text_hypergraph(anchor_text, column = "abstract", id = "doc",
                       construction = "window", window = 3,
                       stop_words = stop_words_en())
win
head(hg_measures(win, what = "nodes"), 5)
```

The anchor abstract alone yields 161 word vertices in 252 distinct
3-token windows. This construction is the right input when sequence
carries meaning (collocations, phrase structure); for corpus-level topic
work the bag is the right grain, which is why the rest of this analysis
uses it.

## Does the semantic view agree?

The bag construction sees shared vocabulary. The kNN construction sees
shared meaning: each abstract's sbert embedding becomes a hyperedge
joining its ten nearest neighbors (`covid_embeddings` ships with the
package). Clustering both hypergraphs the same way asks whether lexical
and semantic topic structure coincide.

```{r}
data(covid_embeddings)
khg <- knn_hypergraph(covid_embeddings, k = 10)
sem_topics <- hg_cluster(khg, k = 4, seed = 1, type = "random_walk")
hg_agreement(topics, sem_topics, what = "table")
hg_agreement(topics, sem_topics)
```

Agreement is real but partial (the label names of two independent
clusterings are arbitrary, so the adjusted Rand index `ari` is the
statistic to read). The resilience topic maps almost entirely
onto one semantic cluster (15 of 27) and the schools topic onto another
(32 of 80), while the large lexical clusters split across semantic ones:
abstracts that use different vocabulary for the same theme are pulled
together by the embeddings, and vice versa. The two constructions are
complementary lenses, not redundant ones -- which is the reason the
package ships both.

## Is the structure real?

A degree-preserving null model (Gotelli 2000 checkerboard swaps) keeps
every document's word count and every word's document count while
shuffling who uses what. If the observed overlap structure survives in
the null, the "topics" would be a degree artifact.

```{r null}
null_fit <- hg_null_test(hg, n = 99, seed = 1)
null_fit
```

Two statistics are invariant under the swap null by construction --
density and average edge size preserve both margins, and their unchanged
values (z = NA) are the built-in falsification check that the chain
preserves what it must. Pairwise participation is indistinguishable from
the null (0.9973 observed vs 0.9974, z = -0.64, p = 0.66): nearly every
pair of abstracts shares a word whether or not topics exist, so that
statistic carries no topical signal in a corpus this dense. The signal is
in the average Jaccard overlap between words: 0.00873 observed against a
null of 0.00867 [0.00865, 0.00871], z = 3.32, p = 0.01. Words co-occur in
the same documents more than their frequencies alone require -- exactly
the excess the topic clustering above exploits. The structure is real.

## From topics to labels: spreading

With the topics named, transductive spreading (`hg_classify`) can label
the rest of the corpus from a handful of examples per topic.
`hg_seeds()` turns the clustering into those examples (the most
representative documents per topic by stationary probability), and
`hg_agreement()` scores each fit against the full partition. How many
examples are enough here?

```{r}
pool <- hg_cluster(hg, k = 4, seed = 1, type = "random_walk",
                   what = "embedding")
do.call(rbind, lapply(c(2, 5, 10), \(m)
  cbind(seeds_per_topic = m,
        hg_agreement(hg_classify(hg, labels = hg_seeds(pool, n = m),
                                 type = "random_walk",
                                 normalization = "class_mass"),
                     topics))))
```

With five labeled abstracts per topic, spreading reproduces 63% of the
full unsupervised partition; with two per topic it manages only 47% --
no better than always predicting the largest topic (80 of 165 is 48%) --
and with ten it reaches 70%. On a small, densely connected corpus, label
spreading needs a genuine seed base: a handful is not enough, and this
sensitivity belongs next to any spread labels a report presents.

## The neural tier on a small corpus

`hg_neural()` trains an HGNN (Feng et al. 2019) on the same hypergraph;
its propagation operator is exactly the spreading operator above, with
learned filters. Training is stochastic, so the sbert-feature run is
repeated over three torch seeds (house rule: a single-seed result is not
a finding). The same 20 seed labels as the five-per-topic spreading run.

```{r neural}
data(covid_embeddings)
seeds20 <- hg_seeds(pool, n = 5)
hg_agreement(hg_neural(hg, labels = seeds20, hidden = 32, epochs = 300,
                       seed = 1, validation = 0), topics)
sbert_runs <- do.call(rbind, lapply(1:3, \(s)
  cbind(method = "hg_neural (sbert)", torch_seed = s,
        hg_agreement(hg_neural(hg, labels = seeds20,
                               features = covid_embeddings, hidden = 32,
                               epochs = 300, seed = s, validation = 0),
                     topics))))
sbert_runs
aggregate(agreement ~ method, data = sbert_runs, mean)
```

The verdict is clear and worth stating plainly: with 20 labels on 165
documents, the neural classifier loses to closed-form spreading.
Bag-of-words features reach 31% agreement (a 130k-parameter first layer
cannot be estimated from 20 examples), sbert features average 52% with
a seed-to-seed range of 38-63%, and `hg_classify` delivered a stable 63%
with no training at all. This mirrors the package's full benchmark study
(see the pkgdown benchmarks article): the neural tier earns its keep
with thousands of labels (`hg_hypergat` R8 0.9665, R52 0.9433;
`hg_neural` R8 0.954), not dozens. `hg_hypergat` -- dual-attention
document-level hypergraphs, run on this same corpus and seed set in
the covid case study -- averages 45% over three torch seeds here, so
few-label work belongs to `hg_classify`.

## When to use which

For a corpus of this size and shape: `hg_pagerank` (global and
personalized, plus its dual) and `hg_centrality` for representative
documents and core vocabulary, with agreement between the walk and the
tensor rankings as the robustness check; `hg_cluster` with
`type = "random_walk"` plus `hg_keywords` for topics, always with an
eigengap look (`what = "eigenvalues"`) and an `hg_stability()` check before
trusting a partition; `knn_hypergraph` on sbert embeddings as the
semantic cross-check; the `nodes = "word"` orientation and the windowed
construction when word communities or word order are the question --
knowing that on a dense corpus the word orientation may only find
frequency strata; `hg_null_test` before claiming any of it is structure;
`hg_seeds` + `hg_classify` for few-label work, scored with
`hg_agreement`; and `hg_neural` only when labels
number in the hundreds or more. All of it is one connected pipeline on
objects built by `text_hypergraph()`.

## References

Chitra, U., & Raphael, B. (2019). Random walks on hypergraphs with
edge-dependent vertex weights. *ICML 36*.

Feng, Y., You, H., Zhang, Z., Ji, R., & Gao, Y. (2019). Hypergraph
neural networks. *AAAI 33*.

Gotelli, N. J. (2000). Null model analysis of species co-occurrence
patterns. *Ecology*, 81(9).

Hayashi, K., Aksoy, S., Park, C., & Park, H. (2020). Hypergraph random
walks, Laplacians, and clustering. *CIKM 29*.

Zhou, D., Huang, J., & Scholkopf, B. (2006). Learning with hypergraphs:
Clustering, classification, and embedding. *NeurIPS 19*.
