---
title: "COVID-19 education research: a hypergraph case study against a published synthesis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 5, cache = FALSE)
library(hypernets)
okabe <- c("#E69F00", "#56B4E9", "#009E73", "#CC79A7",
           "#0072B2", "#D55E00", "#F0E442", "#999999")
```

Saqr et al. (2024) synthesized COVID-19 education research from 4,201
Scopus articles with structural topic modeling (STM): they estimated 40
candidate models between 5 and 45 topics, ranked them by semantic
coherence and exclusivity, and had four experts adjudicate the final
solution -- sixteen topics, organized into three groups (education
related, distance/online education, and diverse issues), with *online
education* the most prevalent topic (14.6%). The corpus bundled with
this package, `covid_abstracts`, is a 165-abstract stratified sample
from a Scopus export of the same research field. That pairing sets up a
complete case study: run the package's entire pipeline on the sample --
construction, null test, centrality, several topic solutions, keyword
mapping, temporal drift, a semantic cross-check, and the full
classification cascade -- and test how much of the published structure
a 4% sample recovers without a human in the model-selection loop. Every
number in the prose is computed in the chunk above it.

## The corpus

```{r}
data(covid_abstracts)
str(covid_abstracts)
xtabs(~ year, data = covid_abstracts)
```

165 abstracts with Scopus id, title, text and year: 40 per year for
2020--2023 and 5 from 2024. The published study covered 2019--2021
(mostly 2021), so the sample extends past the paper's window -- a
difference to keep in mind when prevalences are compared below.

## Construction, and whether the structure is real

Each abstract becomes a vertex; each word becomes a hyperedge joining
every abstract that uses it, weighted by smoothed tf-idf; English stop
words are removed.

```{r}
hg <- text_hypergraph(covid_abstracts, column = "abstract", id = "doc",
                      weight = "tfidf", stop_words = stop_words_en())
hg
hg_measures(hg, what = "summary")
```

165 documents share 4,043 distinct words at density 0.021, and pairwise
participation 0.997 says almost every pair of abstracts shares at least
one word: one tightly connected component, the regime where the
spectral machinery below is well defined.

Before estimating any topics, the null test asks whether the overlap
structure is distinguishable from what the word and document
frequencies alone would produce. The degree-preserving swap null
(Gotelli 2000) keeps every margin fixed and shuffles who uses what.

```{r null}
hg_null_test(hg, n = 99, seed = 1)
```

Density and average edge size are invariant under the swaps (z = NA) --
the built-in check that the chain preserves what it must. Pairwise
participation is at its null value (z = -0.64, p = 0.66): in a corpus
this dense, "abstracts share words" carries no topical signal. The
signal is in the average Jaccard overlap between words: 0.00873
observed against a null of 0.00867, z = 3.32, p = 0.01. Words co-occur
in the same documents more than their frequencies require -- the excess
that topic estimation exploits. The structure is real, so estimating
topics from it is justified.

## The shared frame before the topics

The paper motivates STM by noting that frequent-keyword analysis is
dominated by a few generic terms that swamp the themes. The same
phenomenon is directly visible here: ranking words by the Chitra &
Raphael (2019) hypergraph random walk on the dual hypergraph (words as
vertices, documents as hyperedges) recovers the corpus frame, not its
topics.

```{r}
dual <- dual_hypergraph(hg)
hg_pagerank(dual, sort_by = "pagerank", n = 8)
```

*education, learning, students, online, teaching, covid, teachers,
pandemic* -- the vocabulary every abstract shares. This is the
frequent-words problem the paper describes, reproduced as a
measurement; everything topical lives below this frame, which is what
the clustering has to extract.

The same walk on the primary hypergraph ranks documents, and the
adjacency-tensor centralities (clique-expansion, Z- and H-eigenvector)
rank them by an unrelated nonlinear mechanism, so agreement is a
convergence check:

```{r}
subset(merge(hg_pagerank(hg, sort_by = "pagerank", n = 3),
             covid_abstracts, by.x = "node", by.y = "doc", sort = FALSE),
       select = c(pagerank, year, title))
subset(merge(hg_centrality(hg, sort_by = "Z", n = 3),
             covid_abstracts, by.x = "node", by.y = "doc", sort = FALSE),
       select = c(Z, title))
```

Both mechanisms put the same document first -- a 2022 resilience
synthesis whose vocabulary bridges the corpus -- so "most
representative abstract" is a robust designation, not a
method-dependent one.

## How many topics can 165 abstracts support?

The paper needed 40 candidate models and four experts to choose its
topic count. The spectral route asks the data directly: the eigenvalue
gaps of the Hayashi et al. (2020) random-walk Laplacian indicate how
many clusters the spectrum supports.

```{r scree, fig.height = 4.5}
spectrum <- head(hg_cluster(hg, k = 2, seed = 1, type = "random_walk",
                            what = "eigenvalues"), 20)
plot(spectrum$index, spectrum$value, type = "b", pch = 19, col = okabe[5],
     xlab = "Eigenvalue index", ylab = "Laplacian eigenvalue",
     main = "Spectrum of the random-walk Laplacian (first 20)")
abline(v = c(2, 4, 8), lty = c(2, 3, 3), col = okabe[6])
head(spectrum, 10)
```

The largest non-trivial gap sits after index 2 (0.022), with a mild
shoulder around 8--9: the spectrum supports anything from a coarse
2-way split to roughly nine clusters, and nothing suggests sixteen.
The decisive evidence is stability -- a partition that changes with the
k-means seed is not a finding:

```{r stability}
hg_stability(hg, k = c(3, 4, 8, 16), type = "random_walk")
```

k = 3, 4 and 8 reproduce exactly across seeds (adjusted Rand 1). At
k = 16 -- the published resolution -- the partition changes with the
seed (adjusted Rand 0.72): with about ten documents per cluster, the
sixteen-topic solution is not estimable from this sample. That is the
first result of the case study: the paper's fine resolution needs the
paper's corpus size, and the sample's ceiling is about eight topics.
Sixteen clusters on 165 documents also shows *how* overfitting looks
in keyword space:

```{r}
topics16 <- hg_cluster(hg, k = 16, seed = 1, type = "random_walk")
hg_keywords(hg, topics16, n = 4, collapse = TRUE)
```

Several of these are single-paper shards (*sputum, gamification*;
*superintendents, business*) rather than themes -- exactly what the
seed instability warns about.

## Several solutions: three, four, eight

The three stable resolutions, coarse to fine:

```{r}
topics3 <- hg_cluster(hg, k = 3, seed = 1, type = "random_walk")
xtabs(~ cluster, data = topics3)
hg_keywords(hg, topics3, n = 6, collapse = TRUE)
```

At k = 3 the split is *online teaching and learning* (83 abstracts),
*education during the pandemic in general* (62), and *science,
politics and trust* (20).

```{r}
topics4 <- hg_cluster(hg, k = 4, seed = 1, type = "random_walk")
xtabs(~ cluster, data = topics4)
hg_keywords(hg, topics4, n = 6, collapse = TRUE)
```

At k = 4 the online cluster sheds a distinct *clinical and dental
education* topic (44) and a *systems and resilience* topic (27)
separates from the general one. The finest stable solution:

```{r}
topics8 <- hg_cluster(hg, k = 8, seed = 1, type = "random_walk")
xtabs(~ cluster, data = topics8)
keywords8 <- hg_keywords(hg, topics8, n = 6)
keywords8
```

```{r keyword-panel, fig.height = 9}
par(mfrow = c(4, 2), mar = c(4, 8, 3, 1))
panels <- split(keywords8, keywords8$cluster)
invisible(Map(\(block, panel_col) {
  barplot(rev(block$score), names.arg = rev(block$word), horiz = TRUE,
          las = 1, col = panel_col, border = NA,
          main = unique(block$cluster), xlab = "tf-idf mass in cluster")
}, panels, head(okabe, length(panels))))
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
```

## The eight topics against the published sixteen

Reading the keyword panels against Table 3 of the paper gives a direct
mapping. Each recovered cluster is assigned the published topic whose
characteristic words it shares, and through the paper's own grouping,
one of the three published groups. (The paper reports sixteen topics;
its Table 3 as printed lists seventeen labeled rows, and the group
assignments below follow that table.)

```{r}
mapping <- data.frame(
  cluster = paste("Cluster", 1:8),
  keywords = c("clinical, dental, graduates", "parents, school, leadership",
               "children, anxiety, perceived", "cov, sars, distance",
               "online, learning, teaching", "health, global, mobility",
               "learning, education, system", "science, trust, political"),
  published_topic = c("Medical Education", "School closure / Teachers",
                      "Well-being", "Emergency / Remote education",
                      "Online education", "Higher education",
                      "Education", "(no published counterpart)"),
  published_group = c("diverse", "distance", "diverse", "distance",
                      "distance", "education", "education", "diverse")
)
sizes8 <- aggregate(node ~ cluster, data = topics8, length)
names(sizes8) <- c("cluster", "n_abstracts")
merge(mapping, sizes8, by = "cluster", sort = FALSE)
```

Seven of the eight recovered topics have a clear published
counterpart. The correspondences are specific, not generic: the
*clinical, dental* cluster is the paper's Medical Education topic
(which the paper singles out for the dominance of medical-education
journals in its venue analysis); the *children, anxiety* cluster is
its Well-being topic; the *cov, sars, distance* cluster is the
emergency-response strand; *parents, school, leadership* matches the
school-closure and teachers topics; and the largest recovered cluster
-- *online, learning, teaching*, 41 of 165 abstracts -- is Online
education, which is also the paper's most prevalent topic. The one
cluster without a counterpart, *science, trust, political* (public
trust in science), is a strand this 2020--2024 sample contains that
the paper's mostly-2021 corpus did not surface as a topic of its own.

## Prevalence: the sample against the published corpus

The cluster-to-topic mapping is interpretive, but the topic-to-group
assignment comes from the paper itself, so aggregating both sides to
the three published groups gives a fair quantitative comparison. The
published prevalences (paper, Fig. 2):

```{r}
published <- data.frame(
  topic = c("Online education", "Emerging technologies",
            "Higher education", "Emergency education", "Education",
            "Remote education", "Teachers", "Medical Education",
            "University", "Curriculum", "Well-being", "Equality",
            "Engagement and motivation", "Perceptions",
            "Higher education institutions", "Inclusion",
            "School closure"),
  group = c("distance", "diverse", "education", "distance", "education",
            "distance", "distance", "diverse", "education", "education",
            "diverse", "diverse", "diverse", "diverse", "education",
            "diverse", "distance"),
  prevalence = c(14.6, 11.3, 9.1, 8.6, 7.4, 6.2, 6.1, 5.5, 5.5, 4.3,
                 4.0, 3.8, 3.7, 2.5, 2.4, 2.4, 2.4)
)
published_groups <- aggregate(prevalence ~ group, data = published, sum)
published_groups
```

```{r}
ours <- merge(mapping, sizes8, by = "cluster")
our_groups <- within(
  aggregate(n_abstracts ~ published_group, data = ours, sum),
  share <- round(100 * n_abstracts / sum(n_abstracts), 1)
)
our_groups
```

```{r prevalence-plot, fig.height = 4.5}
comparison <- merge(published_groups, our_groups,
                    by.x = "group", by.y = "published_group")
bars <- t(cbind(published = comparison$prevalence,
                sample = comparison$share))
barplot(bars, beside = TRUE, names.arg = comparison$group,
        col = okabe[c(5, 1)], border = NA, ylab = "Share of corpus (%)",
        main = "Topic-group prevalence: published STM vs hypergraph sample",
        legend.text = c("Published (STM, n = 4,201)",
                        "This sample (hypergraph, n = 165)"),
        args.legend = list(bty = "n"))
```

The group-level prevalences agree to within about five points:
education-related 29.7% here against 28.7% published, distance/online
43.0% against 37.9%, diverse issues 27.3% against 33.2%. Two corpora,
two methods, a 25-fold size difference -- and the same macro-structure,
with distance/online education the largest group in both. The
divergences are also interpretable: the sample slightly overweights
distance/online (its 2022--2024 tail keeps online-learning vocabulary
prevalent) and underweights diverse issues, where the paper's
*Emerging technologies* topic (11.3%) has no counterpart cluster here.

## Topics over time

The sample spans 2020--2024, two years past the paper's window, so it
can ask what happened to the topics after the emergency. Shares by year
at the k = 4 resolution (the yearly counts are too small to carry a
formal test at k = 8, so this is a descriptive read):

```{r drift, fig.height = 4.5}
by_year <- merge(topics4, covid_abstracts, by.x = "node", by.y = "doc")
year_share <- proportions(xtabs(~ cluster + year, data = by_year),
                          margin = 2)
round(year_share, 2)
matplot(colnames(year_share), t(year_share), type = "b", lty = 1,
        pch = c(19, 17, 15, 18), col = head(okabe, 4),
        xlab = "Publication year", ylab = "Share of that year's abstracts",
        main = "Topic shares by year (k = 4)")
legend("topleft", legend = rownames(year_share), col = head(okabe, 4),
       pch = c(19, 17, 15, 18), lty = 1, bty = "n")
```

The online clinical/higher-education cluster falls from 35% of 2020
abstracts to 20% by 2022, while the general schools-and-teachers
cluster rises from 40% to 55%: the literature moves from emergency
online delivery toward broader reflection. The k = 8 emergency cluster
makes the same point more sharply --

```{r}
by_year8 <- merge(topics8, covid_abstracts, by.x = "node", by.y = "doc")
xtabs(~ year, data = subset(by_year8, cluster == "Cluster 4"))
```

-- nine of its seventeen abstracts are from 2020, the front-loaded
crisis-response strand the paper describes as the initial reaction.
(2024 contributes only five abstracts; its shares are not
interpretable.)

## Does the semantic construction agree?

The tf-idf construction sees shared vocabulary. The kNN construction
sees shared meaning: each abstract's sbert embedding becomes a
hyperedge joining its ten nearest neighbors. Clustering both the same
way, at both stable coarse resolutions, measures how much the two
notions of topic coincide. Because the two clusterings' label names
are arbitrary, the adjusted Rand index (the `ari` column) is the
statistic to read.

```{r}
data(covid_embeddings)
khg <- knn_hypergraph(covid_embeddings, k = 10)
sem3 <- hg_cluster(khg, k = 3, seed = 1, type = "random_walk")
hg_agreement(topics3, sem3, what = "table")
sem4 <- hg_cluster(khg, k = 4, seed = 1, type = "random_walk")
rbind(cbind(k = 3, hg_agreement(topics3, sem3)),
      cbind(k = 4, hg_agreement(topics4, sem4)))
```

Agreement is close to zero (adjusted Rand 0.04 at k = 3, 0.08 at
k = 4). That is a finding, not a failure: on this corpus the lexical
and semantic constructions partition the documents along genuinely
different lines -- shared terminology versus embedding proximity --
and a report should say which lens its topics come from. The published
STM, a bag-of-words method, sits on the lexical side; the comparison
above is with the matching construction.

## Scaling the topics to labels: the full cascade

The paper's qualitative phase hand-coded 54 top-cited papers. The
package's transductive and neural verbs automate exactly that step:
label a few documents per topic, spread to the rest. `hg_seeds()`
turns the clustering into those labels -- the `n` most representative
abstracts per topic by stationary probability. With the k = 4
partition as reference, the complete cascade -- closed-form spreading,
spectral-operator HGNN, and dual-attention HyperGAT -- runs on
identical inputs, and `hg_agreement()` scores each fit against the
partition (here the classifier inherits the clustering's own label
names, so the `agreement` column is the score to read). Spreading
first, with its two sensitivity checks:

```{r}
pool <- hg_cluster(hg, k = 4, seed = 1, type = "random_walk",
                   what = "embedding")
do.call(rbind, lapply(c(2, 5, 10), \(m)
  cbind(seeds_per_topic = m,
        hg_agreement(hg_classify(hg, labels = hg_seeds(pool, n = m),
                                 type = "random_walk",
                                 normalization = "class_mass"),
                     topics4))))
do.call(rbind, lapply(c(0.9, 0.99, 0.999), \(x)
  cbind(xi = x,
        hg_agreement(hg_classify(hg, labels = hg_seeds(pool, n = 5),
                                 xi = x, type = "random_walk",
                                 normalization = "class_mass"),
                     topics4))))
```

Five seeds per topic reproduce 63% of the unsupervised partition, ten
reach 70%, and two manage only 47% -- no better than always predicting
the largest topic. The result is insensitive to the spreading
parameter (62--63% across xi from 0.9 to 0.999), so the label budget,
not the tuning, is the binding constraint. The same protocol against
the finer k = 8 reference, with five seeds per topic (40 labels):

```{r}
pool8 <- hg_cluster(hg, k = 8, seed = 1, type = "random_walk",
                    what = "embedding")
hg_agreement(hg_classify(hg, labels = hg_seeds(pool8, n = 5),
                         type = "random_walk",
                         normalization = "class_mass"),
             topics8)
```

67% agreement on the eight-topic labeling -- the eight-way problem is
no harder than the four-way one once each topic has its five seeds.

The neural tier, same 20 labels. `hg_neural()` trains an HGNN whose
propagation operator is the spreading operator with learned filters;
training is stochastic, so the sbert-feature run is repeated over three
torch seeds. `hg_hypergat()`, the dual-attention document-level
architecture, runs on the raw abstracts with the same labels.

```{r neural}
seeds20 <- hg_seeds(pool, n = 5)
spread_run <- cbind(method = "hg_classify (spreading)", torch_seed = NA,
                    hg_agreement(hg_classify(hg, labels = seeds20,
                                             type = "random_walk",
                                             normalization = "class_mass"),
                                 topics4))
bow_run <- cbind(method = "hg_neural (bag-of-words)", torch_seed = 1,
                 hg_agreement(hg_neural(hg, labels = seeds20, hidden = 32,
                                        epochs = 300, seed = 1,
                                        validation = 0), topics4))
sbert_runs <- do.call(rbind, lapply(1:3, \(s)
  cbind(method = "hg_neural (sbert features)", torch_seed = s,
        hg_agreement(hg_neural(hg, labels = seeds20,
                               features = covid_embeddings, hidden = 32,
                               epochs = 300, seed = s, validation = 0),
                     topics4))))
sbert_runs
```

```{r hypergat}
hypergat_runs <- do.call(rbind, lapply(1:3, \(s)
  cbind(method = "hg_hypergat (dual attention)", torch_seed = s,
        hg_agreement(hg_hypergat(covid_abstracts, labels = seeds20,
                                 column = "abstract", id = "doc",
                                 epochs = 10, seed = s, validation = 0),
                     topics4))))
hypergat_runs
```

```{r}
cascade <- rbind(spread_run, bow_run, sbert_runs, hypergat_runs)
aggregate(agreement ~ method, data = cascade, mean)
```

The cascade's verdict is unambiguous and consistent with the package's
full benchmark study: at 20 labels on 165 documents, closed-form
spreading (63%) beats every neural architecture -- HGNN on
bag-of-words features (31%) cannot estimate its first layer from 20
examples, sbert features lift it to 52% with heavy seed-to-seed
variance, and HyperGAT (45%) has a vocabulary-by-300 embedding table
and nothing to fit it with. The same architectures, on the package's
R8/R52 benchmarks with thousands of labels, are the strongest tools it
has (HyperGAT R8 0.9665, R52 0.9433). Label budget selects the
architecture; on a corpus of this size, the closed-form verb is the
right one, and the neural tier is the documented scale-up path.

## What the case shows

Against the published multimethod synthesis, the hypergraph pipeline
on a 4% sample:

- **Recovers the macro-structure.** Group-level topic prevalence
  matches the published STM to within about five points
  (education-related 29.7 vs 28.7, distance/online 43.0 vs 37.9,
  diverse 27.3 vs 33.2), with the same largest group.
- **Recovers the dominant topic and most named topics.** Online
  education is the largest topic in both analyses; seven of eight
  recovered clusters map to specific published topics (Medical
  Education, Well-being, School closure/Teachers, Emergency/Remote
  education, Higher education, Education).
- **Finds its own resolution ceiling.** `hg_stability()` shows k = 8
  is the finest supported solution and the published k = 16 is not
  estimable from 165 documents -- the model-selection question the
  paper answered with 40 candidate models and four experts, answered
  here by a reproducibility criterion.
- **Passes its own falsification checks.** The null test confirms the
  co-occurrence structure is non-random (z = 3.32, p = 0.01) before
  any topic is interpreted; margin-invariant statistics behave as the
  null requires.
- **Adds what the paper's window could not contain.** A public-trust-
  in-science topic with no published counterpart, and a temporal drift
  from emergency online delivery (35% of 2020) toward general
  reflection (55% of 2022) that extends the paper's narrative past its
  2021 horizon.
- **Puts numbers on the labeling step.** Spreading from five examples
  per topic reproduces 63--67% of the partition at either resolution
  and is insensitive to its tuning parameter; every neural
  architecture loses at this label budget and wins at benchmark scale.

The one caution the case study insists on: lexical and semantic
constructions partition this corpus almost independently (adjusted
Rand 0.04--0.08), so "the topics" always means "the topics under a
named construction" -- which is also true, silently, of every
bag-of-words topic model, including the published one.

## References

Chitra, U., & Raphael, B. (2019). Random walks on hypergraphs with
edge-dependent vertex weights. *ICML 36*.

Ding, K., Wang, J., Li, J., Li, D., & Liu, H. (2020). Be more with
less: Hypergraph attention networks for inductive text classification.
*EMNLP 2020*.

Feng, Y., You, H., Zhang, Z., Ji, R., & Gao, Y. (2019). Hypergraph
neural networks. *AAAI 33*.

Gotelli, N. J. (2000). Null model analysis of species co-occurrence
patterns. *Ecology*, 81(9).

Hayashi, K., Aksoy, S., Park, C., & Park, H. (2020). Hypergraph random
walks, Laplacians, and clustering. *CIKM 29*.

Hubert, L., & Arabie, P. (1985). Comparing partitions. *Journal of
Classification*, 2, 193--218.

Saqr, M., Raspopovic Milic, M., Pancheva, K., Jovic, J., Peltekova,
E. V., & Conde, M. A. (2024). A multimethod synthesis of Covid-19
education research: the tightrope between covidization and
meaningfulness. *Universal Access in the Information Society*, 23,
1163--1176.

Zhou, D., Huang, J., & Scholkopf, B. (2006). Learning with hypergraphs:
Clustering, classification, and embedding. *NeurIPS 19*.
