---
title: "Hypergraph PageRank: where the interaction flow spends its time"
author: "hypernets"
date: "2026-08-25"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## The analytic question

The [windowed-hyperedge tutorial](window_hypergraph.html) established which
action combinations recur in the `human_long` pair-programming sessions and
which actions anchor them structurally (H-eigenvector centrality). This
tutorial asks a different question: **if session activity is treated as a
flow that wanders from action to action through the recurrent
combinations, where does that flow spend its time?**

Structural centrality and flow centrality can disagree. An action can sit
in many *distinct* combinations (structurally well-placed) while the
combinations it sits in are rarely *travelled* (little flow). The
instrument for the flow question is the random walk on hypergraphs with
edge-dependent vertex weights of Chitra & Raphael (2019), damped in the
PageRank manner: from an action, the walk picks one of its recurrent
combinations with probability proportional to the combination's window
count, then a member action with probability proportional to its
occurrences inside that combination; with probability 0.15 it teleports to
a uniformly random action. `hypergraph_centrality(type = "pagerank")`
returns the stationary distribution of this walk.

One theoretical point makes the hypergraph version worth having: Chitra &
Raphael prove that with edge-*independent* vertex weights the walk
collapses to an ordinary PageRank on a weighted pairwise graph. The
windowed construction is exactly the case that does *not* collapse — its
incidence cells are within-window occurrence totals, which differ across
combinations — so the scores below use information a pairwise reduction
would destroy.

```{r load}
library(hypernets)
```

## The flow scores

The hypergraph is the same recurrent structure as in the previous
tutorial: three-turn sliding windows, singletons removed, combinations
kept when they recur in at least 50 windows.

```{r build}
hg <- window_hypergraph(human_long, action = "code",
                        actor = "session_id", time = "timestamp",
                        window = 3L, min_size = 2L, min_weight = 50L)
hypergraph_centrality(hg, type = "pagerank")
```

The walk spends 27% of its time on Specify and 17% on Command — together
almost half of all flow on the two core directive actions. Request is
third at 11%. At the other end, Verify receives 4.6%, barely above the
teleportation floor: verification is visited, but the recurrent
combinations that contain it carry little traffic. The window counts
(hyperedge weights) and occurrence totals (vertex weights) are picked up
from the hypergraph automatically; no further arguments are needed.

## What recurrence adds to the ranking

Setting every combination's weight to 1 removes the recurrence
information while keeping the combination structure — the walk then treats
a combination seen 722 times and one seen 50 times as equal.

```{r unitweights}
hypergraph_centrality(hg, type = "pagerank", edge_weights = 1)
```

Two movements stand out against the recurrence-weighted scores above.
Frustrate rises from 9.2% to 12.2% and Correct from 8.0% to 10.1%: these
actions appear in *many distinct* combinations, so equal weighting favors
them. Specify falls from 27.5% to 21.3%. The comparison separates two
kinds of prominence — diversity of contexts (unit weights) versus volume
of traffic (window counts) — and the default deliberately measures the
latter.

## Flow versus structure

The H-eigenvector scores the same hypergraph on presence alone: which
actions sit in combinations whose other members are themselves central.

```{r flowstructure}
cent <- hypergraph_centrality(hg, type = c("H", "pagerank"))
do.call(cbind, cent)
```

The two rankings agree at the top (Specify, then Command) and at the
bottom (Verify last) but disagree in the middle, and the disagreement is
informative. H places Frustrate third (0.36) and puts Correct and Inquire
(0.33) above Request (0.30); PageRank reverses this, placing Request
third (0.11) well above Correct and Inquire (0.08). The reason is the
weight layer: Request sits inside the heaviest combinations in the data —
`Command, Request, Specify` alone recurs 583 times — while Frustrate,
Correct, and Inquire spread across many lighter ones. Structure says
Frustrate is well-connected; flow says the traffic runs through Request.

## Is the ranking an artifact of the damping choice?

Damping is the one free parameter, so the conclusion must survive
changing it. Two calls bracket the default of 0.85.

```{r damping}
hypergraph_centrality(hg, type = "pagerank", damping = 0.5)
hypergraph_centrality(hg, type = "pagerank", damping = 0.95)
```

The ordering of all nine actions is identical at damping 0.5, 0.85, and
0.95; only the spread changes. Less teleportation concentrates mass on
the structure (Specify grows from 20.1% at 0.5 to 29.8% at 0.95, Verify
shrinks from 7.7% to 3.5%), but no two actions trade places. The ranking
can be reported without a damping caveat. In the undamped limit the
scores converge to the walk's stationary distribution — the `pi` column
that [hypergraph_cluster()] reports, where Specify likewise holds 0.31.

## When to use which

- **`type = "pagerank"`** — the question is about *flow or traffic*:
  where activity concentrates, weighted by how often combinations recur.
  It uses both weight layers of a [window_hypergraph()], tolerates
  disconnected structure (teleportation), and its damping default can be
  reported without sensitivity caveats when the ranking is stable, as
  here.
- **`type = "H"` / `"Z"`** — the question is about *structural
  position* in the combination pattern, deliberately ignoring recurrence.
  Divergence from PageRank is itself a finding: it locates actions whose
  prominence is contextual diversity rather than volume.
- **`type = "clique"`** — a pairwise baseline (eigenvector centrality on
  the clique expansion); useful to check whether the hypergraph adds
  anything beyond a graph. When the incidence is binary, PageRank too
  collapses to a pairwise computation — the collapse theorem — so on
  binary hypergraphs prefer the cheaper graph method and reserve the
  hypergraph walk for genuinely edge-dependent weights.
- **`pi` from [hypergraph_cluster()]** — the undamped stationary
  distribution; appropriate when the hypergraph is connected and no
  teleportation smoothing is wanted.

## References

- Benson, A. R. (2019). Three hypergraph eigenvector centralities.
  *SIAM Journal on Mathematics of Data Science*, 1(2), 293–312.
- Chitra, U., & Raphael, B. J. (2019). Random walks on hypergraphs with
  edge-dependent vertex weights. *ICML 2019*, PMLR 97, 1172–1181.
- Hayashi, K., Aksoy, S. G., Park, C. H., & Park, H. (2020). Hypergraph
  random walks, Laplacians, and clustering. *CIKM 2020*, 495–504.
