---
title: "HyperGAT in R: dual-attention text classification"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 5, cache = FALSE)
library(hypernets)
okabe <- c("#E69F00", "#56B4E9", "#009E73", "#CC79A7",
           "#0072B2", "#D55E00", "#F0E442", "#999999")
```

Can dual-attention hypergraph classification -- HyperGAT, the strongest
architecture in the package's literature base -- run natively in R at
published-benchmark quality? This document walks `hg_hypergat()` from
its construction to a full R8 benchmark run: what the model computes,
how it trains, where its accuracy comes from, and where it is the wrong
tool. Every number in the prose is computed in a chunk above it.

## What the model computes

HyperGAT (Ding et al. 2020) departs from every other verb in this
package in one structural decision: instead of one corpus-wide
hypergraph, **every document is its own hypergraph**. The document's
unique words are the vertices; each of its sentences is a hyperedge
joining the words it contains. Two attention layers then pass
information across that structure:

1. **Word to sentence.** Each sentence hyperedge builds a representation
   by attending over its member words -- an attention weight decides
   which words matter for that sentence.
2. **Sentence to word.** Each word rebuilds its representation by
   attending over the sentences it appears in -- a second attention
   decides which sentences matter for that word.

Two such layers, a masked mean pool over the document's words, and a
linear classifier produce the label. Training touches only labeled
documents; the model is inductive, so any document can be scored after
training. The implementation mirrors the official code exactly
(forward-pass parity at float32 precision is part of the package's
verification suite), including its default construction: sentence
hyperedges only, without the optional LDA topic hyperedges.

The construction is easy to see on a real document:

```{r}
data(covid_abstracts)
first_doc <- head(covid_abstracts, 1)
strsplit(first_doc$abstract, "[.!?;]+")
```

Each element of that split is one future hyperedge; after tokenization
and stop-word removal its remaining words are the vertices it joins.

## A small worked example

Six two-sentence documents, four labeled:

```{r}
docs <- c(
  cooking_1 = "Simmer the soup slowly. Add onions and carrots to the pot.",
  cooking_2 = "This soup recipe needs salt. Serve the dish on a cold night.",
  cooking_3 = "Chop onions for the stew. Salt the broth and simmer.",
  space_1 = "The telescope revealed a distant galaxy. Stars everywhere.",
  space_2 = "Astronomers aimed the telescope. The stars were sharp.",
  space_3 = "A comet passed the galaxy. Astronomers watched the stars."
)
toy_fit <- hg_hypergat(
  docs,
  labels = c(cooking_1 = "cooking", cooking_2 = "cooking",
             space_1 = "space", space_2 = "space"),
  embed_dim = 16, hidden = 8, epochs = 30, lr = 0.05,
  validation = 0, seed = 1
)
toy_fit
```

The two unlabeled documents get the right classes from their sentence
structure alone: `cooking_3` shares kitchen vocabulary with the cooking
sentences, `space_3` shares astronomy vocabulary with the space ones.
`score` is the softmax probability of the winning class and `margin`
its lead over the runner-up -- the same tidy contract as `hg_classify()`
and `hg_neural()`.

## R8 at full scale

R8 is the eight-class Reuters newswire benchmark. The TextGCN-cleaned
files used elsewhere in the package's benchmarks contain no sentence
punctuation, so this run uses the official HyperGAT repository's own
corpus files, which keep sentences and ship their own train/test split
(5,501 / 2,190). Training follows the official protocol: 10 epochs,
batch size 8, learning rate 0.001 with a step decay, balanced class
weights, 10% of the training labels held out to pick the best epoch.

```{r r8-train, cache = TRUE}
official <- file.path("..", "local_testing_and_equivalence",
                      "HyperGAT_TextClassification", "data")
r8_text <- readLines(file.path(official, "R8_corpus.txt"),
                     encoding = "UTF-8", warn = FALSE)
r8_meta <- within(
  read.delim(file.path(official, "R8_labels.txt"), header = FALSE,
             col.names = c("row", "split", "label"),
             colClasses = "character"),
  node <- sprintf("R8_%05d", seq_along(row))
)
names(r8_text) <- r8_meta$node
timing <- system.time(
  r8_fit <- hg_hypergat(r8_text,
                        labels = subset(r8_meta,
                                        grepl("train", split, fixed = TRUE),
                                        select = c(node, label)),
                        min_count = 5, seed = 1)
)
timing
```

`hg_hypergat()` takes the training labels directly as the tidy
`(node, label)` table, and `hg_agreement()` scores the fit against the
held-out rows of the same table:

```{r}
r8_test <- subset(r8_meta, grepl("test", split, fixed = TRUE))
hg_agreement(r8_fit, r8_test)
```

Test accuracy 0.9652 on the 2,187 scored documents, from a
three-minute training run on a laptop CPU (177 seconds when this
document was authored) -- one verb call on raw text. Three of the
2,190 held-out documents were dropped in preprocessing for having no
usable tokens (the warning above); counting them as errors instead
gives 0.9639 on the full test set. The package's 3-seed benchmark of
the same verb, which additionally mirrors the official frequency
window and scores the full split, averages 0.9665; the full published
HyperGAT (LDA hyperedges, lemmatization, GPU) reports 0.9797. About a
point of accuracy separates this R run from the published
architecture.

## Training dynamics

```{r r8-curves, fig.height = 4}
r8_history <- attr(r8_fit, "history")
par(mfrow = c(1, 2))
plot(r8_history$epoch, r8_history$loss, type = "b", pch = 19,
     col = okabe[5], xlab = "Epoch", ylab = "Mean batch loss",
     main = "Training loss")
plot(r8_history$epoch, r8_history$val_accuracy, type = "b", pch = 17,
     col = okabe[3], xlab = "Epoch", ylab = "Validation accuracy",
     main = "Validation accuracy")
par(mfrow = c(1, 1))
r8_history
```

The loss falls an order of magnitude over the first three epochs
(0.3761 to 0.0374) and flattens once the step decay cuts the learning
rate after epoch 3. Validation accuracy is already 0.9635 after one
epoch and settles at 0.9708, the checkpoint the returned model keeps.
On R8 there is nothing to gain from training past roughly five epochs
-- the model's capacity meets this dataset early.

## Where the errors live

```{r r8-confusion, fig.height = 6}
scored <- merge(r8_fit, r8_test, by = "node", suffixes = c("_fit", ""))
confusion <- xtabs(~ label + predicted, data = scored)
confusion
shades <- colorRampPalette(c("white", okabe[5]))(64)
scaled <- sweep(unclass(confusion), 1, rowSums(confusion), "/")
image(seq_len(ncol(scaled)), seq_len(nrow(scaled)),
      t(scaled[rev(seq_len(nrow(scaled))), ]),
      col = shades, axes = FALSE, xlab = "Predicted", ylab = "Truth",
      main = "Row-normalized confusion")
axis(1, at = seq_len(ncol(scaled)), labels = colnames(scaled), las = 2)
axis(2, at = seq_len(nrow(scaled)), labels = rev(rownames(scaled)),
     las = 1)
```

```{r}
per_class <- within(
  data.frame(class = colnames(confusion),
             n_test = as.integer(rowSums(confusion)),
             precision = as.numeric(diag(confusion) / colSums(confusion)),
             recall = as.numeric(diag(confusion) / rowSums(confusion)),
             row.names = NULL),
  f1 <- 2 * precision * recall / (precision + recall)
)
per_class
```

The diagonal dominates every class, and the residual errors are
semantic rather than random. The one systematic confusion is *interest*
versus *money-fx* (11 and 7 cross-errors) -- adjacent monetary-policy
categories that human coders also separate inconsistently. The weakest
class is *ship*, the smallest (36 test documents, recall 0.72), which
leaks into *crude* and *trade* -- shipping stories are usually about
oil cargoes and trade routes. *earn*, half the test set, is nearly
perfect at recall 0.99. A report built on these predictions should
flag the interest/money-fx boundary and the ship class specifically;
overall accuracy hides both.

## In context

The package's full benchmark study (three torch seeds; pkgdown
benchmarks article) puts this run in context:

```{r}
read.csv(file.path("..", "vignettes", "articles",
                   "benchmark-hypergat.csv"))
```

R8 0.9665 and R52 0.9433 over three seeds -- the package's best numbers
on both datasets, with R52 above published transductive TextGCN
(0.9356) and both within about a point of the full published HyperGAT
(0.9797 / 0.9498), which additionally uses LDA topic hyperedges and a
WordNet lemmatizer this implementation does not.

## Where it is the wrong tool

The architecture's price is its parameter count: the word-embedding
table alone is vocabulary-by-300. With few labels there is nothing to
estimate it from. On the bundled 165-abstract covid corpus with 20
labeled documents (five per topic from the clustering in the covid
analysis tutorial), three torch seeds:

```{r covid-fewlabel, cache = TRUE}
hg <- text_hypergraph(covid_abstracts, column = "abstract", id = "doc",
                      weight = "tfidf", stop_words = stop_words_en())
covid_topics <- hg_cluster(hg, k = 4, seed = 1, type = "random_walk")
pool <- hg_cluster(hg, k = 4, seed = 1, type = "random_walk",
                   what = "embedding")
covid_runs <- do.call(rbind, lapply(1:3, \(s)
  cbind(method = "hg_hypergat", torch_seed = s,
        hg_agreement(hg_hypergat(covid_abstracts,
                                 labels = hg_seeds(pool, n = 5),
                                 column = "abstract", id = "doc",
                                 epochs = 10, seed = s, validation = 0),
                     covid_topics))))
covid_runs
aggregate(agreement ~ method, data = covid_runs, mean)
```

Mean agreement 0.45 with the reference topic partition -- against a
stable 0.63 for the closed-form `hg_classify()` with the same 20 labels.
The pattern is consistent across the package's whole neural tier: with
dozens of labels, use label spreading; with thousands, HyperGAT is the
strongest tool the package has.

## When to use which

`hg_hypergat()` is the right verb when documents have sentence
structure and labels number in the thousands: it delivers the package's
best benchmark accuracies in minutes on a CPU. `hg_neural()` (HGNN) fits
when a corpus-level hypergraph with rich vertex features (for instance
sbert embeddings) is the natural object. `hg_classify()` remains the
choice for few labels, and on corpora without meaningful sentence
boundaries (pre-cleaned text, transcripts stripped of punctuation) the
sentence construction has nothing to build from -- check the corpus
before choosing the architecture.

## References

Ding, K., Wang, J., Li, J., Li, D., & Liu, H. (2020). Be more with
less: Hypergraph attention networks for inductive text classification.
*EMNLP 2020*.

Feng, Y., You, H., Zhang, Z., Ji, R., & Gao, Y. (2019). Hypergraph
neural networks. *AAAI 33*.

Zhou, D., Huang, J., & Scholkopf, B. (2006). Learning with hypergraphs:
Clustering, classification, and embedding. *NeurIPS 19*.
