---
title: "Hypergraph text analysis with hypernets, stage by stage"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(hypernets)
```

A co-occurrence graph records that two words appeared together. A corpus
knows more than that: a word binds *all* the documents containing it at
once, a token window binds *all* the words inside it, and a document sits
near *all* its semantic neighbors. A hypergraph — edges that contain any
number of vertices — represents each of those facts directly instead of
breaking them into pairs. hypernets' text family builds three such hypergraphs
from a corpus and analyzes them with tidy verbs, on top of the hypergraph
family's spectral methods of Zhou, Huang & Schölkopf (2006) and Hayashi,
Aksoy, Park & Park (2020) (HyperNetX-parity verified) and its incidence
and measures layer.

The running corpus is the bundled `covid_abstracts`: 165 COVID-19
education-research abstracts, up to 40 per year, 2020–2024.

```{r data}
head(covid_abstracts, 3)
```

Each stage below is one question, one verb, and a reading of the actual
output.

# Stage 1 — Construction: how does a corpus become a hypergraph?

`text_hypergraph()` tokenizes deterministically in base R (lowercase, split
on non-letter runs, possessives kept as one token) and hands a weighted
incidence structure to `group_hypergraph()`. The smallest
possible corpus makes every number checkable by eye:

```{r tiny}
tiny <- text_hypergraph(
  c(a = "salt and soup", b = "soup and stars"),
  weight = "tfidf"
)
tiny
```

Two documents became the vertices; the four distinct words became
hyperedges — *soup* connects both documents, *stars* only one. The tidy
accessor shows the exact incidence, one row per document–word pair:

```{r tiny-weights}
as.data.frame(tiny)
```

The `weight` column is smoothed tf-idf,
`tf * (log((1 + N)/(1 + df)) + 1)`: *and* and *soup* occur in both of the
two documents, so their idf is exactly 1; *salt* and *stars* occur in one,
so they carry `log(3/2) + 1 = 1.405`. The smoothing (Manning, Raghavan &
Schütze 2008, the scikit-learn `smooth_idf` variant) is deliberate — a
plain `log(N/df)` gives weight zero to a word in every document, which
would silently disconnect vertices.

Two more accessors complete the corpus bookkeeping:

```{r tiny-tables}
as.data.frame(tiny, what = "documents")
as.data.frame(tiny, what = "vocabulary")
```

## The real corpus, and the two orientations

On real text, two decisions matter. First the stop list: `stop_words_en()`
ships the English function words, and corpus boilerplate is added
explicitly — in abstracts, that is words like *study* and *results*.
Second `min_count`: a word seen fewer than a handful of times cannot
connect anything and only inflates the vocabulary.

```{r construct}
stops <- c(
  stop_words_en(),
  "using", "used", "use", "based", "results", "study", "research",
  "paper", "article", "findings", "data"
)

hg <- text_hypergraph(
  covid_abstracts,
  column = "abstract",
  id = "doc",
  weight = "tfidf",
  stop_words = stops,
  min_count = 3L
)
hg
```

165 document-vertices, 1,453 word-hyperedges; the median word touches 4
abstracts, the most ubiquitous one 140 of 165. The `nodes` argument flips
the orientation — words as vertices, documents as hyperedges — which is
the right shape for questions about vocabulary rather than documents:

```{r construct-word}
word_hg <- text_hypergraph(
  covid_abstracts,
  column = "abstract",
  id = "doc",
  nodes = "word",
  weight = "tfidf",
  stop_words = stops,
  min_count = 3L
)
word_hg
```

Same incidence, transposed meaning. Every verb below accepts either
orientation; which one you build determines whether the answers are about
documents or about words.

# Stage 2 — Structure: is this hypergraph analyzable at all?

Before any spectral method, `hg_measures()` reports the structural facts.
The summary table answers the "can I even cluster this?" question:

```{r measures-summary}
hg_measures(hg, what = "summary")
```

Density 0.043 says a word-hyperedge contains on average 4.3% of all
documents; a pairwise participation of 0.996 says nearly every pair of
abstracts shares at least one word — one dense component, so the spectral
machinery applies to the whole corpus (a disconnected hypergraph is
refused by the clustering verbs with an explicit error, not analyzed
wrongly). The node table ranks the documents themselves:

```{r measures-nodes}
head(hg_measures(hg))
```

`hyperdegree` counts how many distinct (kept) words a document
contributes to; `strength` sums the sizes of those word-hyperedges, so it
mixes how many words a document uses with how widely shared they are. The
first abstract sits in 71 word-hyperedges whose sizes sum to 1,322 — a
document built from broadly shared vocabulary, which will act as a hub in
anything walk-based. On small hypergraphs the
`edges` and `overlap` tables are readable directly:

```{r measures-overlap}
hg_measures(tiny, what = "edges")
hg_measures(tiny, what = "overlap")
```

The overlap table is one row per hyperedge pair — *and* and *soup* share
both their documents (Jaccard 1), each shares exactly one document with
*salt* and *stars*. On the full corpus this table has
`1453 * 1452 / 2` rows; it stays a data.frame you filter with arguments,
not something to print.

# Stage 3 — Centrality: which words organize the corpus?

`hg_centrality()` computes three centralities: `clique` (eigenvector
centrality on the weighted clique expansion — the "collapse to a graph"
baseline) and the tensor `Z`- and `H`-eigenvector centralities (Benson
2019), which use the hypergraph's higher-order structure directly instead
of collapsing it. `sort_by` and `n` make top-k selection an argument, not
a post-processing step. The windowed hypergraph of the 2020 abstracts
(built fully in Stage 6) is small enough to compare definitions:

```{r centrality}
early <- subset(covid_abstracts, year == 2020)
windowed <- text_hypergraph(
  early,
  column = "abstract",
  id = "doc",
  construction = "window",
  window = 3,
  stop_words = stops,
  min_count = 3L
)
hg_centrality(windowed, type = c("clique", "H"), sort_by = "H", n = 8)
```

The two definitions agree on the theme vocabulary but not on its order:
the clique expansion puts *education* far ahead (0.417) with *covid*
second; the H-eigenvector keeps *education* first but demotes *covid* to
eighth, promoting *learning*, *students* and *teachers*. The reading:
*covid* co-occurs broadly but diffusely — pairwise collapse rewards that;
the tensor centrality rewards words that sit in strong *whole windows*
together, which is the teaching-and-learning cluster. When the ranking is
the finding, report which centrality produced it.

# Stage 4 — Clustering: do the abstracts group by theme?

`hg_cluster()` embeds the hypergraph Laplacian and k-means the embedding
(deterministically, with an explicit seed). Two Laplacians are available:
`"zhou"` (Zhou et al. 2006) and `"random_walk"` — the Hayashi et al.
(2020) construction whose random walk uses *edge-dependent vertex
weights*. tf-idf is exactly such a weighting (the same word counts
differently in different documents), so `"random_walk"` is the
theory-matched choice here:

```{r cluster}
clusters <- hg_cluster(hg, k = 4, type = "random_walk", seed = 1)
aggregate(node ~ cluster, data = clusters, FUN = length)
```

```{r cluster-years}
merged <- merge(clusters, covid_abstracts, by.x = "node", by.y = "doc")
aggregate(year ~ cluster, data = merged, FUN = mean)
```

Sizes 16/44/29/76, and every cluster's mean year sits between 2021.1 and
2021.7: the partition is thematic, not temporal — this literature's
vocabulary separates by topic while staying stable across the pandemic
years. Does the Laplacian choice matter? Refit with `"zhou"` and
cross-tabulate:

```{r cluster-laplacian}
clusters_zhou <- hg_cluster(hg, k = 4, type = "zhou", seed = 1)
table(random_walk = clusters$cluster, zhou = clusters_zhou$cluster)
```

It matters a lot: no clean diagonal, whole clusters reshuffle. The two
Laplacians answer differently because only one of them consumes the
edge-dependent weights that tf-idf provides. The decisive practice: with
tf-idf weighting, report the `"random_walk"` partition and use `"zhou"`
only as a robustness view — groupings that survive both are the ones to
build interpretation on. (The number of clusters is the other
consequential choice; `vignette("text-hypergraphs")` runs the k = 3
sensitivity check — the two compact themes are stable, the large residual
cluster is not.)

# Stage 5 — Classification: can six labels classify 165 documents?

`hg_classify()` is transductive label spreading (Zhou et al. 2006):
labels for a few vertices propagate over the hyperedges until every
vertex has a score for every class. Labeling the three alphabetically
first abstracts of 2020 as `"early"` and of 2024 as `"late"` asks whether
pandemic-era vocabulary dates an abstract:

```{r classify}
labels <- c(
  "2-s2.0-85085897904" = "early",
  "2-s2.0-85087124472" = "early",
  "2-s2.0-85087646998" = "early",
  "2-s2.0-85107489479" = "late",
  "2-s2.0-85112557152" = "late",
  "2-s2.0-85113355756" = "late"
)
predictions <- hg_classify(hg, labels = labels)
predicted_years <- merge(predictions, covid_abstracts,
                         by.x = "node", by.y = "doc")
aggregate(year ~ predicted, data = predicted_years, FUN = mean)
table(
  predicted = predicted_years$predicted,
  actually_2020 = predicted_years$year == 2020
)
```

The classifier commits to `"early"` for exactly three abstracts — every
one genuinely from 2020 — and calls everything else `"late"`, 2020
included. Correct where it commits, and honest about how weak the year
signal is: this literature's vocabulary converged almost immediately. The
`margin` column (best minus second-best score) is the per-document
confidence; low margins mark documents the labels barely reach. The
damping parameter `xi` controls how far labels spread, and here it is not
doing the work:

```{r classify-xi}
predictions_90 <- hg_classify(hg, labels = labels, xi = 0.90)
table(xi_099 = predictions$predicted, xi_090 = predictions_90$predicted)
```

Dropping xi from 0.99 to 0.90 flips 2 of 165 documents. The conservative
result is a property of the corpus structure, not of the tuning
parameter — which is exactly what a sensitivity check is for.

# Stage 6 — Windowed hyperedges: does word order change the answer?

The bag construction cannot see that *distance* and *learning* are a
phrase. `construction = "window"` slides a token window over each
document (never crossing document boundaries); each distinct window
content — the set of words within `window` tokens of each other — is one
hyperedge, weighted by its window count. This is the sequential
construction of Ding et al. (2020, Sec. 3.2) with weights — the text
front end of `window_hypergraph()`, with which it shares its incidence
matrix (a shipped package test; a word repeated inside one window is
collapsed by the set-valued hyperedge here and counted by the engine).
Stop-word and `min_count` filtering run
*before* windowing, so a removed token closes its gap.

```{r window}
windowed
hg_centrality(windowed, type = "clique", sort_by = "clique", n = 10)
```

Compare the bag ranking on the same 40 abstracts:

```{r window-vs-bag}
bag_2020 <- text_hypergraph(
  early,
  column = "abstract",
  id = "doc",
  nodes = "word",
  weight = "tfidf",
  stop_words = stops,
  min_count = 3L
)
hg_centrality(bag_2020, type = "clique", sort_by = "clique", n = 10)
```

Same corpus, same stop list, one different construction: the windowed
ranking puts **distance** third and the bag ranking does not place it in
the top ten at all. Adjacency is the signal — *distance* matters in 2020
almost exclusively inside *distance learning* and *distance education*.
The window size is arbitrary, so it gets the standard check
(`vignette("constructions")` runs it): at `window = 4`, *distance* rises
to second — the conclusion survives.

One practical constraint: distinct windows multiply fast, and the
engine's incidence is dense. Window modest corpora (here: 40 abstracts,
2,000 windows, a 5 MB incidence) or filter hard first.

# Stage 7 — kNN embedding hyperedges: does meaning beat vocabulary?

Two abstracts can discuss the same phenomenon with disjoint vocabulary.
`construction = "knn"` places each document in an embedding space and
forms one hyperedge per document — itself plus its `k` nearest neighbors,
weighted by cosine similarity. The bundled `covid_embeddings` matrix
(sbert's pinned `all-MiniLM-L6-v2`, L2-normalized, built by
`data-raw/covid_embeddings.R`) keeps everything offline; with the sbert
package installed, omitting `embeddings` encodes the text in-place, and
`knn_hypergraph(embeddings, k)` takes any embedding matrix directly. The
binary variant of the construction is verified support-identical to the
CRAN oracle `HyperG::knn_hypergraph()`.

```{r knn}
knn_hg <- text_hypergraph(
  covid_abstracts,
  column = "abstract",
  id = "doc",
  construction = "knn",
  k = 10,
  embeddings = covid_embeddings
)
knn_hg
knn_clusters <- hg_cluster(knn_hg, k = 4, type = "random_walk", seed = 1)
aggregate(node ~ cluster, data = knn_clusters, FUN = length)
```

Balanced groups — 41/41/37/46 — where the bag clustering of Stage 4 left
one dominant 76-document residual: semantic neighborhoods spread
structure more evenly than shared vocabulary. The two views agree only
partially:

```{r knn-vs-bag}
table(knn = knn_clusters$cluster, bag = clusters$cluster)
```

Two themes are recovered by both constructions (the 32- and 37-document
blocks); the rest reshuffles. Neither is "the" truth — shared vocabulary
and semantic proximity are different questions — and a grouping stable
across both is the one to trust. The choice of `k` gets its sensitivity
check in `vignette("constructions")`: against `k = 15`, 143 of 165
documents keep their cluster alignment; against `k = 5`, 125 do — robust
upward, softer downward. And every number in this stage is conditional on
one pinned encoder; a claim that matters should be re-run with a second
model (`model =` makes that a one-argument change).

# When to use which

- **Bag, documents as nodes** — the default. Document clustering
  (`hg_cluster`, `"random_walk"` whenever weights are tf-idf) and
  few-label classification (`hg_classify`).
- **Bag, words as nodes** — vocabulary structure: central terms
  (`hg_centrality`), word communities.
- **Window** — when order carries signal: phrase vocabulary,
  collocations, sequence-like corpora. Words are the vertices; keep
  corpora modest or filter hard.
- **kNN** — when meaning matters more than word overlap:
  paraphrase-heavy corpora, or as the explicit spectral alternative to
  UMAP+HDBSCAN. Documents are the vertices; the geometry is the
  encoder's, so say which encoder.
- **`hg_measures` first, always** — density, participation, and the
  node table say whether the spectral verbs are even applicable, and
  which hubs will dominate them.
- Everything is deterministic given a seed; every arbitrary choice
  (stop list, `min_count`, `k`, `window`, `xi`, number of clusters) has
  a one-argument sensitivity check — run it and report whether the
  conclusion survives. The vignettes show each check on this corpus.

# References

Benson, A. R. (2019). Three hypergraph eigenvector centralities.
*SIAM Journal on Mathematics of Data Science*, 1(2).

Ding, K., Wang, J., Li, J., Li, D., & Liu, H. (2020). Be more with less:
Hypergraph attention networks for inductive text classification.
*EMNLP 2020*. doi:10.18653/v1/2020.emnlp-main.399.

Hayashi, K., Aksoy, S. G., Park, C. H., & Park, H. (2020). Hypergraph
random walks, Laplacians, and clustering. *CIKM 2020*.
doi:10.1145/3340531.3412034.

Manning, C. D., Raghavan, P., & Schütze, H. (2008). *Introduction to
Information Retrieval*. Cambridge University Press.

Zhou, D., Huang, J., & Schölkopf, B. (2006). Learning with hypergraphs:
Clustering, classification, and embedding. *NeurIPS 2006*.
