---
title: "Windowed sequence hyperedges: multi-way co-occurrence in coded sequences"
author: "hypernets"
date: "2026-08-25"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
```

## The analytic question

Coded interaction sequences record one action at a time: a person specifies
a requirement, the partner interrupts, a command follows. A transition
network answers *what follows what*; a co-occurrence network answers *which
pairs appear together*. Neither answers the question this tutorial takes
up: **which actions occur together as groups within short spans of
activity, and do those groups recur often enough to constitute a
structure?**

A pairwise network cannot hold the answer, because a group of three
actions inside one span is not the same observation as its three pairs.
The hypergraph keeps the group intact: each short span of a sequence
becomes one *hyperedge* over the distinct actions inside it. The
construction follows Ding, Wang, Li, Li & Liu (2020); the resulting
weighted incidence matrix carries edge-dependent vertex weights in the
sense of Chitra & Raphael (2019), which the spectral machinery of Zhou,
Huang & Schölkopf (2006) and Hayashi, Aksoy, Park & Park (2020) consumes
directly.

The data are `human_long`, bundled with hypernets: 10,796 coded human
turns from 429 human–AI pair-programming sessions, nine action codes, each
code carrying an a-priori cluster label (Directive, Evaluative,
Metacognitive) that the analysis below never uses — it will serve as an
external check at the end.

```{r load}
library(hypernets)
```

## How a sequence becomes a hypergraph

Before the real data, a four-event sequence small enough to verify by
hand. With `window = 2` the windows are (a,b), (b,a), (a,c); the first two
contain the same set of states, so they collapse into one hyperedge
observed twice.

```{r tiny}
hg_tiny <- window_hypergraph(list(s1 = c("a", "b", "a", "c")), window = 2L)
as.data.frame(hg_tiny)
```

The table has one row per hyperedge. The set {a, b} was seen in 2 windows
(`weight = 2`), the set {a, c} in 1. Weight is therefore *recurrence*: how
many spans of the sequence produced exactly this combination. Repeats
inside a window count as occurrences in the incidence matrix but not as
extra members — a window (a, a) yields the singleton hyperedge {a} with an
incidence cell of 2. This is what makes the incidence matrix
edge-dependent-weighted rather than binary.

## Recurrent action combinations in real sessions

Which combinations of actions recur inside three-turn spans of the
sessions? The long format goes in directly: `action` names the code
column, `actor` groups turns into sessions, `time` orders them.

```{r full}
hg_all <- window_hypergraph(human_long, action = "code",
                            actor = "session_id", time = "timestamp",
                            window = 3L)
hg_all
```

The 429 sessions produce 9,942 windows (six sessions were shorter than
three turns and contributed none — the constructor counts them rather than
dropping them silently). The 129 hyperedges are exactly the 9 singletons,
36 pairs, and 84 triples that nine codes allow: with this much data,
*every possible combination occurs at least once*. Presence alone is
saturated, so any presence-based summary would be flat. The information is
in the weights:

```{r top10}
head(as.data.frame(hg_all, sort_by = "weight"), 10)
```

The distribution is far from flat. `Command, Specify` dominates with 722
windows, followed by `Command, Request, Specify` (583). Specify appears in
eight of the top nine combinations: specification is the connective tissue
of these sessions. The gap between the strongest combination (722) and the
weakest (1) is what the saturated presence structure hides.

Since rare combinations are noise for structural analysis, the
`min_weight` argument keeps only recurrent ones — the same role `min_freq`
plays in higher-order rule extraction. Fifty windows out of 9,942 is a
conservative floor; `min_size = 2` also removes the nine singletons, which
carry no co-occurrence information.

```{r recurrent}
hg <- window_hypergraph(human_long, action = "code",
                        actor = "session_id", time = "timestamp",
                        window = 3L, min_size = 2L, min_weight = 50L)
hg
```

Fifty-six combinations — 22 pairs and 34 triples — recur at least 50
times. This is the structure the rest of the analysis works on.

## How concentrated is the recurrent structure?

```{r measures}
hypergraph_measures(hg)
```

Three numbers matter here. First, hyperdegree now ranges from 8 to 33
(median 15): after filtering, the codes participate in very different
numbers of recurrent combinations, where the unfiltered hypergraph had
forced every code to the same value. Second, pairwise participation is
0.92 — almost every pair of codes still shares at least one recurrent
hyperedge, so the structure is concentrated but not fragmented. Third, the
intersection profile shows 264 hyperedge pairs overlapping in two codes:
the triples are largely *extensions* of the strong pairs (for example
`Command, Specify` extending to `Command, Request, Specify`), not separate
formations.

## Which actions anchor the combinations?

The H-eigenvector centrality of Benson (2019) scores each node on the
uniform-size hypergraph tensor; unlike degree it rewards being in
combinations whose *other members* are themselves central.

```{r centrality}
hypergraph_centrality(hg, type = "H")
```

Specify (0.43) is the clear anchor, with Command (0.38) second — the two
directive actions sit inside most strong combinations. Verify (0.26),
Refine (0.28), and Interrupt (0.28) are peripheral: they occur, and recur,
but in combinations anchored by others. The ordering matches what the
top-weight table already suggested, now as a single per-node score.

## Do the actions organize into higher-order groups?

The random-walk Laplacian of Hayashi et al. (2020) uses both weight
layers this hypergraph carries: the occurrence-count incidence
(edge-dependent vertex weights) and the window counts, which
`hypergraph_cluster()` picks up automatically as hyperedge weights.
Spectral k-means is stochastic, so the run is seeded and repeated under a
second seed as a stability check.

```{r cluster}
cl <- hypergraph_cluster(hg, k = 2L, type = "random_walk", seed = 1)
cl_check <- hypergraph_cluster(hg, k = 2L, type = "random_walk", seed = 99)
identical(as.data.frame(cl), as.data.frame(cl_check))
as.data.frame(cl)
```

The partition is identical under both seeds, so it can be reported as
stable. Cluster 2 is {Correct, Frustrate, Refine}; Cluster 1 is everything
else, with Specify holding the largest stationary probability
(`pi = 0.31`) — a random walker on the recurrent combinations spends
almost a third of its time on Specify.

This is where the withheld labels come in: the coding scheme's Evaluative
cluster is {Correct, Frustrate, Refine, Verify}. The hypergraph recovered
three of its four members from co-occurrence structure alone. The
defector is informative: Verify co-occurs with Specify and Command (the
combinations `Command, Specify, Verify` and `Request, Specify, Verify`
each recur over 80 times), so *within short spans* verification behaves
as part of the directive workflow, not the evaluative one.

```{r clusterplot, fig.alt = "Spectral embedding of the nine action codes, colored by cluster"}
plot(cl)
```

The embedding shows the same picture geometrically: the corrective codes
separate on the second spectral dimension while Specify sits at the center
of the main group.

## Does the window size change the conclusions?

Window size is the one consequential arbitrary choice in this pipeline, so
the headline result must survive a different value.

```{r sensitivity}
hg4 <- window_hypergraph(human_long, action = "code",
                         actor = "session_id", time = "timestamp",
                         window = 4L, min_size = 2L, min_weight = 50L)
head(as.data.frame(hg4, sort_by = "weight"), 8)
```

With four-turn windows the same combinations lead: `Command, Request,
Specify` (447), `Command, Interrupt, Specify` (357), `Command, Specify`
(343), and their four-code extension `Command, Interrupt, Request,
Specify` now enters at 241 — wider windows admit larger sets, but the
membership of the dominant combinations is unchanged. The conclusion —
directive combinations anchored by Specify dominate, corrective actions
form their own block — does not depend on the choice of three versus four.

## When to use which

- **`window_hypergraph()`** — sequences, and the question is about
  combinations *local in time*. `step = 1` (sliding) weights every span;
  `step = window` (tumbling) counts disjoint spans and suits length
  comparisons across groups. Raise `min_weight` before structural
  analysis on dense data; this tutorial's saturation of all 129 possible
  sets is the norm, not the exception, once windows are plentiful.
- **`group_hypergraph()`** — the grouping is already categorical (a whole
  session, a document, a team) and position in time is irrelevant. A
  whole-sequence window reproduces it exactly.
- **`build_hypergraph()`** — the input is an estimated pairwise network
  and the hyperedges should be its cliques.
- **`clique_expansion()`** — hands the hypergraph back to pairwise
  machinery (Nestimate, cograph) when a graph method is needed; the
  multi-way distinctions collapse into edge weights at that point, which
  is exactly what this tutorial's construction exists to avoid.

## References

- Benson, A. R. (2019). Three hypergraph eigenvector centralities.
  *SIAM Journal on Mathematics of Data Science*, 1(2), 293–312.
- Chitra, U., & Raphael, B. J. (2019). Random walks on hypergraphs with
  edge-dependent vertex weights. *ICML 2019*, PMLR 97, 1172–1181.
- Ding, K., Wang, J., Li, J., Li, D., & Liu, H. (2020). Be more with less:
  Hypergraph attention networks for inductive text classification.
  *EMNLP 2020*, 4927–4936.
- Hayashi, K., Aksoy, S. G., Park, C. H., & Park, H. (2020). Hypergraph
  random walks, Laplacians, and clustering. *CIKM 2020*, 495–504.
- Zhou, D., Huang, J., & Schölkopf, B. (2006). Learning with hypergraphs:
  Clustering, classification, and embedding. *NeurIPS 2006*, 1601–1608.
