---
title: "Benchmarks: R8, R52, MR, Ohsumed, 20NG"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

How far does closed-form hypergraph label spreading go on the five standard
text-classification benchmarks? This article reports test-set accuracy of
`text_hypergraph(sparse = TRUE)` + `hg_classify()` on R8, R52, MR, Ohsumed,
and 20-Newsgroups, against a tf-idf nearest-centroid baseline and against
the published accuracy tables in Ding et al. (2020). Every number below was
produced by the harness in `benchmarks/` of the package repository; nothing
is copied from memory.

## Setup

The corpora and train/test splits are the exact files of Yao, Mao & Luo
(2019), which Ding et al. (2020) also use -- any other cleaning would make
the numbers incomparable. `benchmarks/download.sh` fetches them. Each
dataset becomes a sparse document--word hypergraph (documents are vertices,
words are hyperedges, tf-idf incidence weights); the training documents are
the labeled seeds; the spreading solution classifies every test document in
one conjugate-gradient solve per class.

```{r, eval = FALSE}
hg <- text_hypergraph(docs, weight = "tfidf", sparse = TRUE)
fit <- hg_classify(hg, labels = train_labels, normalization = "class_mass")
```

Two checks tie the pipeline to the published setup. The split sizes match
Ding et al.'s Table 1 exactly on all five datasets. The vocabulary sizes
match exactly on R8 (7,688 hyperedges) and R52 (8,892); on MR, Ohsumed and
20NG ours are smaller (18,151 vs 18,764; 13,349 vs 14,157; 36,970 vs
42,757) because the package tokenizer drops digit-containing tokens. MR has
one document unreachable from the rest of the corpus; the harness runs on
the giant component and would score unreachable test documents as errors
(here that document is not in the test set).

## Full-split results

`accuracy` is test-set accuracy with a 1,000-resample percentile bootstrap
CI; `fit_s` is the wall-clock seconds for the transductive solve itself.

```{r}
results <- read.csv("benchmark-results.csv")
knitr::kable(
  subset(results,
         select = c(dataset, method, accuracy, ci_low, ci_high, macro_f1,
                    build_s, fit_s)),
  digits = 4
)
```

Three findings. First, on the two class-balanced datasets the hypergraph
classifier beats the centroid baseline outright: 0.8477 vs 0.7796 on 20NG
and 0.7684 vs 0.6851 on MR. Second, on the skewed multi-class corpora (R8,
R52, Ohsumed) the centroid wins; label spreading pays for class imbalance
even after normalization, and R52's 52 heavily skewed classes are its worst
case (0.5284 vs 0.8797). Third, the `hypergraph_zhou` rows -- the raw Zhou
(2006) argmax with no normalization -- collapse onto the majority class
(0.4947 on R8 is exactly the majority-class rate; 0.1020 on 20NG), which is
why `normalization = "class_mass"` (Zhu, Ghahramani & Lafferty 2003) exists
and is used everywhere else in this article. The EDVW walk
(`hypergraph_random_walk_cmn`) is best on Ohsumed (0.5110 vs 0.4702) and
second elsewhere.

The cost side: the largest solve -- 20 classes over 18,846 documents x
36,970 hyperedges -- fits in 1.2 seconds on a laptop CPU, with no training,
no gradient steps and no GPU. Ding et al.'s Table 3 reports 1,479 MB of GPU
memory for TextGCN and 180 MB for HyperGAT on the same corpus.

## Against the published tables

Published rows are means over 10 runs from Ding et al. (2020), Table 2;
our transductive rows are deterministic given the split.

```{r}
published <- read.csv("benchmark-published.csv")
knitr::kable(published, digits = 4)
```

On 20NG the hypergraph classifier (0.8477) is within two points of
transductive TextGCN (0.8643) and HyperGAT (0.8662), and above fastText
(0.7938). On MR (0.7684) it edges out published transductive TextGCN
(0.7674) and trails HyperGAT (0.7832). On R8, R52 and Ohsumed the trained
models are clearly ahead. The honest summary: a closed-form spectral method
with zero trained parameters recovers most of the accuracy of trained
graph neural networks on balanced corpora and is not competitive on
skewed ones.

## The low-label regime

Label spreading is designed for few labels, so the full-split table above
is its worst case. Here the training seeds are subsampled (stratified, at
least one per class, five draws per fraction; the SD column is over
draws) and both methods receive the same seeds.

```{r}
lowlabel <- read.csv("benchmark-lowlabel.csv")
knitr::kable(lowlabel, digits = 4)
```

On MR the hypergraph classifier leads at every fraction, from 72 seeds
(0.5675 vs 0.5541) to 1,422 seeds (0.6984 vs 0.6677): with few labels, the
document--word structure carries information the centroid cannot use. On R8
and Ohsumed the centroid stays ahead at every fraction -- strong tf-idf
class signatures survive subsampling better than spreading does. Few
labels help the case for structure, but they do not overturn the class
balance effect.

## The neural tier: HGNN

`hg_neural()` trains the two-layer hypergraph convolutional network of
Feng et al. (2019) natively in R ({torch}), on the same sparse
document--word hypergraph and tf-idf features. Its propagation matrix is
exactly the Zhou operator the transductive classifier spreads with --
verified to machine precision in the tests, with forward-pass parity
against the official implementation (DHG) at float32 precision. Training
is stochastic, so every number is the mean of three seeds (SD and range
shown); each run holds out 10% of the seeds (stratified) to pick the best
epoch. Two configurations are reported: the original paper's
(`lr = 0.001`, 200 epochs, tuned for citation graphs with dense features)
and the package default (`lr = 0.01`, 600 epochs, selected by validation
accuracy for high-dimensional sparse text features).

```{r}
neural <- read.csv("benchmark-neural.csv")
knitr::kable(
  subset(neural,
         select = c(dataset, epochs, lr, accuracy, sd, acc_min, acc_max,
                    macro_f1, fit_s)),
  digits = 4
)
```

The tuned configuration lifts R8 to 0.9539 (above the centroid's 0.9246;
published TextGCN 0.9707) and R52 to 0.8440 (a 32-point jump over
transduction's 0.5284, though still under the centroid), and holds MR at
0.7692 -- the best HGNN number in this article for that dataset. On
Ohsumed and 20NG, however, HGNN underperforms the closed-form
transduction at every configuration tested (20NG: 0.5355 tuned, 0.6347
at paper defaults, vs 0.8477 for transduction). The loss curves say this
is not an optimization failure alone: the corpus-level document-node
design oversmooths -- two rounds of propagation through high-degree word
hyperedges blur exactly the distinctions 20 newsgroups need. The
literature's answer is document-level hypergraphs with attention over
hyperedges: HyperGAT, next.

## HyperGAT: document-level hypergraphs with dual attention

`hg_hypergat()` implements Ding et al. (2020) natively: every document
becomes its own hypergraph (its unique words as vertices, its sentences
as hyperedges), a first attention layer aggregates words into sentence
representations, a second aggregates sentences back into words, and a
masked mean pool feeds the classifier. The implementation mirrors the
official code (verified against its source, with forward-pass parity at
float32 precision). `semantic = "none"` uses sentence hyperedges alone;
`semantic = "lda"` adds the paper's online-LDA topic hyperedges, trained on
labeled documents only, and precomputed official keyword lists can be passed
to replay the upstream preprocessing boundary exactly. The benchmark table
below predates that implementation and remains explicitly sentence-only, so
its closest published comparator is the paper's "w/o semantic" ablation.
One further honest difference: the official
preprocessing lemmatizes with WordNet; ours does not. The benchmark runs
on the official repository's own corpus files (the TextGCN cleaned files
used above contain no sentence punctuation at all), whose R8 split is
5,501/2,190 rather than TextGCN's 5,485/2,189. Official training
protocol throughout: 10 epochs, batch 8, StepLR, balanced class
weights, 10% validation checkpointing; three torch seeds.

```{r}
hypergat <- read.csv("benchmark-hypergat.csv")
knitr::kable(hypergat, digits = 4)
```

This is the package's strongest result. R8 reaches 0.9665 (SD 0.0034) --
above every other method in this article, within 1.3 points of the full
published HyperGAT (0.9797) and at published transductive TextGCN level
(0.9707). R52 reaches 0.9433 (SD 0.0018), which **exceeds published
transductive TextGCN (0.9356)** and sits 0.7 points from published
HyperGAT (0.9498) -- against our previous R52 best of 0.8797 (the
centroid). Wall-clock: under four minutes per training run on a laptop
CPU. The architecture, not more optimization of the corpus-level
design, was what the skewed news corpora needed.

## When to use which

Use `hg_classify(normalization = "class_mass")` when classes are roughly
balanced or labels are scarce, and always over the raw argmax: the raw
rule's majority-class collapse on skewed seeds is total. Use
`type = "random_walk"` when tf-idf weights should shape the walk itself;
it was best on Ohsumed here. Use the neural tier when plentiful labels can pay for training, and pick
the architecture by corpus shape: `hg_hypergat()` for classification of
documents with sentence structure -- it holds the package's best R8
(0.9665) and R52 (0.9433, above published transductive TextGCN) -- and
`hg_neural()` (HGNN) when the corpus-level hypergraph with rich vertex
features is the natural object (it holds MR at 0.7692). Neither beats
the closed-form classifier everywhere: on 20NG the transduction's
0.8477 still leads everything we trained. For skewed many-class corpora
without training, a tf-idf centroid remains a strong, nearly free
baseline; the full published HyperGAT (with LDA hyperedges) holds the
state of the art. The package's contribution is that every non-published
row above runs natively in R -- the transductive ones in seconds, the
neural ones in minutes on a CPU -- each from one verb.

## References

Feng, Y., You, H., Zhang, Z., Ji, R., & Gao, Y. (2019). Hypergraph
neural networks. *AAAI 33*.

Ding, K., Wang, J., Li, J., Li, D., & Liu, H. (2020). Be more with less:
Hypergraph attention networks for inductive text classification.
*EMNLP 2020*.

Yao, L., Mao, C., & Luo, Y. (2019). Graph convolutional networks for text
classification. *AAAI 2019*.

Zhou, D., Huang, J., & Scholkopf, B. (2006). Learning with hypergraphs:
Clustering, classification, and embedding. *NeurIPS 19*.

Zhu, X., Ghahramani, Z., & Lafferty, J. (2003). Semi-supervised learning
using Gaussian fields and harmonic functions. *ICML 20*.
