---
title: "Sixteen topics in COVID-19 education research: whole abstracts as a hypergraph"
author: "honets"
date: "2026-09-06"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "", fig.width = 9, dpi = 110,
                      out.width = "100%")
library(honets)
has_torch <- requireNamespace("torch", quietly = TRUE) &&
  torch::torch_is_installed()
okabe_ito <- c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00",
               "#CC79A7", "#999999", "#000000")
```

> The `covid` corpus bundled with **sbert**: abstracts of COVID-19 research
> on education, children, schools and society, 2020 to 2024, each with its
> publication year. Here every abstract is one node of a document-word
> hypergraph and is read whole. No sentence encoder is involved, so nothing
> is truncated at a token window. Every number and figure is recomputed at
> knit time.

# The corpus

Records indexed without an abstract are dropped, as are exact duplicate
abstracts. Bare numbers never enter the vocabulary (the tokeniser keeps
alphabetic tokens only), so the year and "19" need no stop list; the
scientific boilerplate that every abstract shares does, because a word in
every document separates nothing.

```{r corpus}
# clean_text() repairs the export: HTML, references, URLs, the trailing
# copyright notice (a publisher cluster at any fine cut if left in) and the
# database's placeholder for a missing abstract; unique() collapses
# duplicated records
abstracts <- clean_text(sbert::covid, column = "Abstract",
                        remove = "no abstract available", min_content = 0.5)
abstracts <- unique(abstracts)
abstracts$doc <- sprintf("abs_%04d", seq_len(nrow(abstracts)))
table(abstracts$Year)
boilerplate <- c("covid", "coronavirus", "sars", "cov", "pandemic",
                 "disease", "study", "studies", "result", "results",
                 "conclusion", "conclusions", "background", "method",
                 "methods", "objective", "aim")
stops <- c(stop_words_en(), boilerplate)
```

# The hypergraph

Documents are nodes; every word kept after the stop lists that occurs in at
least five abstracts is a hyperedge binding the abstracts that use it, with
tf-idf incidence weights, stored sparse.

```{r hypergraph}
hg <- text_hypergraph(abstracts, column = "Abstract", id = "doc",
                      weight = "tfidf", stop_words = stops,
                      min_count = 5L, sparse = TRUE)
hg
```

# Sixteen topics, and what the spectrum says about that

The count used throughout is **sixteen**, the number of topics in the
published structural topic model of this literature (Saqr et al. 2023,
*Universal Access in the Information Society*), so the partition is the one
a reader of that paper can set beside its topics. The Laplacian spectrum
should be read alongside that choice rather than hidden by it: the gap
after the second eigenvalue is the large one, the gaps after the third and
fourth are an order of magnitude smaller, and from the fifth on there is
nothing. Two and four are the cuts the hypergraph itself insists on;
sixteen is a finer partition below those seams, supported by the words
each topic owns (next section) rather than by the spectrum.

```{r eigen}
hg_cluster(hg, k = 18, seed = 1, what = "eigenvalues")
```

```{r cluster}
topics <- hg_cluster(hg, k = 16, seed = 1)
table(topics$cluster)
```

# Heavy words and distinctive words

`hg_keywords()` ranks each topic's words under the score named in `type`.
Ranked by the score itself, every topic in this corpus returns the same
list: "education", "students", "learning", "online", "teaching" are the
corpus, not the topics. Class-based tf-idf demotes them only mildly, because
BERTopic's idf term is `log(A / f + 1)` and these words are frequent in
every class.

```{r heavy}
hg_keywords(hg, topics, n = 6, collapse = TRUE,
            type = c("frequency", "ctfidf"))
```

`sort_by = "share"` ranks by the fraction of a word's total score that falls
in the topic, and `min_docs` is the support floor that keeps a word owned by
a topic because it occurs in a handful of abstracts from reaching the top.
This is where the sixteen topics separate.

```{r distinctive}
distinctive <- hg_keywords(hg, topics, n = 8, type = "frequency",
                           sort_by = "share", min_docs = 15)
distinctive_table <- hg_keywords(hg, topics, n = 8, type = "frequency",
                                 sort_by = "share", min_docs = 15,
                                 collapse = TRUE)
knitr::kable(distinctive_table, row.names = FALSE)
```

```{r plot-distinctive, fig.height = 13}
plot(distinctive, value = "share", ncol = 4)
```

# Four views side by side

The same sixteen topics under frequency, class-based tf-idf, PageRank
centrality in the topic's own word hypergraph (words as nodes, the topic's
abstracts as hyperedges, tf-idf weights) and, when torch is present,
HyperGAT attention, all ranked by share with the same support floor, so the
columns differ only in how the words are scored.

```{r views}
views <- hg_keywords(hg, topics, n = 6, sort_by = "share", min_docs = 15,
                     centrality = "pagerank",
                     type = c("frequency", "ctfidf", "centrality"))
```

```{r plot-views, fig.height = 30}
plot(views, value = "share")
```

```{r attention, eval = has_torch}
attention <- hg_hypergat(abstracts, labels = topics, column = "Abstract",
                         id = "doc", epochs = 10, seed = 1,
                         what = "attention")
fit <- hg_hypergat(abstracts, labels = topics, column = "Abstract", id = "doc",
                   epochs = 10, seed = 1)
mean(fit$predicted == fit$label)
hg_keywords(hg, topics, n = 8, scores = attention,
            sort_by = "share", min_docs = 15, collapse = TRUE)
```

# Topics across the pandemic years

Each abstract carries its publication year, so the partition doubles as a
lens on how the literature moved. Sixteen topics cannot be told apart by
colour, so each topic gets its own panel: the bar is the topic's share of
that year's abstracts, labelled by the three words the topic owns most.
2023 holds 105 abstracts and should be read with that in mind; 2024, with
five, is left to the count table.

```{r years, fig.height = 14}
labels <- hg_keywords(hg, topics, n = 3, type = "frequency",
                      sort_by = "share", min_docs = 15, collapse = TRUE)
year_topic <- merge(topics, subset(abstracts, select = c(doc, Year)),
                    by.x = "node", by.y = "doc")
year_n <- as.data.frame(table(Year = year_topic$Year))
shown <- subset(merge(year_topic, year_n, by = "Year"), Freq >= 30)
composition <- as.data.frame(
  prop.table(table(shown$Year, shown$cluster), margin = 1)
)
names(composition) <- c("year", "cluster", "share")
composition <- merge(composition, labels, by = "cluster")
composition$topic <- factor(composition$words, levels = labels$words)
composition$label <- sprintf("%.0f%%", 100 * composition$share)
ggplot2::ggplot(composition,
                ggplot2::aes(x = .data$year, y = .data$share)) +
  ggplot2::geom_col(fill = okabe_ito[5], width = 0.8) +
  ggplot2::geom_text(ggplot2::aes(label = .data$label), vjust = -0.4,
                     size = 3) +
  ggplot2::facet_wrap(ggplot2::vars(.data$topic), ncol = 3) +
  ggplot2::scale_y_continuous(labels = \(x) sprintf("%.0f%%", 100 * x),
                              expand = ggplot2::expansion(mult = c(0, .3))) +
  ggplot2::labs(x = "publication year", y = "share of the year's abstracts") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(strip.text = ggplot2::element_text(size = 9))
```

```{r years-table}
knitr::kable(
  as.data.frame.matrix(table(year_topic$Year, year_topic$cluster)),
  caption = "Abstracts per year and topic"
)
```

# Document scope against sentence scope, at sixteen topics

The centrality view above relates two words when they share an abstract.
`text_hypergraph(construction = "sentence")` builds the finer relation:
words are nodes and every sentence is a hyperedge, so two words are linked
only when they share a sentence. Passed to `hg_keywords()` in place of the
document hypergraph, with the same partition, it ranks each topic's words
by their centrality among the topic's sentences rather than its abstracts.
The comparison uses the same sixteen topics as the rest of the document.

```{r sentences}
sh <- text_hypergraph(abstracts, column = "Abstract", id = "doc",
                      construction = "sentence", stop_words = stops,
                      min_count = 5L, sparse = TRUE)
sh
document_scope <- hg_keywords(hg, topics, n = 8, type = "centrality",
                              sort_by = "share", min_docs = 15)
sentence_scope <- hg_keywords(sh, topics, n = 8, type = "centrality",
                              sort_by = "share", min_docs = 15)
document_scope
sentence_scope
```

```{r scopes-overlap}
shared <- merge(document_scope, sentence_scope, by = c("cluster", "word"))
overlap <- table(factor(shared$cluster, levels = unique(document_scope$cluster)))
overlap
mean(overlap)
```

The two scopes share seven or eight of eight words in every one of the
sixteen topics. Ranked by share, a topic's *owned* vocabulary is what
surfaces under either scope, and centrality only orders it; the finer
co-occurrence relation changes that order at the margin. Sentence scope
does not remove the corpus vocabulary either: ranked by raw centrality
instead of share, "learning", "students", "education" and "online" top
every topic under both scopes, because the words of the whole corpus are
also the words of most sentences.

The scope matters more, and in the other direction, for word communities.
Cutting the document-scoped word hypergraph (`nodes = "word"`) into
sixteen gives communities of a few hundred words each; cutting the
sentence-scoped one gives two large communities and fourteen small ones,
the usual spectral behaviour on a sparse relation.

```{r word-communities}
word_bag <- text_hypergraph(abstracts, column = "Abstract", id = "doc",
                            nodes = "word", stop_words = stops,
                            min_count = 5L, sparse = TRUE)
communities <- lapply(list(document = word_bag, sentence = sh), \(h) {
  parts <- hg_cluster(h, k = 16, seed = 1)
  weights <- as.data.frame(h, what = "weights")
  degree <- tapply(weights$weight, weights$word, sum)
  sizes <- sort(table(parts$cluster), decreasing = TRUE)
  data.frame(
    community = names(sizes),
    n_words = as.integer(sizes),
    top_words = vapply(names(sizes), \(cl) {
      w <- parts$node[parts$cluster == cl]
      paste(utils::head(w[order(-degree[w])], 6), collapse = ", ")
    }, character(1))
  )
})
knitr::kable(communities$document, row.names = FALSE,
             caption = "Word communities, document scope (words as nodes, abstracts as hyperedges)")
knitr::kable(communities$sentence, row.names = FALSE,
             caption = "Word communities, sentence scope (words as nodes, sentences as hyperedges)")
```

So for this corpus the sentence construction is the better *relation* but
not the better *tool*: it sharpens the ordering within a topic's owned
words slightly and it fragments word communities. Where it should pay off
is a corpus whose documents are long and mixed, where sharing a document
means little and sharing a sentence means a lot.

# Reading it

Ranked by the raw score, the sixteen topics look alike because the corpus
has one vocabulary; ranked by share with a support floor they are sixteen
different literatures, under document and sentence scope alike. The three views agree on what each topic owns, since the share
column is computed from the same partition; they differ in which owned word
comes first. The attention view, when torch is present, is trained on these
same four labels, and its agreement with them is printed above its table.
The year composition is descriptive: it reports shares, not a test of change
over time.

```{r session}
sessionInfo()
```
