---
title: "Four ways to cut the same corpus: weights and walks at sixteen topics"
author: "honets"
date: "2026-09-06"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "", fig.width = 9, dpi = 110,
                      out.width = "100%")
library(honets)
```

The default spectral cut of a text hypergraph, `type = "zhou"`, reads only
which documents each word binds: every hyperedge counts once and the
incidence cells set by `weight =` are ignored. So `weight = "tfidf"` and
`weight = "n"` give the same partition to machine precision. The cells do
enter two other places: the `type = "random_walk"` Laplacian, which walks
on them, and the hyperedge weights that `edge_weights =` adds to either
Laplacian. This document puts the four resulting cuts of the COVID-19
education corpus side by side at sixteen topics so the choice can be made
by looking, not by argument.

| Cut | Laplacian | What a word's weight does |
|---|---|---|
| `zhou` | Zhou, unit hyperedge weights | nothing: membership only |
| `zhou_idf` | Zhou, hyperedge weight = idf of the word | a word in most documents binds them less |
| `walk_counts` | random walk on raw counts | a document's heavy words pull more |
| `walk_tfidf` | random walk on tf-idf cells | heavy and rare words pull more |

# The corpus and the four cuts

```{r corpus}
abstracts <- clean_text(sbert::covid, column = "Abstract",
                        remove = "no abstract available", min_content = 0.5)
abstracts <- unique(abstracts)
abstracts$doc <- sprintf("abs_%04d", seq_len(nrow(abstracts)))
boilerplate <- c("covid", "coronavirus", "sars", "cov", "pandemic",
                 "disease", "study", "studies", "result", "results",
                 "conclusion", "conclusions", "background", "method",
                 "methods", "objective", "aim")
stops <- c(stop_words_en(), boilerplate)
tfidf <- text_hypergraph(abstracts, column = "Abstract", id = "doc",
                         weight = "tfidf", stop_words = stops,
                         min_count = 5L, sparse = TRUE)
counts <- text_hypergraph(abstracts, column = "Abstract", id = "doc",
                          weight = "n", stop_words = stops,
                          min_count = 5L, sparse = TRUE)
cuts <- list(
  zhou = hg_cluster(tfidf, k = 16, seed = 1),
  zhou_idf = hg_cluster(tfidf, k = 16, seed = 1, edge_weights = "idf"),
  walk_counts = hg_cluster(counts, k = 16, seed = 1, type = "random_walk"),
  walk_tfidf = hg_cluster(tfidf, k = 16, seed = 1, type = "random_walk")
)
```

That the default cut ignores the cells is checked, not asserted:

```{r same}
eigen_tfidf <- hg_cluster(tfidf, k = 8, seed = 1, what = "eigenvalues")
eigen_counts <- hg_cluster(counts, k = 8, seed = 1, what = "eigenvalues")
max(abs(eigen_tfidf$value - eigen_counts$value))
```

# How much they agree

Adjusted Rand index between every pair of partitions. One means the same
sixteen topics, zero means no more agreement than chance.

```{r agreement}
ari <- outer(names(cuts), names(cuts), Vectorize(\(a, b) {
  agreement <- hg_agreement(cuts[[a]], cuts[[b]])
  agreement$ari
}))
dimnames(ari) <- list(names(cuts), names(cuts))
knitr::kable(round(ari, 2), caption = "Adjusted Rand index between cuts")
```

# Balance and distinctiveness

Topic sizes, and how much of each topic's top-six owned vocabulary it
actually owns: the mean share of a word's corpus count that falls in the
topic, over the six words `sort_by = "share"` ranks first with at least 15
supporting abstracts. The weakest topic's mean share says whether every
topic has a vocabulary of its own.

```{r balance}
balance <- function(partition) {
  owned <- hg_keywords(tfidf, partition, n = 6, type = "frequency",
                       sort_by = "share", min_docs = 15)
  topic_sizes <- hg_topic_sizes(tfidf, partition)
  sizes <- topic_sizes$n
  by_topic <- tapply(owned$share, owned$cluster, mean)
  data.frame(smallest = min(sizes), largest = max(sizes),
             `size CV` = round(stats::sd(sizes) / mean(sizes), 2),
             `mean share` = round(mean(owned$share), 3),
             `weakest topic` = round(min(by_topic), 3),
             check.names = FALSE)
}
balances <- do.call(rbind, lapply(cuts, balance))
knitr::kable(balances)
```

# The sixteen topics under each cut

Each cut's topics, described by the six words they own, one table per cut.
Topic numbers are not comparable across cuts; read the words.

```{r words-zhou}
hg_keywords(tfidf, cuts$zhou, n = 6, type = "frequency",
            sort_by = "share", min_docs = 15, collapse = TRUE)
```

```{r words-zhou-idf}
hg_keywords(tfidf, cuts$zhou_idf, n = 6, type = "frequency",
            sort_by = "share", min_docs = 15, collapse = TRUE)
```

```{r words-walk-counts}
hg_keywords(tfidf, cuts$walk_counts, n = 6, type = "frequency",
            sort_by = "share", min_docs = 15, collapse = TRUE)
```

```{r words-walk-tfidf}
hg_keywords(tfidf, cuts$walk_tfidf, n = 6, type = "frequency",
            sort_by = "share", min_docs = 15, collapse = TRUE)
```

# Where the default and the idf-weighted cut differ

The two Zhou cuts agree at ARI 0.42, the highest pair. `hg_agreement(what
= "mapping")` names, for each topic of the default cut, the idf-weighted
topic that takes most of its abstracts and the share it takes: a topic that
maps at 0.9 survived the reweighting intact; one that maps at 0.4 was
split. The two keyword tables are joined on so the mapping can be read in
words.

```{r mapping}
mapping <- hg_agreement(cuts$zhou, cuts$zhou_idf, what = "mapping")
words_zhou <- hg_keywords(tfidf, cuts$zhou, n = 3, type = "frequency",
                          sort_by = "share", min_docs = 15, collapse = TRUE)
words_idf <- hg_keywords(tfidf, cuts$zhou_idf, n = 3, type = "frequency",
                         sort_by = "share", min_docs = 15, collapse = TRUE)
mapping <- merge(mapping, subset(words_zhou, select = c(cluster, words)),
                 by.x = "label_x", by.y = "cluster", sort = FALSE)
mapping <- merge(mapping, subset(words_idf, select = c(cluster, words)),
                 by.x = "label_y", by.y = "cluster", sort = FALSE,
                 suffixes = c("_zhou", "_idf"))
mapping_shown <- subset(mapping,
                        select = c(label_x, words_zhou, label_y, words_idf,
                                   share))
knitr::kable(
  mapping_shown, digits = 2, row.names = FALSE,
  col.names = c("zhou topic", "its words", "goes mostly to", "those words",
                "share kept")
)
```

# The spectra

The eigenvalue gaps that say how many clusters each Laplacian supports.
The idf weighting does not change that answer: two, then four, then
nothing.

```{r spectra}
gaps <- function(hg, ...) {
  eigen <- hg_cluster(hg, k = 8, seed = 1, what = "eigenvalues", ...)
  round(head(eigen$gap, 6), 3)
}
spectra <- rbind(
  zhou = gaps(tfidf),
  zhou_idf = gaps(tfidf, edge_weights = "idf"),
  walk_counts = gaps(counts, type = "random_walk"),
  walk_tfidf = gaps(tfidf, type = "random_walk")
)
knitr::kable(spectra, col.names = paste("gap after", 1:6))
```

# Reading it

Three of the four cuts are about equally balanced and equally distinctive;
the random walk on tf-idf is the outlier, with the most uneven sizes and a
weakest topic that owns little. The idf-weighted Zhou cut keeps most of the
default's topics and reshuffles the rest, so it is a variant of the same
partition rather than a different one. Whether that variant is better is a
reading of the two word tables, and the tables are above.

```{r session}
sessionInfo()
```
