---
title: "Hypergraphs: relations that bind more than two nodes"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is an imported dependency and the graph/plotting engine. The guard
# keeps this source document usable in stripped-down documentation sandboxes.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

> Each section below is one verb, worked end to end on the same data. The
> sections run in order in a single R session, so later ones may reuse
> objects the earlier ones built.

## build_hypergraph(): promoting cliques to group interactions {#build_hypergraph}

### The analytic question

Sometimes only pairwise data is available, but the process being studied is
plausibly a group process. Three people who all know each other might form a
genuine three-way relationship — or they might just be three separate
friendships that happen to close a triangle. A graph cannot tell you which.

`build_hypergraph()` takes the position of Burgio et al. (2020): treat each
triangle (and larger clique) as a **candidate** group interaction, and
promote it to a hyperedge with probability `p`. At `p = 1` every clique
becomes a group; at `p = 0` none do and the pairwise structure is preserved
unchanged. The parameter is an explicit statement of how much you believe
triadic closure implies group interaction.

If you *observe* group membership directly, do not infer it — use
[`group_hypergraph()`](hypergraphs.html#group_hypergraph) instead. If your data are
sequences, use [`window_hypergraph()`](hypergraphs.html#window_hypergraph).

```{r build-hypergraph-load}
library(honets)
set.seed(1)
m <- matrix(runif(100), 10, 10)
m <- (m + t(m)) / 2
diag(m) <- 0
dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
```

```{r build-hypergraph-cograph, eval = has_cograph}
library(cograph)
```

### Building one

```{r build-hypergraph-build}
hg <- build_hypergraph(m, threshold = 0.5)
hg
```

24 pairwise edges retained as 2-hyperedges, plus 15 triangles promoted to
3-hyperedges.

```{r build-hypergraph-table}
as.data.frame(hg, top = 6)
```

`weight` is `NA` here: hyperedges from cliques are unweighted. Only
[`window_hypergraph()`](hypergraphs.html#window_hypergraph) produces weights, from window
counts.

### Seeing it

A hypergraph has no direct network drawing, but `clique_expansion()`
projects it back to a pairwise network that cograph plots directly. Read it
knowing what the projection destroys: every hyperedge became a clique, and
the group boundaries are gone.

```{r build-hypergraph-hgplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "Clique expansion of the hypergraph"}
net <- clique_expansion(hg)
splot(net, minimum = 0.05,
      title = "Hypergraph, projected to a pairwise network")
```

### `p` is the modelling assumption

```{r build-hypergraph-p}
do.call(rbind, lapply(c(0, 0.25, 0.5, 0.75, 1), function(p) {
  h <- build_hypergraph(m, threshold = 0.5, p = p, seed = 1)
  edges_h <- as.data.frame(h)
  data.frame(p = p, hyperedges = nrow(edges_h),
             pairwise = sum(edges_h$size == 2),
             triples = sum(edges_h$size == 3))
}))
```

At `p = 0` you get 24 hyperedges, all of size 2 — the original graph,
re-expressed. At `p = 1`, all 15 triangles are promoted. In between, the
sampling is Bernoulli per clique, so **`seed` is required for
reproducibility**:

```{r build-hypergraph-seed}
first <- build_hypergraph(m, threshold = 0.5, p = 0.5, seed = 42)
second <- build_hypergraph(m, threshold = 0.5, p = 0.5, seed = 42)
identical(first, second)
```

Never report a result at `0 < p < 1` from a single draw. Repeat across seeds
and report the variability, exactly as you would for any stochastic method.

```{r build-hypergraph-seedvar}
sizes <- vapply(1:20, function(s) {
  h <- build_hypergraph(m, threshold = 0.5, p = 0.5, seed = s)
  edges_h <- as.data.frame(h)
  nrow(edges_h)
}, integer(1))
range(sizes)
```

### `include_pairwise` — a hypergraph of groups only

```{r build-hypergraph-nopairs}
build_hypergraph(m, threshold = 0.5, include_pairwise = FALSE)
```

`FALSE` drops the 2-hyperedges, giving a "fully higher-order" hypergraph
containing only the k ≥ 3 interactions. Use it to isolate the group structure
from the dyadic structure — for instance to check whether a centrality result
survives when the pairwise scaffolding is removed.

### `max_size` and `threshold`

```{r build-hypergraph-maxsize}
do.call(rbind, lapply(c(3, 4, 5), function(ms) {
  h <- build_hypergraph(m, threshold = 0.5, max_size = ms)
  edges_h <- as.data.frame(h)
  data.frame(max_size = ms, hyperedges = nrow(edges_h))
}))
```

`max_size = 3` (the default) keeps triangles. Raising it to 4 finds two
4-cliques; there are no 5-cliques, so 5 adds nothing. Unlike the Vietoris-Rips
complex in [`build_simplicial()`](simplicial-complexes.html#build_simplicial), this does **not**
explode combinatorially, because only maximal-clique membership is
enumerated rather than every face.

`threshold` binarises the weight matrix before clique-finding, and it moves
the result far more than `max_size` does:

```{r build-hypergraph-threshold}
do.call(rbind, lapply(c(0.3, 0.5, 0.7), function(t) {
  h <- build_hypergraph(m, threshold = t)
  edges_h <- as.data.frame(h)
  data.frame(threshold = t, hyperedges = nrow(edges_h),
             pairwise = sum(edges_h$size == 2),
             triples = sum(edges_h$size == 3))
}))
```

### `type = "vr"` raises rather than pretending

```{r build-hypergraph-vr, error = TRUE}
build_hypergraph(m, type = "vr")
```

A genuine metric Vietoris-Rips construction needs a distance-ball graph,
which this verb does not build. Rather than silently aliasing `"clique"` and
giving you a result labelled something it is not, it stops. Use
[`build_simplicial(type = "vr")`](simplicial-complexes.html#build_simplicial) if you want a real
VR construction.

### What to do next

- **[`hypergraph_measures()`](hypergraphs.html#hypergraph_measures)** — hyperdegree,
  overlap and density for the object you just built.
- **[`hypergraph_centrality()`](hypergraphs.html#hypergraph_centrality)** — which nodes
  matter, by tensor eigenvector or EDVW PageRank.
- **[`clique_expansion()`](hypergraphs.html#clique_expansion)** — project back to a
  pairwise network, and see what the round trip costs.
- **[`group_hypergraph()`](hypergraphs.html#group_hypergraph)** — when membership is
  observed rather than inferred.

### References

- Burgio, G., Matamalas, J. T., Gómez, S., & Arenas, A. (2020). Evolution of
  cooperation in the presence of higher-order interactions: from networks to
  hypergraphs. *Entropy*, 22(7), 744.
- Battiston, F., et al. (2020). Networks beyond pairwise interactions.
  *Physics Reports*, 874, 1-92.


---

## Windowed sequence hyperedges: multi-way co-occurrence in coded sequences {#window_hypergraph}

### The analytic question

Coded interaction sequences record one action at a time: a person specifies
a requirement, the partner interrupts, a command follows. A transition
network answers *what follows what*; a co-occurrence network answers *which
pairs appear together*. Neither answers the question this tutorial takes
up: **which actions occur together as groups within short spans of
activity, and do those groups recur often enough to constitute a
structure?**

A pairwise network cannot hold the answer, because a group of three
actions inside one span is not the same observation as its three pairs.
The hypergraph keeps the group intact: each short span of a sequence
becomes one *hyperedge* over the distinct actions inside it. The
construction follows Ding, Wang, Li, Li & Liu (2020); the resulting
weighted incidence matrix carries edge-dependent vertex weights in the
sense of Chitra & Raphael (2019), which the spectral machinery of Zhou,
Huang & Schölkopf (2006) and Hayashi, Aksoy, Park & Park (2020) consumes
directly.

The data are `human_long`, bundled with honets: 10,796 coded human
turns from 429 human–AI pair-programming sessions, nine action codes, each
code carrying an a-priori cluster label (Directive, Evaluative,
Metacognitive) that the analysis below never uses — it will serve as an
external check at the end.

```{r window-hypergraph-load}
library(honets)
```

```{r window-hypergraph-cograph, eval = has_cograph}
library(cograph)
```

### How a sequence becomes a hypergraph

Before the real data, a four-event sequence small enough to verify by
hand. With `window = 2` the windows are (a,b), (b,a), (a,c); the first two
contain the same set of states, so they collapse into one hyperedge
observed twice.

```{r window-hypergraph-tiny}
hg_tiny <- window_hypergraph(list(s1 = c("a", "b", "a", "c")), window = 2L)
as.data.frame(hg_tiny)
```

The table has one row per hyperedge. The set {a, b} was seen in 2 windows
(`weight = 2`), the set {a, c} in 1. Weight is therefore *recurrence*: how
many spans of the sequence produced exactly this combination. Repeats
inside a window count as occurrences in the incidence matrix but not as
extra members — a window (a, a) yields the singleton hyperedge {a} with an
incidence cell of 2. This is what makes the incidence matrix
edge-dependent-weighted rather than binary.

### Recurrent action combinations in real sessions

Which combinations of actions recur inside three-turn spans of the
sessions? The long format goes in directly: `action` names the code
column, `actor` groups turns into sessions, `time` orders them.

```{r window-hypergraph-full}
hg_all <- window_hypergraph(human_long, action = "code",
                            actor = "session_id", time = "timestamp",
                            window = 3L)
hg_all
```

The 429 sessions produce 9,942 windows (six sessions were shorter than
three turns and contributed none — the constructor counts them rather than
dropping them silently). The 129 hyperedges are exactly the 9 singletons,
36 pairs, and 84 triples that nine codes allow: with this much data,
*every possible combination occurs at least once*. Presence alone is
saturated, so any presence-based summary would be flat. The information is
in the weights:

```{r window-hypergraph-top10}
as.data.frame(hg_all, sort_by = "weight", top = 10)
```

The distribution is far from flat. `Command, Specify` dominates with 722
windows, followed by `Command, Request, Specify` (583). Specify appears in
eight of the top nine combinations: specification is the connective tissue
of these sessions. The gap between the strongest combination (722) and the
weakest (1) is what the saturated presence structure hides.

Since rare combinations are noise for structural analysis, the
`min_weight` argument keeps only recurrent ones — the same role `min_freq`
plays in higher-order rule extraction. Fifty windows out of 9,942 is a
conservative floor; `min_size = 2` also removes the nine singletons, which
carry no co-occurrence information.

```{r window-hypergraph-recurrent}
hg <- window_hypergraph(human_long, action = "code",
                        actor = "session_id", time = "timestamp",
                        window = 3L, min_size = 2L, min_weight = 50L)
hg
```

Fifty-six combinations — 22 pairs and 34 triples — recur at least 50
times. This is the structure the rest of the analysis works on.

### Seeing the recurrent combinations

```{r window-hypergraph-whgplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "Clique expansion of the window hypergraph"}
net_all <- clique_expansion(hg_all)
splot(net_all, minimum = 0.05,
      title = "Windowed co-occurrence, projected")
```

### How concentrated is the recurrent structure?

```{r window-hypergraph-measures}
hypergraph_measures(hg)
```

Three numbers matter here. First, hyperdegree now ranges from 8 to 33
(median 15): after filtering, the codes participate in very different
numbers of recurrent combinations, where the unfiltered hypergraph had
forced every code to the same value. Second, pairwise participation is
0.92 — almost every pair of codes still shares at least one recurrent
hyperedge, so the structure is concentrated but not fragmented. Third, the
intersection profile shows 264 hyperedge pairs overlapping in two codes:
the triples are largely *extensions* of the strong pairs (for example
`Command, Specify` extending to `Command, Request, Specify`), not separate
formations.

### Which actions anchor the combinations?

The H-eigenvector centrality of Benson (2019) scores each node on the
uniform-size hypergraph tensor; unlike degree it rewards being in
combinations whose *other members* are themselves central.

```{r window-hypergraph-centrality}
hypergraph_centrality(hg, type = "H")
```

Specify (0.43) is the clear anchor, with Command (0.38) second — the two
directive actions sit inside most strong combinations. Verify (0.26),
Refine (0.28), and Interrupt (0.28) are peripheral: they occur, and recur,
but in combinations anchored by others. The ordering matches what the
top-weight table already suggested, now as a single per-node score.

### Do the actions organize into higher-order groups?

The random-walk Laplacian of Hayashi et al. (2020) uses both weight
layers this hypergraph carries: the occurrence-count incidence
(edge-dependent vertex weights) and the window counts, which
`hypergraph_cluster()` picks up automatically as hyperedge weights.
Hayashi et al. provide two clustering branches: RDC-Spec uses the smallest
Laplacian eigenvectors and k-means, while RDC-Sym factorizes the normalized
similarity as `T ~= U U'` and assigns each node to its largest factor. Both
are available through `algorithm =`. Spectral k-means is stochastic, so the
run is seeded and repeated under a second seed as a stability check.

```{r window-hypergraph-cluster}
cl <- hypergraph_cluster(hg, k = 2L, type = "random_walk", seed = 1)
cl_check <- hypergraph_cluster(hg, k = 2L, type = "random_walk", seed = 99)
membership <- as.data.frame(cl)
membership_check <- as.data.frame(cl_check)
identical(membership, membership_check)
membership
```

```{r window-hypergraph-symnmf}
cl_sym <- hypergraph_cluster(
  hg, k = 2L, type = "random_walk", algorithm = "symnmf",
  nstart = 5L, seed = 1
)
cl_sym
as.data.frame(cl_sym)
```

RDC-Sym exposes the final Eq. 16 loss, its objective trace, iteration count,
and convergence flag under `cl_sym$params`. The paper's J-NMF and JS-NMF
methods are also implemented as `hypergraph_joint_cluster()`, but require a
separate node-by-node relation matrix (patent citations in the paper). They
are intentionally not invoked here because this sequence data has no such
second view.

The partition is identical under both seeds, so it can be reported as
stable. Cluster 2 is {Correct, Frustrate, Refine}; Cluster 1 is everything
else, with Specify holding the largest stationary probability
(`pi = 0.31`) — a random walker on the recurrent combinations spends
almost a third of its time on Specify.

This is where the withheld labels come in: the coding scheme's Evaluative
cluster is {Correct, Frustrate, Refine, Verify}. The hypergraph recovered
three of its four members from co-occurrence structure alone. The
defector is informative: Verify co-occurs with Specify and Command (the
combinations `Command, Specify, Verify` and `Request, Specify, Verify`
each recur over 80 times), so *within short spans* verification behaves
as part of the directive workflow, not the evaluative one.

```{r window-hypergraph-clusterplot, fig.width = 12, fig.height = 5, fig.alt = "Spectral embedding of the nine action codes, colored by cluster"}
plot(cl)
```

The embedding shows the same picture geometrically: the corrective codes
separate on the second spectral dimension while Specify sits at the center
of the main group.

### Does the window size change the conclusions?

Window size is the one consequential arbitrary choice in this pipeline, so
the headline result must survive a different value.

```{r window-hypergraph-sensitivity}
hg4 <- window_hypergraph(human_long, action = "code",
                         actor = "session_id", time = "timestamp",
                         window = 4L, min_size = 2L, min_weight = 50L)
as.data.frame(hg4, sort_by = "weight", top = 8)
```

With four-turn windows the same combinations lead: `Command, Request,
Specify` (447), `Command, Interrupt, Specify` (357), `Command, Specify`
(343), and their four-code extension `Command, Interrupt, Request,
Specify` now enters at 241 — wider windows admit larger sets, but the
membership of the dominant combinations is unchanged. The conclusion —
directive combinations anchored by Specify dominate, corrective actions
form their own block — does not depend on the choice of three versus four.

### The same sessions as a temporal hypergraph

`window_hypergraph()` answers which combinations recur, and forgets *when*
in a session they occurred. `temporal_hypergraph()` keeps that: read the
same log with the session as the hyperedge, the code as the member and
the turn's position as its time, and every session becomes one hyperedge
whose members enter at their step. The clock is order. Here
`order_in_session` counts every turn of the session, the partner's
included, so a human turn's step is its place among all turns. A wide
table of states, one row per session and one column per position, needs
no column names at all.

```{r window-hypergraph-temporal}
sessions <- temporal_hypergraph(human_long, actor = "code",
                                group = "session_id",
                                time = "order_in_session")
sessions
```

A snapshot with `window = 3` at step 5 is the co-occurrence hypergraph of
steps 5 to 7: one hyperedge per session still running, over the codes of
its human turns among those three turns. Most sessions show one or two
codes there, because the partner holds some of the three turns;
`window_hypergraph()` above windows over consecutive human turns instead,
which is why its sets are larger.

```{r window-hypergraph-temporal-snapshot}
steps_five_to_seven <- hypergraph_snapshot(sessions, at = 5, window = 3)
steps_five_to_seven
```

`mode = "cumulative"` at a step is what each session had shown by then,
and `hg_growth()` follows the memberships as they enter.

```{r window-hypergraph-temporal-growth}
growth_sessions <- hg_growth(sessions, mode = "cumulative")
head(growth_sessions, 5)
```

### When to use which

- **`temporal_hypergraph()`** — sequences, and the question is about
  *when* combinations occur: what a session had shown by a step, or on a
  span of steps, with the session kept intact as one hyperedge.
- **`window_hypergraph()`** — sequences, and the question is about
  combinations *local in time*. `step = 1` (sliding) weights every span;
  `step = window` (tumbling) counts disjoint spans and suits length
  comparisons across groups. Raise `min_weight` before structural
  analysis on dense data; this tutorial's saturation of all 129 possible
  sets is the norm, not the exception, once windows are plentiful.
- **`group_hypergraph()`** — the grouping is already categorical (a whole
  session, a document, a team) and position in time is irrelevant. A
  whole-sequence window reproduces it exactly.
- **`build_hypergraph()`** — the input is an estimated pairwise network
  and the hyperedges should be its cliques.
- **`clique_expansion()`** — hands the hypergraph back to pairwise
  machinery in cograph when a graph method is needed; the
  multi-way distinctions collapse into edge weights at that point, which
  is exactly what this tutorial's construction exists to avoid.

### References

- Benson, A. R. (2019). Three hypergraph eigenvector centralities.
  *SIAM Journal on Mathematics of Data Science*, 1(2), 293–312.
- Chitra, U., & Raphael, B. J. (2019). Random walks on hypergraphs with
  edge-dependent vertex weights. *ICML 2019*, PMLR 97, 1172–1181.
- Ding, K., Wang, J., Li, J., Li, D., & Liu, H. (2020). Be more with less:
  Hypergraph attention networks for inductive text classification.
  *EMNLP 2020*, 4927–4936.
- Hayashi, K., Aksoy, S. G., Park, C. H., & Park, H. (2020). Hypergraph
  random walks, Laplacians, and clustering. *CIKM 2020*, 495–504.
- Zhou, D., Huang, J., & Schölkopf, B. (2006). Learning with hypergraphs:
  Clustering, classification, and embedding. *NeurIPS 2006*, 1601–1608.


---

## group_hypergraph(): when membership is observed {#group_hypergraph}

### The analytic question

A great deal of data is *already* group data. Authors on a paper. Students in
a seminar. Codes applied in one session. Items in one basket. The group
membership is recorded directly — nothing needs inferring.

The usual move is to project it to a pairwise network: everyone in a group
gets connected to everyone else. That projection **destroys information**
irreversibly. A five-author paper and ten separate two-author papers can
produce exactly the same graph, and no analysis of the graph can tell them
apart.

`group_hypergraph()` skips the projection. Each group becomes one hyperedge
spanning all its members, and the group size survives into the analysis.

```{r group-hypergraph-load}
library(honets)
```

```{r group-hypergraph-cograph, eval = has_cograph}
library(cograph)
```

### Building one

Long-format data: one row per (member, group) observation.

```{r group-hypergraph-build}
hg <- group_hypergraph(human_long, actor = "code", group = "session_id")
hg
```

429 sessions become 429 hyperedges over the 9 action codes. The size
distribution is the thing a pairwise projection would have thrown away: **80
sessions used all nine codes**, while five used only one.

```{r group-hypergraph-table}
as.data.frame(hg, top = 4)
```

### A small worked example

```{r group-hypergraph-small}
df <- data.frame(
  person  = c("Alice", "Bob", "Carol", "Alice", "Bob", "Dave", "Carol",
              "Dave", "Eve"),
  session = c("S1", "S1", "S1", "S2", "S2", "S3", "S3", "S3", "S3"),
  stringsAsFactors = FALSE
)
small <- group_hypergraph(df, actor = "person", group = "session")
small
as.data.frame(small)
```

Three sessions, three hyperedges: a triple, a pair, and another triple. Note
that S1 = {Alice, Bob, Carol} is **one** relation, not three.

### Seeing it

```{r group-hypergraph-grpplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "Clique expansion of the group hypergraph"}
net <- clique_expansion(hg)
splot(net, minimum = 0.05,
      title = "Group co-membership, projected")
```

### Weighted membership

If the data record a quantity per (member, group) pair — hours contributed,
messages sent, counts — pass the column as `weight` and the incidence matrix
carries it.

```{r group-hypergraph-weighted}
dfw <- data.frame(
  person  = c("A", "B", "A", "A", "B"),
  grp     = c("g1", "g1", "g1", "g2", "g2"),
  n       = c(2, 5, 3, 1, 4),
  stringsAsFactors = FALSE
)
hw <- group_hypergraph(dfw, actor = "person", group = "grp", weight = "n")
as.data.frame(hw, what = "memberships")
```

Repeated `(member, group)` pairs are **summed**: A appears twice in `g1` with
2 and 3, giving 5. Without `weight`, the incidence is 0/1 and repeats simply
collapse:

```{r group-hypergraph-binary}
hb <- group_hypergraph(dfw, actor = "person", group = "grp")
as.data.frame(hb, what = "memberships")
```

Those edge-dependent vertex weights are exactly what
[`hypergraph_centrality(type = "pagerank")`](hypergraphs.html#hypergraph_centrality) and
the [Laplacian family](hypergraphs.html#hypergraph_laplacian) consume — a weighted
hypergraph carries information a binary one does not.

### Missing values are dropped silently

```{r group-hypergraph-na}
dfn <- data.frame(
  person = c("A", "B", NA, "C", "D"),
  grp    = c("g1", "g1", "g1", NA, "g2"),
  stringsAsFactors = FALSE
)
group_hypergraph(dfn, actor = "person", group = "grp")
```

Rows with `NA` in either column are removed before construction. Check your
own missingness before calling — this verb will not warn you about it, and a
silently shrunken hypergraph is easy to miss.

### What the group structure buys you

```{r group-hypergraph-structure}
m <- hypergraph_measures(hg)
as.data.frame(m, what = "global")
```

**Pairwise participation is 1.0**: every pair of codes co-occurs in at least
one session. A projected pairwise network would therefore be the *complete
graph on nine nodes* — utterly uninformative, exactly as the first-order
transition network is in the
[three-families comparison](overview.html#three_families).

The hypergraph keeps its information because the hyperedges remain
distinguishable by size and membership:

```{r group-hypergraph-sizes}
hg_measures(hg, what = "distribution", measure = "size")
```

### Naming

The argument is `member`, not `player`: the values become hypergraph nodes
and need not be people — codes, items, genes and words are all common. The
verb is named for what it returns (a hypergraph) rather than for its input
format.

### What to do next

- **[`hypergraph_measures()`](hypergraphs.html#hypergraph_measures)** — hyperdegree,
  overlap structure and density.
- **[`hypergraph_centrality()`](hypergraphs.html#hypergraph_centrality)** — which members
  are central across the groups.
- **[`hypergraph_cluster()`](hypergraphs.html#hypergraph_laplacian)** — group the members
  by the hypergraph's own random walk.
- **[`clique_expansion()`](hypergraphs.html#clique_expansion)** — the projection this
  verb avoids, when you deliberately want it.
- **[`window_hypergraph()`](hypergraphs.html#window_hypergraph)** — the sequence-based
  route to a hypergraph, when groups are local in time rather than given.

### References

- Perc, M., Gómez-Gardeñes, J., Szolnoki, A., Floría, L. M., & Moreno, Y.
  (2013). Evolutionary dynamics of group interactions on structured
  populations: a review. *Journal of the Royal Society Interface*, 10(80),
  20120997.
- Battiston, F., et al. (2020). Networks beyond pairwise interactions.
  *Physics Reports*, 874, 1-92.


---

## hypergraph_measures(): structure without a projection {#hypergraph_measures}

### The analytic question

Graph descriptives — degree, density, clustering — all assume relations come
in pairs. Applied to a projected hypergraph they answer questions about the
projection rather than about the data.

`hypergraph_measures()` computes the equivalents directly on the hypergraph:
how many groups is each node in, how large are the groups, and how much do
the groups overlap. The last of these has no graph analogue at all, and is
usually the most informative.

```{r hypergraph-measures-load}
library(honets)
hg <- group_hypergraph(human_long, actor = "code", group = "session_id")
m <- hypergraph_measures(hg)
m
```

```{r hypergraph-measures-cograph, eval = has_cograph}
library(cograph)
```

### Per-node measures

```{r hypergraph-measures-nodes}
as.data.frame(m, sort_by = "hyperdegree")
```

- **`hyperdegree`** — how many hyperedges contain this node. Command appears
  in 423 of 429 sessions.
- **`node_strength`** — the summed incidence weight. On a binary incidence
  this is a weighted count; the gap between it and hyperdegree tells you how
  much the weights matter.
- **`max_edge_size`** — the largest group this node ever belonged to. All 9
  here, because every code appears in at least one nine-code session.

The spread is narrow: 423 down to 218, a ratio under 2. Every action is
present in most sessions, so **hyperdegree does not discriminate strongly** —
worth knowing before building an argument on it. Compare with
[`hypergraph_centrality()`](hypergraphs.html#hypergraph_centrality), which uses the
overlap structure rather than raw membership counts.

### Whole-hypergraph measures

```{r hypergraph-measures-global}
as.data.frame(m, what = "global")
```

**Density 0.729** is the fraction of possible (node, hyperedge) memberships
that exist — a hypergraph analogue of graph density, and high here because
sessions use most of the code set.

**Pairwise participation 1.0** is the one to notice. It is the fraction of
node *pairs* that co-occur in at least one hyperedge, and it is at ceiling:
every pair of codes shares a session. That is precisely why a pairwise
projection of these data is uninformative — it would be the complete graph.

### Hyperedge sizes

```{r hypergraph-measures-edges}
as.data.frame(m, what = "edges", top = 5)
hg_measures(hg, what = "distribution", measure = "size")
```

The size distribution is the structure a projection destroys. Here it is
strongly right-skewed: most sessions use six or more of the nine codes.

### Seeing the structure

The per-node table becomes a picture two ways: the co-membership weights as
a heatmap, and the projected network itself. The heatmap is the honest one —
it shows the pairwise weights the measures are computed from without
implying the hypergraph *was* a graph.

```{r hypergraph-measures-heat, eval = has_cograph, fig.width = 8, fig.height = 7, fig.alt = "Heatmap of pairwise co-membership weights"}
net <- clique_expansion(hg)
plot_heatmap(net)
```

```{r hypergraph-measures-hgmplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "Clique expansion of the hypergraph"}
splot(net, minimum = 0.05,
      title = "Co-membership, projected to a pairwise network")
```

### The intersection profile

The most distinctively hypergraph-shaped output, shown by `print()`: how many
*pairs of hyperedges* share exactly k members.

```{r hypergraph-measures-profile}
m
```

96 pairs of sessions share no codes at all; 20,355 pairs share five. Read as
a distribution, it says these sessions are **highly overlapping** — the
typical pair has five codes in common. A collection of specialised, disjoint
groups would put its mass at the low end instead.

This is a property of the *set system*. No pairwise projection retains it.

### A contrasting example

```{r hypergraph-measures-contrast}
disjoint <- group_hypergraph(
  data.frame(member = c("A","B","C","D","E","F"),
             grp    = c("g1","g1","g2","g2","g3","g3"),
             stringsAsFactors = FALSE),
  actor = "member", group = "grp")
md <- hypergraph_measures(disjoint)
as.data.frame(md, what = "global")
md
```

Three disjoint pairs: density 0.33, pairwise participation 0.2, and every
hyperedge pair shares zero members. Compare against the sessions above to see
how far apart two hypergraphs can be on measures that a graph would flatten.

### On an inferred hypergraph

```{r hypergraph-measures-inferred}
set.seed(1)
mat <- matrix(runif(100), 10, 10)
mat <- (mat + t(mat)) / 2
diag(mat) <- 0
dimnames(mat) <- list(LETTERS[1:10], LETTERS[1:10])
hg_thr <- build_hypergraph(mat, threshold = 0.5)
mi <- hypergraph_measures(hg_thr)
as.data.frame(mi, what = "global")
as.data.frame(mi, sort_by = "hyperdegree", top = 4)
```

Clique-derived hypergraphs have much lower density, because most hyperedges
are size 2 or 3 rather than spanning most of the node set.

### What to do next

- **[`hypergraph_centrality()`](hypergraphs.html#hypergraph_centrality)** — ranking
  nodes, where hyperdegree does not separate them.
- **[`hypergraph_cluster()`](hypergraphs.html#hypergraph_laplacian)** — grouping nodes
  using the hypergraph's random walk.
- **[`clique_expansion()`](hypergraphs.html#clique_expansion)** — the projection, and
  what it costs.
- **[`q_analysis()`](simplicial-complexes.html#q_analysis)** — the overlap question again, in the
  simplicial family, where subset closure is assumed.

### References

- Lee, G., Choe, M., & Shin, K. (2024). A survey on hypergraph
  representation, learning and mining. *Data Mining & Knowledge Discovery*,
  37, 1-39.
- Do, M. T., Yoon, S., Hooi, B., & Shin, K. (2020). Structural patterns and
  generative models of real-world hypergraphs. *KDD 2020*, 176-186.


---

## Hypergraph PageRank: where the interaction flow spends its time {#hypergraph_centrality}

### The analytic question

The [windowed-hyperedge tutorial](hypergraphs.html#window_hypergraph) established which
action combinations recur in the `human_long` pair-programming sessions and
which actions anchor them structurally (H-eigenvector centrality). This
tutorial asks a different question: **if session activity is treated as a
flow that wanders from action to action through the recurrent
combinations, where does that flow spend its time?**

Structural centrality and flow centrality can disagree. An action can sit
in many *distinct* combinations (structurally well-placed) while the
combinations it sits in are rarely *travelled* (little flow). The
instrument for the flow question is the random walk on hypergraphs with
edge-dependent vertex weights of Chitra & Raphael (2019), damped in the
PageRank manner: from an action, the walk picks one of its recurrent
combinations with probability proportional to the combination's window
count, then a member action with probability proportional to its
occurrences inside that combination; with probability 0.15 it teleports to
a uniformly random action. `hypergraph_centrality(type = "pagerank")`
returns the stationary distribution of this walk.

One theoretical point makes the hypergraph version worth having: Chitra &
Raphael prove that with edge-*independent* vertex weights the walk
collapses to an ordinary PageRank on a weighted pairwise graph. The
windowed construction is exactly the case that does *not* collapse — its
incidence cells are within-window occurrence totals, which differ across
combinations — so the scores below use information a pairwise reduction
would destroy.

```{r hypergraph-centrality-load}
library(honets)
```

```{r hypergraph-centrality-cograph, eval = has_cograph}
library(cograph)
```

### The flow scores

The hypergraph is the same recurrent structure as in the previous
tutorial: three-turn sliding windows, singletons removed, combinations
kept when they recur in at least 50 windows.

```{r hypergraph-centrality-build}
hg <- window_hypergraph(human_long, action = "code",
                        actor = "session_id", time = "timestamp",
                        window = 3L, min_size = 2L, min_weight = 50L)
hypergraph_centrality(hg, type = "pagerank")
```

The walk spends 27% of its time on Specify and 17% on Command — together
almost half of all flow on the two core directive actions. Request is
third at 11%. At the other end, Verify receives 4.6%, barely above the
teleportation floor: verification is visited, but the recurrent
combinations that contain it carry little traffic. The window counts
(hyperedge weights) and occurrence totals (vertex weights) are picked up
from the hypergraph automatically; no further arguments are needed.

### What recurrence adds to the ranking

Setting every combination's weight to 1 removes the recurrence
information while keeping the combination structure — the walk then treats
a combination seen 722 times and one seen 50 times as equal.

```{r hypergraph-centrality-unitweights}
hypergraph_centrality(hg, type = "pagerank", edge_weights = 1)
```

Two movements stand out against the recurrence-weighted scores above.
Frustrate rises from 9.2% to 12.2% and Correct from 8.0% to 10.1%: these
actions appear in *many distinct* combinations, so equal weighting favors
them. Specify falls from 27.5% to 21.3%. The comparison separates two
kinds of prominence — diversity of contexts (unit weights) versus volume
of traffic (window counts) — and the default deliberately measures the
latter.

### Seeing the flow

```{r hypergraph-centrality-hgcplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "Clique expansion of the hypergraph, showing where the walk flows"}
net <- clique_expansion(hg)
splot(net, minimum = 0.05,
      title = "The co-membership structure the walk moves on")
```

### Flow versus structure

The H-eigenvector scores the same hypergraph on presence alone: which
actions sit in combinations whose other members are themselves central.

```{r hypergraph-centrality-flowstructure}
hypergraph_centrality(hg, type = c("H", "pagerank"), sort_by = "pagerank")
```

The table is ordered by PageRank, so reading down the `H` column shows
where the two measures part company. They agree at the top (Specify, then
Command) and at the bottom (Verify last) but disagree in the middle, and
the disagreement is informative. H's third-highest score goes to
Frustrate (0.36), and it puts Correct and Inquire (0.33) above Request
(0.30); PageRank reverses this, placing Request third (0.11) well above
Correct and Inquire (0.08). The reason is the
weight layer: Request sits inside the heaviest combinations in the data —
`Command, Request, Specify` alone recurs 583 times — while Frustrate,
Correct, and Inquire spread across many lighter ones. Structure says
Frustrate is well-connected; flow says the traffic runs through Request.

### Is the ranking an artifact of the damping choice?

Damping is the one free parameter, so the conclusion must survive
changing it. Two calls bracket the default of 0.85.

```{r hypergraph-centrality-damping}
hypergraph_centrality(hg, type = "pagerank", damping = 0.5)
hypergraph_centrality(hg, type = "pagerank", damping = 0.95)
```

The ordering of all nine actions is identical at damping 0.5, 0.85, and
0.95; only the spread changes. Less teleportation concentrates mass on
the structure (Specify grows from 20.1% at 0.5 to 29.8% at 0.95, Verify
shrinks from 7.7% to 3.5%), but no two actions trade places. The ranking
can be reported without a damping caveat. In the undamped limit the
scores converge to the walk's stationary distribution — the `pi` column
that [hypergraph_cluster()] reports, where Specify likewise holds 0.31.

### When to use which

- **`type = "pagerank"`** — the question is about *flow or traffic*:
  where activity concentrates, weighted by how often combinations recur.
  It uses both weight layers of a [window_hypergraph()], tolerates
  disconnected structure (teleportation), and its damping default can be
  reported without sensitivity caveats when the ranking is stable, as
  here.
- **`type = "H"` / `"Z"`** — the question is about *structural
  position* in the combination pattern, deliberately ignoring recurrence.
  Divergence from PageRank is itself a finding: it locates actions whose
  prominence is contextual diversity rather than volume.
- **`type = "clique"`** — a pairwise baseline (eigenvector centrality on
  the clique expansion); useful to check whether the hypergraph adds
  anything beyond a graph. When the incidence is binary, PageRank too
  collapses to a pairwise computation — the collapse theorem — so on
  binary hypergraphs prefer the cheaper graph method and reserve the
  hypergraph walk for genuinely edge-dependent weights.
- **`pi` from [hypergraph_cluster()]** — the undamped stationary
  distribution; appropriate when the hypergraph is connected and no
  teleportation smoothing is wanted.

### References

- Benson, A. R. (2019). Three hypergraph eigenvector centralities.
  *SIAM Journal on Mathematics of Data Science*, 1(2), 293–312.
- Chitra, U., & Raphael, B. J. (2019). Random walks on hypergraphs with
  edge-dependent vertex weights. *ICML 2019*, PMLR 97, 1172–1181.
- Hayashi, K., Aksoy, S. G., Park, C. H., & Park, H. (2020). Hypergraph
  random walks, Laplacians, and clustering. *CIKM 2020*, 495–504.


---

## The hypergraph Laplacian family {#hypergraph_laplacian}

### The analytic question

Spectral methods — clustering, label propagation, diffusion — all run off a
Laplacian. Defining one for a hypergraph is not automatic, because a
hyperedge connects many nodes at once and there is no single obvious way to
distribute that connection.

This family implements two answers:

- **`type = "zhou"`** (default) — the weighted normalised Laplacian of Zhou,
  Huang & Schölkopf (2006). Each hyperedge spreads its weight evenly over its
  members.
- **`type = "random_walk"`** — the edge-dependent vertex weight walk of
  Hayashi et al. (2020). A step picks a hyperedge in proportion to its
  weight, then a node within it in proportion to that node's weight *in that
  hyperedge*. When those within-edge weights vary, this carries information
  the Zhou construction averages away.

```{r hypergraph-laplacian-load}
library(honets)
hg <- window_hypergraph(human_long, action = "code", actor = "session_id",
                        time = "timestamp",
                        window = 3L, min_size = 2L, min_weight = 50L)
hg
```

### The Laplacian itself

```{r hypergraph-laplacian-lap}
L <- hypergraph_laplacian(hg)
dim(L)
round(L[1:4, 1:4], 3)
```

It is a symmetric 9×9 matrix with the properties a Laplacian should have:

```{r hypergraph-laplacian-props}
ev <- sort(eigen(L, symmetric = TRUE, only.values = TRUE)$values)
round(head(ev, 4), 5)
```

The smallest eigenvalue is exactly 0 — the constant vector is always in the
kernel. The **second** eigenvalue (0.548) is the spectral gap: how hard it is
to cut this hypergraph in two. A large gap means well-connected, hard to
split.

The stationary distribution of the associated walk comes along as an
attribute, and the [PageRank verb](hypergraphs.html#hypergraph_centrality) is the damped
version of the same walk:

```{r hypergraph-laplacian-pi}
round(sort(attr(L, "pi"), decreasing = TRUE), 4)
```

### Clustering

```{r hypergraph-laplacian-cluster}
cl <- hypergraph_cluster(hg, k = 3, seed = 1)
cl
```

```{r hypergraph-laplacian-clustertable}
as.data.frame(cl)
```

Three clusters of three. Reading the membership: **Command, Inquire and
Interrupt** group together; **Correct, Frustrate and Refine** form a second;
**Request, Specify and Verify** a third. The second is recognisably a
repair-and-revision cluster, the third a specify-and-check cluster.

`summary()` gives the sizes:

```{r hypergraph-laplacian-clsummary}
summary(cl)
```

#### Check the eigengap before trusting `k`

The print method reports **Eigengap after k: 0.0380**. That is small. A
well-separated `k`-cluster structure produces a large gap after the `k`-th
eigenvalue; a small one means the choice of `k` is not strongly supported by
the spectrum.

```{r hypergraph-laplacian-eigengap}
do.call(rbind, lapply(2:5, function(k) {
  c_k <- hypergraph_cluster(hg, k = k, seed = 1)
  sizes_k <- summary(c_k)
  data.frame(k = k, sizes = paste(sizes_k$size, collapse = "/"))
}))
```

`k` is your assumption, not a discovery. Report the eigengap alongside it, or
show that the conclusion survives several values.

#### Seeding

`hypergraph_cluster()` runs k-means on the spectral embedding, which is
stochastic. Pass `seed`:

```{r hypergraph-laplacian-seed}
first <- hypergraph_cluster(hg, k = 3, seed = 1)
second <- hypergraph_cluster(hg, k = 3, seed = 1)
identical(first, second)
```

`nstart` (default 25) controls how many k-means restarts are used; raise it
if clusters move between seeds.

### Seeing the clustering

`plot()` gives two diagnostic panels: the Laplacian spectrum with the
chosen `k` marked, and the spectral embedding the k-means step actually
clustered. The eigengap after `k` is the evidence for (or against) the
choice — read the scree panel before believing the assignment.

```{r hypergraph-laplacian-clplot, fig.width = 12, fig.height = 6, fig.alt = "Laplacian spectrum and spectral embedding, coloured by cluster"}
plot(cl)
```

### Transductive label spreading

Label a few nodes, and let the hypergraph structure propagate labels to the
rest.

```{r hypergraph-laplacian-trans}
labs <- stats::setNames(c("directive", "evaluative"), c("Command", "Verify"))
tr <- hypergraph_transduction(hg, labels = labs, xi = 0.5)
tr
as.data.frame(tr)
```

`score` is the confidence in the predicted class; `margin` is the gap to the
runner-up. Small margins mean the structure did not decide strongly.

#### The default `xi` can override your own seed labels

`xi` trades off *smoothness over the hypergraph* against *fidelity to the
seeds*. It defaults to 0.99, which weights smoothness heavily — heavily
enough that a seed label can be overturned:

```{r hypergraph-laplacian-xi}
do.call(rbind, lapply(c(0.1, 0.5, 0.9, 0.99), function(x) {
  fit <- hypergraph_transduction(hg, labels = labs, xi = x)
  d <- as.data.frame(fit)
  data.frame(xi = x,
             Verify_seeded_as = "evaluative",
             Verify_predicted = d$predicted[d$node == "Verify"],
             n_evaluative = sum(d$predicted == "evaluative"))
}))
```

At `xi = 0.99` — **the default** — Verify is predicted `directive` despite
being seeded `evaluative`, and the evaluative class disappears entirely. At
`xi ≤ 0.9` the seed holds.

This is the documented behaviour of Zhou's algorithm, not a bug: with one
seed per class and a strongly-connected structure, 99% smoothing swamps the
1% fidelity term. But it means **you must check that your seeds survive**.
Lower `xi`, or seed more nodes:

```{r hypergraph-laplacian-moreseeds}
labs2 <- stats::setNames(
  c("directive", "directive", "evaluative", "evaluative"),
  c("Command", "Specify", "Verify", "Correct"))
tr2 <- hypergraph_transduction(hg, labels = labs2, xi = 0.9)
as.data.frame(tr2)
```

With two seeds per class the evaluative class survives at the same `xi`.

### Seeing the spreading

The full node-by-class score matrix, with seed nodes outlined in black and
each node's winning class marked. Rows where the winning and runner-up
scores are close are the assignments to distrust — that closeness is
visible here and invisible in a table of predicted labels.

```{r hypergraph-laplacian-trplot, fig.width = 8, fig.height = 6, fig.alt = "Heatmap of transduction scores by node and class"}
plot(tr2)
```

### Both Laplacians

```{r hypergraph-laplacian-types}
do.call(rbind, lapply(c("zhou", "random_walk"), function(ty) {
  Lt <- hypergraph_laplacian(hg, type = ty)
  e <- sort(eigen(Lt, symmetric = TRUE, only.values = TRUE)$values)
  data.frame(type = ty, lambda_2 = round(e[2], 4),
             lambda_max = round(max(e), 4))
}))
```

Use `random_walk` when the incidence carries genuinely edge-dependent vertex
weights — as `window_hypergraph()` produces, where a cell is an occurrence
count within a window. On a binary incidence the two constructions carry the
same information, and Chitra & Raphael's collapse theorem says the walk is
then equivalent to one on a weighted clique expansion.

### Connectivity is required

```{r hypergraph-laplacian-disconnected, error = TRUE}
iso <- group_hypergraph(
  data.frame(member = c("A", "B", "C", "D"),
             grp    = c("g1", "g1", "g2", "g2"), stringsAsFactors = FALSE),
  actor = "member", group = "grp")
hypergraph_laplacian(iso)
```

A disconnected hypergraph has a degenerate spectrum — the multiplicity of
eigenvalue 0 equals the number of components — so these verbs refuse it with
a classed `honets_hypergraph_disconnected` condition rather than returning
something meaningless. Split the components and analyse them separately.

### What to do next

- **[`hypergraph_centrality()`](hypergraphs.html#hypergraph_centrality)** — the damped
  version of this same random walk.
- **[`hypergraph_measures()`](hypergraphs.html#hypergraph_measures)** — descriptive
  structure before the spectral analysis.
- **[`window_hypergraph()`](hypergraphs.html#window_hypergraph)** — where the
  edge-dependent weights come from.

### References

- Zhou, D., Huang, J., & Schölkopf, B. (2006). Learning with hypergraphs:
  clustering, classification, and embedding. *NeurIPS 19*, 1601-1608.
- Hayashi, K., Aksoy, S. G., Park, C. H., & Park, H. (2020). Hypergraph
  random walks, Laplacians, and clustering. *CIKM 2020*, 495-504.
- Chitra, U., & Raphael, B. J. (2019). Random walks on hypergraphs with
  edge-dependent vertex weights. *ICML 2019*, PMLR 97, 1172-1181.
- von Luxburg, U. (2007). A tutorial on spectral clustering. *Statistics and
  Computing*, 17(4), 395-416.


---

## clique_expansion(): projecting back, and what it costs {#clique_expansion}

### The analytic question

Most network tools take a pairwise network. `clique_expansion()` turns a
hypergraph into one, so the whole first-order toolkit becomes available —
plotting, community detection, standard centralities.

It is also the clearest way to *see* what a hypergraph representation is
worth, because the projection is **lossy and irreversible**, and you can
measure exactly how much it throws away.

```{r clique-expansion-load}
library(honets)
```

```{r clique-expansion-cograph, eval = has_cograph}
library(cograph)
```

### Projecting

Every pair of nodes sharing a hyperedge gets an edge, weighted by how much
they share.

```{r clique-expansion-project}
hg <- group_hypergraph(human_long, actor = "code", group = "session_id")
net <- clique_expansion(hg)
class(net)
summary(net)
w <- hg_project(hg, what = "matrix")
round(w[1:5, 1:5])
```

Nine nodes, 36 edges — the complete graph on nine nodes. Command and
Interrupt share 390 sessions; Frustrate and Inquire share 179.

The return is a `netobject` / `cograph_network`, so it drops straight into
the packages that consume those.

### What `weighted` actually controls

It binarises the **incidence**, not the output. The output is always a count
of shared hyperedges (weighted by incidence values when they exist).

```{r clique-expansion-weighted}
dw <- data.frame(member = c("A", "B", "C", "A", "B"),
                 grp    = c("g1", "g1", "g1", "g2", "g2"),
                 n      = c(2, 5, 3, 1, 4), stringsAsFactors = FALSE)
hw <- group_hypergraph(dw, actor = "member", group = "grp", weight = "n")
as.data.frame(hw, what = "memberships")

hg_project(hw, weighted = TRUE, what = "matrix")
hg_project(hw, weighted = FALSE, what = "matrix")
```

With `TRUE`, A-B is `2×5 + 1×4 = 14` — the incidence values multiply. With
`FALSE`, A-B is `2` — they simply share two hyperedges.

On a hypergraph whose incidence is already 0/1 the argument does nothing:

```{r clique-expansion-noop}
hb <- group_hypergraph(dw, actor = "member", group = "grp")
w_true <- hg_project(hb, weighted = TRUE, what = "matrix")
w_false <- hg_project(hb, weighted = FALSE, what = "matrix")
identical(w_true, w_false)
```

### Seeing the projection

The projected object is a `cograph_network`, so it plots directly — which is
most of the reason to project at all.

```{r clique-expansion-projplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "The projected pairwise network"}
splot(net, minimum = 0.05, title = "Hypergraph, projected")
```

### The information the projection destroys

This is the part worth internalising. Take a hypergraph with two 3-groups,
project it, and rebuild a hypergraph from the projection:

```{r clique-expansion-roundtrip}
d <- data.frame(member = c("A", "B", "C", "A", "B", "D"),
                grp    = c("g1", "g1", "g1", "g2", "g2", "g2"),
                stringsAsFactors = FALSE)
h1 <- group_hypergraph(d, actor = "member", group = "grp")
edges_h1 <- as.data.frame(h1)
edges_h1
```

Two genuine three-way groups. Project:

```{r clique-expansion-proj}
n1 <- clique_expansion(h1)
n1
```

Rebuild:

```{r clique-expansion-rebuild}
h2 <- build_hypergraph(n1, threshold = 0)
edges_h2 <- as.data.frame(h2)
edges_h2
```

**Two hyperedges went in; seven came out.** The two real triples are
recovered — but so are five pairwise edges that were never groups in their
own right, only by-products of the projection. A hypergraph and its
projection are not interchangeable, and the round trip cannot tell a genuine
group from a triangle of coincidences.

That gap is the entire case for
[`build_hypergraph()`](hypergraphs.html#build_hypergraph)'s `p` parameter: when you
*must* infer groups from pairwise data, `p` is where you state how much you
believe the inference.

### Projecting collapses the coding sessions completely

```{r clique-expansion-collapse}
m <- hypergraph_measures(hg)
as.data.frame(m, what = "global")
```

Pairwise participation is **1.0**: every pair of codes co-occurs somewhere.
So the projection is the complete graph, and any first-order centrality on it
is near-degenerate — the same saturation that makes the first-order
transition network uninformative in the
[three-families comparison](overview.html#three_families).

```{r clique-expansion-degenerate}
deg <- rowSums(w > 0)
range(deg)
```

Every node has degree 8. The hypergraph, meanwhile, still distinguishes the
codes clearly by size and overlap structure — see
[`hypergraph_centrality()`](hypergraphs.html#hypergraph_centrality).

### When to use it anyway

- To hand a hypergraph to a tool that only accepts pairwise networks.
- To compare a higher-order result against its first-order counterpart, and
  show that the higher-order structure is doing work.
- When Chitra & Raphael's collapse theorem applies: with **edge-independent**
  vertex weights, a random walk on the hypergraph is *equivalent* to one on
  the weighted clique expansion. In that case the projection is lossless for
  walk-based purposes, and the hypergraph adds nothing — which is itself
  worth knowing, and is why
  [`window_hypergraph()`](hypergraphs.html#window_hypergraph)'s edge-dependent weights
  matter.

### What to do next

- **[`build_hypergraph()`](hypergraphs.html#build_hypergraph)** — the inverse direction,
  with an explicit belief parameter.
- **[`hypergraph_centrality()`](hypergraphs.html#hypergraph_centrality)** — centrality
  without projecting.
- **[`hypergraph_measures()`](hypergraphs.html#hypergraph_measures)** — the overlap
  structure that projection flattens.

### References

- Chitra, U., & Raphael, B. J. (2019). Random walks on hypergraphs with
  edge-dependent vertex weights. *ICML 2019*, PMLR 97, 1172-1181.
- Agarwal, S., Branson, K., & Belongie, S. (2006). Higher order learning with
  graphs. *ICML 2006*, 17-24.
- Tian, Y., & Zafarani, R. (2024). Higher-order network analysis methods.
  *ACM Computing Surveys*.
