---
title: "honets vignettes"
subtitle: "Higher-order network analysis in R"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      comment = "")
```

A network is *higher-order* when a relation reaches past a single pair of
nodes at a single moment. There are two ways to reach past it, and three
structures that do so. honets implements all three under one taxonomy.

The tables below route the main analysis verbs to worked documents. The
package reference indexes every export; compact `hg_*()` aliases share the
same documentation as their long `hypergraph_*()` forms.

## Start here

**[Three higher-order readings of one dataset](overview.html#three_families)** — the
same 429 coded sessions read as a memory network, a simplicial complex and a
hypergraph. The three disagree, and where they disagree is the finding. Read
this first if you are deciding which family your question belongs to.

## Memory networks

*A node is a state plus the memory of how it was reached. Built from
categorical sequences.*

| Vignette | Answers |
|---|---|
| [`build_hon()`](memory-networks.html#build_hon) | What follows what, given how you got here? |
| [`markov_order_test()`](memory-networks.html#markov_order_test) | Is there memory at all, and how far back? |
| [`path_dependence()`](memory-networks.html#path_dependence) | *Which* contexts carry the memory? |
| [`build_mogen()`](memory-networks.html#build_mogen) | Which single order do the data support? |
| [`build_hypa()`](memory-networks.html#build_hypa) | Which paths occur more or less often than chance? |
| [`hon_centrality()`](memory-networks.html#hon_centrality) | Which states carry the flow? |
| [`bootstrap_hon()` and `compare_hon()`](memory-networks.html#hon_inference) | Do the rules survive resampling, and do two cohorts differ? |
| [`build_honem()`](memory-networks.html#build_honem) | How do I get vector features out of it? |
| [`pathways()`](overview.html#pathways) | How do I hand this to the other two families? |

## Simplicial complexes

*A relation is a set of mutually related nodes plus every subset — which
gives it a geometry, and therefore a topology.*

| Vignette | Answers |
|---|---|
| [`build_simplicial()`](simplicial-complexes.html#build_simplicial) | How do I turn a network, a distance matrix or a memory network into a shape? |
| [`betti_numbers()`](simplicial-complexes.html#betti_numbers) | Where are the holes — the structures that are conspicuously *absent*? |
| [`persistent_homology()`](simplicial-complexes.html#persistent_homology) | Which of those features are real rather than artefacts of a threshold? |
| [`persistence_landscape()` and `bottleneck_distance()`](simplicial-complexes.html#persistence_landscape) | How do I compare two of these, or average a sample of them? |
| [`q_analysis()` and `simplicial_degree()`](simplicial-complexes.html#q_analysis) | How strongly do the groups overlap, and which nodes sit deepest? |

## Hypergraphs

*A relation is an arbitrary set of nodes bound as a unit, with no requirement
that its subsets are relations too.*

| Vignette | Answers |
|---|---|
| [`group_hypergraph()`](hypergraphs.html#group_hypergraph) | My data already record group membership — how do I keep it? |
| [`window_hypergraph()`](hypergraphs.html#window_hypergraph) | My data are sequences — how do I get local group structure? |
| [`build_hypergraph()`](hypergraphs.html#build_hypergraph) | I only have pairwise data — how do I infer groups, and state my assumption? |
| [`hypergraph_measures()`](hypergraphs.html#hypergraph_measures) | How is this hypergraph structured? |
| [`hypergraph_centrality()`](hypergraphs.html#hypergraph_centrality) | Which nodes are central without projecting? |
| [`hypergraph_laplacian()`, `hypergraph_cluster()`, `hypergraph_transduction()`](hypergraphs.html#hypergraph_laplacian) | Spectral clustering and label spreading on a hypergraph. |
| [`clique_expansion()`](hypergraphs.html#clique_expansion) | How do I project back to a pairwise network, and what does it cost? |
| [`hypergraph_project()` and `hypergraph_line_graph()`](legal-hypergraphs.html) | How do I create association and s-line projections? |
| [`temporal_hypergraph()` and snapshot verbs](legal-hypergraphs.html) | How do higher-order relations enter, persist and leave over time? |
| [`hypergraph_edges()` and `hypergraph_edge_centrality()`](legal-hypergraphs.html) | Which hyperedges are large, connected or structurally central? |
| [`hypergraph_motifs()`](legal-hypergraphs.html) | Are Y/T/O patterns enriched against a configuration null? |
| [`hypergraph_communities()` and `hypergraph_community_quality()`](legal-hypergraphs.html) | Which association-graph partition is stable, and how good is it? |
| [`hg_neural()` and `hg_hypergat()`](../articles/benchmarks.html) | How do the native neural models compare with closed-form methods? |

## Text hypergraphs

| Vignette | Answers |
|---|---|
| [`text_hypergraph()` and tidy `hg_*()` verbs](../articles/text-hypergraphs.html) | How do I construct and analyse a corpus hypergraph? |
| [Window and kNN constructions](../articles/text-constructions.html) | How do token order and embedding neighbourhoods change the representation? |
| [Benchmark article](../articles/benchmarks.html) | How do transduction, HGNN and HyperGAT perform on the paper datasets? |

## The verb grammar

The same naming rules hold in every family, so a verb from one reads like a
verb from another.

- **`build_*()`** constructors always return a `net_*` object. Where a family
  admits several construction routes, they are selected with `type =`.
- **`*_centrality()`, `*_measures()`, `*_degree()`** always return a tidy
  `data.frame`, one row per node or per structure.
- **`bootstrap_*()`, `compare_*()`, `*_test()`** return a `net_*` result
  carrying estimates, intervals and p-values.
- **Every `net_*` result supports `print()`, `summary()`, `plot()` and
  `as.data.frame()`.** The accessor is the supported way in — reaching into a
  result with `$` is never required. A secondary table is selected with
  `what =`, a subset with a named argument, an ordering with `sort_by =`.

```{r grammar, echo = TRUE, eval = FALSE}
as.data.frame(hon)                      # the extracted rules
as.data.frame(hon, what = "nodes")      # the higher-order nodes
as.data.frame(ph,  what = "betti")      # Betti curves, not the diagram
as.data.frame(hm,  what = "global")     # whole-hypergraph summary
```

## Recurring cautions

Four warnings appear across these vignettes often enough to collect here.

**Thresholds are choices, and they move the answer.** `min_freq` in
`build_hon()`, `threshold` in `build_simplicial()`, `min_count` in
`path_dependence()`. Show the sweep or report the value — a Betti number
without its threshold is not a result.

**Stochastic verbs need a seed, and one seed is not a finding.**
`build_hypergraph(p < 1)`, `hypergraph_cluster()`, every permutation test.
Pass `seed`, then check the conclusion survives across several.

**Information criteria disagree, and the disagreement is informative.** AIC
and BIC select different orders in both `build_mogen()` and
`markov_order_test()` on the bundled data. Report which you used and why.

**Defaults are not always safe.** `build_simplicial(type = "vr")` defaults to
`max_dim = 10L`, which on a 20-point cloud produces 784,625 simplices.
`hypergraph_transduction()` defaults to `xi = 0.99`, which can override your
own seed labels. Both are documented in their vignettes.

## Rebuilding these

```sh
Rscript -e 'devtools::install(".")'
Rscript docs/_knit_all.R              # all of them
Rscript docs/_knit_all.R build_hon    # one, by basename
```

Each vignette runs live against the installed package, so a knit that
succeeds is also a check that the documented behaviour is the actual
behaviour.
