---
title: "Legal hypergraphs on toy data: the whole workflow, step by step"
author: "honets"
date: "2026-09-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
library(honets)
# igraph is a suggested package; the community section needs it.
has_igraph <- requireNamespace("igraph", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE,
                      fig.width = 7, fig.height = 5, out.width = "100%",
                      dpi = 150)
```

> This is the legal-hypergraphs workflow (Coupette, Hartung and Katz
> 2024) on data small enough to read by eye. Every step is one call. The
> real analysis, on the authors' released data, is `legal-hypergraphs.html`;
> the verbs are the same, only the data change.

## The two kinds of data

The paper uses two datasets. **Tribunals**: a case is decided by three
arbitrators, from the day the tribunal is constituted to the day the case
concludes. The three arbitrators are bound together for that interval.
**Citations**: a decision cites earlier decisions in blocks. The decisions
cited together in one block are bound together on the day of the citing
decision.

We make both up with base R. A seed makes the data the same every time.

```{r seed}
set.seed(2024)
```

### Fake tribunals

Twenty arbitrators and forty cases. Every case has three seats. A few
arbitrators are much busier than the rest, as in the real data, so some
pairs and even trios sit together more than once.

```{r arbitrators}
arbitrators <- sprintf("arbitrator_%02d", 1:20)
```

```{r cases}
cases <- sprintf("case_%02d", 1:40)
```

```{r seats}
seats <- unlist(lapply(cases, function(case) {
  sample(arbitrators, size = 3, prob = 1 / seq_along(arbitrators))
}))
```

Each case is constituted on a random day between 2010 and 2019 and
concludes one to four years later. Eight cases are still pending, so their
conclusion date is missing.

```{r dates-tribunals}
constituted <- as.Date("2010-01-01") + sort(sample(0:3650, size = 40))
concluded <- constituted + sample(365:1460, size = 40)
concluded[33:40] <- NA
```

Each case belongs to an economic sector, a property of the case, not of a
seat.

```{r sector}
sector <- sample(c("energy", "mining", "finance", "construction"),
                 size = 40, replace = TRUE)
```

One row per seat: the case, the arbitrator, the two dates, the sector and
whether the case has concluded. This is the shape of `icsid_tribunals`.

```{r tribunal-table}
toy_tribunals <- data.frame(
  case = rep(cases, each = 3),
  arbitrator = seats,
  constituted = rep(constituted, each = 3),
  concluded = rep(concluded, each = 3),
  sector = rep(sector, each = 3),
  is_concluded = rep(!is.na(concluded), each = 3),
  stringsAsFactors = FALSE
)
head(toy_tribunals, 6)
```

### Fake citations

Sixty decisions, one per quarter from 2005. Each decision after the fifth
cites between four and eight earlier decisions, spread over one or two blocks.
Later decisions are cited more often, as in a growing court.

```{r decisions}
decisions <- sprintf("decision_%02d", 1:60)
decided <- as.Date("2005-01-01") + seq(0, by = 91, length.out = 60)
toy_decisions <- data.frame(decision = decisions, date = decided,
                            stringsAsFactors = FALSE)
head(toy_decisions, 3)
```

```{r citation-counts}
n_cites <- c(rep(0, 5), sample(4:8, size = 55, replace = TRUE))
citing_index <- rep(seq_along(decisions), n_cites)
```

```{r cited}
cited_index <- vapply(citing_index, function(i) {
  sample(seq_len(i - 1), size = 1, prob = seq_len(i - 1))
}, integer(1))
```

```{r blocks}
block_number <- sample(1:2, size = length(citing_index), replace = TRUE)
```

One row per citation: the citing decision, the cited decision, the block
the citation sits in, and both dates. A decision cited twice in the same
block is one citation. This is the shape of `gfcc_citations`.

```{r citation-table}
toy_citations <- unique(data.frame(
  citing = decisions[citing_index],
  cited = decisions[cited_index],
  block = sprintf("%s_block_%d", decisions[citing_index], block_number),
  date_citing = decided[citing_index],
  date_cited = decided[cited_index],
  stringsAsFactors = FALSE
))
head(toy_citations, 6)
```

## Models

`temporal_hypergraph()` reads co-presence data: `actor` is the node, `group`
the column whose shared value binds actors into one hyperedge. `start` and
`end` give the interval. A missing end stays active through the end of
observation, which we declare as the end of 2020.

```{r observation-end}
observation_end <- as.Date("2020-12-31")
```

```{r tribunals-build}
tribunals <- temporal_hypergraph(
  toy_tribunals, actor = "arbitrator", group = "case",
  start = "constituted", end = "concluded",
  observation_end = observation_end
)
tribunals
```

`summary()` is the paper's Table 1 for this dataset.

```{r tribunals-summary}
summary(tribunals)
```

The citations are the same shape with `time` instead of an interval: a block
is a contact on the citing date. The decision table is the node universe,
so a decision exists from its own date even before anyone cites it.

```{r blocks-build}
blocks <- temporal_hypergraph(
  toy_citations, actor = "cited", group = "block", time = "date_citing",
  nodes = toy_decisions
)
blocks
```

```{r blocks-summary}
summary(blocks)
```

### Four representations of the same data

`hypergraph_snapshot()` with `mode = "cumulative"` and no `at` is the static
aggregate: every hyperedge that ever appeared, as an ordinary hypergraph.

```{r aggregate-tribunals}
tribunals_all <- hypergraph_snapshot(tribunals, mode = "cumulative")
tribunals_all
```

```{r aggregate-blocks}
blocks_all <- hypergraph_snapshot(blocks, mode = "cumulative")
blocks_all
```

`hg_representations()` reads an aggregate the four ways of the paper's Table
2: the multi-hypergraph (`mh`), the binary hypergraph of distinct member sets
(`bh`), and the graph it reduces to with (`mg`) or without (`bg`) edge
multiplicities.

```{r tribunals-representations}
hg_representations(tribunals_all)
```

For citations the graph is the citation graph: `graph = "citation"` with
`edge_source` naming the attribute that holds the citing decision.

```{r blocks-representations}
hg_representations(blocks_all, graph = "citation", edge_source = "citing")
```

### One decision as a hypergraph

`hg_subset()` with `where =` keeps the hyperedges whose attribute takes a
value: the two blocks of decision 56, four cited decisions each.

```{r decision}
one_decision <- hg_subset(blocks_all, where = c(citing = "decision_56"))
one_decision
```

`plot()` draws every block as a blob around the decisions it cites,
coloured by block size.

```{r decision-plot, fig.width = 6, fig.height = 6}
plot(one_decision, color_by = "size", padding = 0.05,
     legend_title = "block size")
```

## Growth

`hg_growth()` counts, at every event time, the nodes and hyperedges present.
With `mode = "cumulative"` the counts accumulate and a decision enters at its
own date.

```{r blocks-growth}
growth_blocks <- hg_growth(blocks, mode = "cumulative")
tail(growth_blocks, 3)
```

```{r blocks-growth-plot, fig.height = 5}
plot(growth_blocks,
     columns = c("n_nodes", "n_edges", "n_edges_distinct", "n_memberships"))
```

`hg_edges()` with `what = "distribution"` gives the distribution of block
sizes as the network stood on chosen dates.

```{r census-dates}
census_dates <- as.Date(c("2008-12-31", "2012-12-31", "2016-12-31",
                          "2019-12-31"))
```

```{r blocks-sizes}
sizes_by_date <- hg_edges(blocks, what = "distribution", at = census_dates,
                          snapshot_mode = "cumulative")
head(sizes_by_date)
```

```{r blocks-sizes-plot}
plot(sizes_by_date)
```

The `s` argument counts, for each block, the other blocks sharing at least
`s` decisions with it.

```{r blocks-intersections}
intersections <- hg_edges(blocks_all, s = 1:3, what = "distribution",
                          measure = "n_incident_edges")
head(intersections)
```

```{r blocks-intersections-plot, fig.height = 4}
plot(intersections)
```

## Arbitrators and tribunals over time

`hg_measures()` with `what = "distribution"` gives the degree distribution.
The `hyperdegree` of an arbitrator is the number of tribunals sat on.

```{r tribunals-degrees}
tribunal_counts <- hg_measures(tribunals_all, what = "distribution",
                               measure = "hyperdegree")
tail(tribunal_counts, 3)
```

```{r tribunals-degrees-plot, fig.height = 4}
plot(tribunal_counts)
```

For interval data `hg_growth()` reports the active counts and, in the
`_cumulative` columns, the counts of a model that never retires a tribunal.

```{r tribunals-growth}
growth_tribunals <- hg_growth(tribunals)
tail(growth_tribunals, 3)
```

```{r tribunals-growth-plot, fig.height = 4}
plot(growth_tribunals,
     columns = c("n_nodes", "n_edges", "n_nodes_cumulative", "n_edges_cumulative"),
     facets = FALSE)
```

`hg_edges()` with `what = "summary"` gives, per snapshot, the mean and
quartiles of a hyperedge measure; `n_neighbors` is the number of
arbitrators within one hop of a tribunal.

```{r tribunals-neighbourhood}
neighbourhood <- hg_edges(tribunals, what = "summary", measure = "n_neighbors")
tail(neighbourhood, 3)
```

```{r tribunals-neighbourhood-plot, fig.height = 4}
plot(neighbourhood, columns = c("mean", "q25", "median", "q75"), facets = FALSE)
```

`components = TRUE` adds the connectivity of the active hypergraph at each
time.

```{r tribunals-components}
connectivity <- hg_growth(tribunals, components = TRUE)
tail(connectivity, 3)
```

## Repeated collaborations

`hg_null_test()` compares a statistic with its distribution under a null that
keeps every arbitrator's number of seats and every tribunal's size.
`repeated_edges` counts tribunals whose trio already occurred,
`repeated_pairs` the pairs that sat together more than once. Our seats were
drawn at random, so the repeats are only what busy arbitrators produce by
chance, and the test says so: the observed counts sit inside the null
range. On the real data they lie far above it.

```{r tribunals-repeats}
hg_null_test(tribunals_all, statistic = c("repeated_pairs", "repeated_edges"),
             method = "assignment", n = 199, seed = 1)
```

## Hyperedge centralities

The centrality analysis runs on the tribunals active on the last observed
date. `multiedges = FALSE` collapses two tribunals with the same three
members into one hyperedge.

```{r active}
active <- hypergraph_snapshot(tribunals, at = observation_end,
                              mode = "active", multiedges = FALSE)
active
```

`hg_edge_centrality()` builds the s-line graph, in which two tribunals are
adjacent when they share at least `s` arbitrators, and ranks the tribunals
by betweenness on it. `top = 5` keeps the five most central.

```{r betweenness}
top_betweenness <- hg_edge_centrality(active, s = 1, measure = "betweenness",
                                      top = 5)
top_betweenness
```

`hg_subset()` takes that table and cuts the snapshot down to those cases.

```{r betweenness-subset}
central_tribunals <- hg_subset(active, edges = top_betweenness)
central_tribunals
```

`plot()` draws them coloured by sector.

```{r betweenness-plot, fig.width = 7, fig.height = 6}
plot(central_tribunals, color_by = "sector", legend_title = "sector",
     label_size = 3)
```

Closeness ranks a different set.

```{r closeness}
top_closeness <- hg_edge_centrality(active, s = 1, measure = "closeness",
                                    top = 5)
top_closeness
```

```{r closeness-subset}
close_tribunals <- hg_subset(active, edges = top_closeness)
```

```{r closeness-plot, fig.width = 7, fig.height = 6}
plot(close_tribunals, color_by = "sector", legend_title = "sector",
     label_size = 3)
```

The static aggregate ranks tribunals that never overlapped in time. As for
the active snapshot, duplicate trios are collapsed first.

```{r aggregate-binary}
tribunals_binary <- hypergraph_snapshot(tribunals, mode = "cumulative",
                                        multiedges = FALSE)
tribunals_binary
```

```{r aggregate-closeness}
top_aggregate <- hg_edge_centrality(tribunals_binary, s = 1,
                                    measure = "closeness", top = 5)
top_aggregate
```

```{r aggregate-subset}
aggregate_tribunals <- hg_subset(tribunals_binary, edges = top_aggregate)
```

`linetype_by` outlines the concluded cases differently from the pending
ones.

```{r aggregate-plot, fig.width = 7, fig.height = 6}
plot(aggregate_tribunals, color_by = "sector", linetype_by = "is_concluded",
     legend_title = "sector", label_size = 3)
```

## Motifs

`hg_motifs()` counts the three four-node motifs of a 3-uniform hypergraph,
Y (two of the four possible triples), T (three) and O (all four), against
configuration-model nulls that keep degrees and sizes. Random seats give
counts close to their null means; the real tribunals show a clear excess.

```{r motifs}
motifs_aggregate <- hg_motifs(tribunals_all, n = 200, seed = 1)
motifs_aggregate
```

```{r motifs-plot, fig.height = 4}
plot(motifs_aggregate)
```

The active snapshot is smaller and has its own census.

```{r motifs-active}
motifs_active <- hg_motifs(active, n = 200, seed = 1)
motifs_active
```

```{r motifs-active-plot, fig.height = 4}
plot(motifs_active)
```

## Communities across representations

The paper clusters the citation data eight ways: the association graph of
`hg_project()`, in which a block of size |e| gives each pair of its members
the weight 1 / (|e| - 1), on the binary (`bh`) or multi (`mh`) hypergraph,
with (`bhs`, `mhs`) or without self-association, the term that joins the
citing decision to the decisions it cites; and the citation graph, binary
(`bg`) or multi (`mg`), directed or undirected (`bgu`, `mgu`). Our fake
decisions only ever cite earlier ones, so every citation is backward already.

`hg_communities()` runs Infomap repeatedly on the chosen projection and keeps
the run most similar to all the others as the medoid. The four classic
citation graphs first.

```{r communities-graphs, eval = has_igraph}
communities_bg <- hg_communities(blocks_all, n_runs = 5, trials = 5,
                                 method = "citation", edge_source = "citing",
                                 duplicate_edges = "collapse", directed = TRUE)
communities_bgu <- hg_communities(blocks_all, n_runs = 5, trials = 5,
                                  method = "citation", edge_source = "citing",
                                  duplicate_edges = "collapse")
communities_mg <- hg_communities(blocks_all, n_runs = 5, trials = 5,
                                 method = "citation", edge_source = "citing",
                                 directed = TRUE)
communities_mgu <- hg_communities(blocks_all, n_runs = 5, trials = 5,
                                  method = "citation", edge_source = "citing")
communities_bg
```

Then the four association graphs.

```{r communities-hypergraphs, eval = has_igraph}
communities_bh <- hg_communities(blocks_all, n_runs = 5, trials = 5,
                                 duplicate_edges = "collapse")
communities_bhs <- hg_communities(blocks_all, n_runs = 5, trials = 5,
                                  duplicate_edges = "collapse",
                                  self_association = TRUE, edge_source = "citing")
communities_mh <- hg_communities(blocks_all, n_runs = 5, trials = 5)
communities_mhs <- hg_communities(blocks_all, n_runs = 5, trials = 5,
                                  self_association = TRUE, edge_source = "citing")
communities_mhs
```

`hg_compare_communities()` sets the eight fits side by side and scores each
medoid on its own projection. Sixty random decisions have no community
structure to find, so the citation graphs come back as one community and
the association graphs split off a few singletons; the table reports that
faithfully. The real data separate into dozens of communities.

```{r compare, eval = has_igraph}
comparison <- hg_compare_communities(
  bg = communities_bg, bgu = communities_bgu, mg = communities_mg,
  mgu = communities_mgu, bh = communities_bh, bhs = communities_bhs,
  mh = communities_mh, mhs = communities_mhs, hg = blocks_all,
  edge_source = "citing"
)
comparison
```

```{r compare-plot, eval = has_igraph, fig.height = 4}
plot(comparison)
```

The similarity table holds the AMI, ARI and NMI between every pair of
medoids.

```{r similarity-table, eval = has_igraph}
similarity <- as.data.frame(comparison, what = "similarity")
head(similarity)
```

```{r similarity-plot, eval = has_igraph, fig.width = 6, fig.height = 5}
plot(comparison, what = "similarity")
```

The quality table holds coverage, performance, modularity and the worst
community conductance of each medoid on its own projection.

```{r quality, eval = has_igraph}
quality <- as.data.frame(comparison, what = "quality")
quality
```

## What you have seen

Two fake tables, one row per seat and one row per citation, went through
the whole workflow of the paper: Table 1 by `summary()`, Table 2 by
`hg_representations()`, a decision's hypergraph by `hg_subset()` and
`plot()`, growth and size distributions by `hg_growth()`, `hg_edges()` and
`hg_measures()`, repeated collaborations by `hg_null_test()`, hyperedge
centralities by `hg_edge_centrality()`, motifs by `hg_motifs()`, and
communities by `hg_communities()` and `hg_compare_communities()`. Replace
the two tables with your own and nothing else changes. What changes is the
answers: made-up data give null results everywhere, which is exactly what
the tests are for.

## References

Coupette, C., Hartung, D., and Katz, D. M. (2024). Legal hypergraphs.
Philosophical Transactions of the Royal Society A, 382(2270), 20230141.
