---
title: "Legal hypergraphs: the paper workflow in honets"
author: "honets"
date: "2026-09-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
library(honets)
# igraph is a suggested package; the community section needs it.
has_igraph <- requireNamespace("igraph", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE,
                      fig.width = 7, fig.height = 5, out.width = "100%",
                      dpi = 200)
```

> The sections run in order in a single R session; later ones reuse the
> objects the earlier ones built. One step per chunk: each chunk saves one
> object or draws one figure.

Coupette, Hartung and Katz (2024) model legal relations as temporal
hypergraphs: a citation block binds every decision it cites to the decision
that cites them, and a tribunal binds its three arbitrators for as long as
the case runs. The paper works through two datasets, the citation blocks of
the German Federal Constitutional Court (GFCC) and the tribunals of the
International Centre for Settlement of Investment Disputes (ICSID), and
applies three families of methods to them: hyperedge centralities, motifs
against a configuration null, and communities of a weighted projection. This
vignette reproduces that workflow in honets on the authors' released data,
figure by figure. Hypergraph construction and mathematics are native to
honets; once a projection is an ordinary graph, cograph supplies the graph
algorithms and the drawing.

## Data

Both datasets ship with honets, rebuilt from the authors' Zenodo archive
(record 8081507, CC BY-NC 4.0) by `data-raw/legal_hypergraphs.R`.
`icsid_tribunals` has one row per tribunal seat: the case, the arbitrator
who held the seat, the dates on which the tribunal was constituted and the
case concluded, and the case metadata. A pending case has no conclusion
date.

```{r data-icsid}
head(icsid_tribunals)
```

`gfcc_decisions` has one row per decision with its date, and
`gfcc_citations` one row per citation: the citing decision, the cited
decision, the citation block the citation belongs to, and the dates of both
decisions.

```{r data-gfcc}
head(gfcc_decisions)
head(gfcc_citations)
```

## Models

`temporal_hypergraph()` reads co-presence data the way Dynet does: `actor`
names the node column and `group` the column whose shared value binds
actors into one hyperedge, here the case. `start` and `end` give the
interval on which the hyperedge is active, the paper's interval-event
model. The dates become days since the first constitution, the unit and
origin the object reports. A case without a conclusion date stays active
through the end of observation, which the paper sets at its cut-off,
2023-06-15; `observation_end` declares that window without rewriting the
data. Every other column that is constant within a case, such as the
economic sector, is kept as an attribute of the hyperedge; the seat, which
varies within a case, is not.

The cut-off is used twice below, so it is one object.

```{r observation-end}
observation_end <- as.Date("2023-06-15")
```

```{r icsid-build}
tribunals <- temporal_hypergraph(
  icsid_tribunals, actor = "arbitrator", group = "case",
  start = "constituted", end = "concluded",
  observation_end = observation_end
)
tribunals
```

`summary()` is the paper's Table 1 for this dataset.

```{r icsid-summary}
summary(tribunals)
```

The citation table is the same shape: the cited decision is the actor and
the citation block the group. A block has a single `time`, the date of the
decision that contains it, so it is a contact hyperedge: an instantaneous
event, as in a contact log. The paper's point-aggregation model, in which
a block stays in the network once it has appeared, is not a property of
the data but a way of looking at it, and it is asked for below with
`mode = "cumulative"`. The citing decision is constant within a block, so
`citing` becomes the block's attribute. The decision table is the node
universe, so that a decision exists from its own date whether or not it
has been cited. `sparse = TRUE` stores the 46,257 by 3,618 incidence as a
sparse matrix.

```{r gfcc-build}
blocks <- temporal_hypergraph(
  gfcc_citations, actor = "cited", group = "block", time = "date_citing",
  nodes = gfcc_decisions, sparse = TRUE
)
blocks
```

```{r gfcc-summary}
summary(blocks)
```

The two summaries are the paper's Table 1: 742 cases with 441 arbitrators
and 2,226 memberships, and 46,257 citation blocks with 77,284 memberships
over 2,026 decision dates between 1951 and 2022. The paper counts 1,077
event times for ICSID because it adds its cut-off date as the end of every
pending case; here those cases have no end.

### Four representations of the same data

`hypergraph_snapshot()` with `mode = "cumulative"` and no `at` returns the
static aggregate of a temporal hypergraph, every hyperedge begun by the end
of observation, as a `net_hypergraph` that every static verb accepts.

```{r aggregate-icsid}
tribunals_all <- hypergraph_snapshot(tribunals, mode = "cumulative")
tribunals_all
```

The same call on the citation blocks gives the aggregate citation
hypergraph.

```{r aggregate-gfcc}
blocks_all <- hypergraph_snapshot(blocks, mode = "cumulative")
blocks_all
```

`hg_representations()` reports an aggregate the four ways the paper's
Table 2 reads it: the multi-hypergraph as it is (`mh`), the binary
hypergraph of distinct member sets (`bh`), and the graph the hypergraph
reduces to, with (`mg`) or without (`bg`) edge multiplicities. For the
tribunals the graph is the clique expansion, in which every pair of members
of a tribunal is an edge.

```{r icsid-representations}
hg_representations(tribunals_all)
```

For the citation blocks the graph is the citation graph, in which the citing
decision is joined to every decision in the block; `graph = "citation"`
selects it and `edge_source` names the attribute that holds the citing
decision. The multi-graph has five edges fewer than the paper's 77,284
because five citations repeat a target inside one block, which a graph edge
list can hold and a hypergraph incidence cannot, and the binary hypergraph
has two member sets fewer for the same reason.

```{r gfcc-representations}
hg_representations(blocks_all, graph = "citation", edge_source = "citing")
```

The representation changes what a node looks like: a decision has a median
of 6 outgoing citations as a graph node and is a member of a median of 10
blocks as a hypergraph node, and an arbitrator has a median of 4 colleagues
but sits on a median of 2 tribunals.

### Decisions as hypergraphs

`hg_subset()` with `where =` keeps the hyperedges whose attribute takes a
value, here the blocks of one citing decision: the hypergraph that a single
decision is in this model. The decision
153-001 is the headscarf ban; its 68 citation blocks bind 85 cited
decisions, 28 of them cited alone.

```{r decision}
headscarf <- hg_subset(blocks_all, where = c(citing = "153-001"))
headscarf
```

`plot()` places the cited decisions with a spring layout of the clique
projection and draws every block as a blob around its members, coloured by
cardinality: the paper's Figure 3. A decision cited alone is drawn as its
node.

```{r decision-plot, fig.width = 7, fig.height = 7}
plot(headscarf, color_by = "size", labels = FALSE, padding = 0.035,
     legend_title = "block size")
```

The decision 125-260 on data retention has the same structure with larger
blocks.

```{r decision-2}
retention <- hg_subset(blocks_all, where = c(citing = "125-260"))
retention
```

```{r decision-2-plot, fig.width = 7, fig.height = 7}
plot(retention, color_by = "size", labels = FALSE, padding = 0.035,
     legend_title = "block size")
```

### Growth

`hg_growth()` counts, at every event time, the nodes and hyperedges present:
the hyperedges as they are, the distinct member sets, and the memberships.
With `mode = "cumulative"` every count is cumulative and a decision enters
at its own date, the paper's growing citation network. The result is a
time series table whose `time` column is on the hypergraph's clock, days
since the first decision; the plot method draws it on the calendar.

```{r gfcc-growth}
growth_blocks <- hg_growth(blocks, mode = "cumulative")
tail(growth_blocks, 3)
```

Its `plot()` method draws each count against time, the paper's Figure 4a.

```{r gfcc-growth-plot, fig.height = 5}
plot(growth_blocks,
     columns = c("n_nodes", "n_edges", "n_edges_distinct", "n_memberships"))
```

`hg_edges()` on a temporal hypergraph evaluates every snapshot requested with
`at` (dates are read on the hypergraph's clock), and `what = "distribution"`
returns the empirical distribution of a hyperedge measure with its
complementary cumulative share, one block of rows per date. The cumulative
mode again gives the network as it stood on each date.

```{r census-dates}
census_dates <- as.Date(c("1960-12-31", "1975-12-31", "1990-12-31",
                          "2005-12-31", "2020-12-31"))
```

```{r gfcc-sizes}
sizes_by_date <- hg_edges(blocks, what = "distribution", at = census_dates,
                          snapshot_mode = "cumulative")
head(sizes_by_date)
```

Its `plot()` draws one curve per date, the paper's Figure 4b; the
distribution shifts to the right as blocks grow longer.

```{r gfcc-sizes-plot}
plot(sizes_by_date)
```

The `s` argument counts, for each hyperedge, the other hyperedges that share
at least `s` nodes with it. The paper's Figure 4c draws that count for
`s = 2, 3, 4` on the multi-hypergraph.

```{r gfcc-intersections}
intersections <- hg_edges(blocks_all, s = 2:4, what = "distribution",
                          measure = "n_incident_edges")
head(intersections)
```

```{r gfcc-intersections-plot, fig.height = 4}
plot(intersections)
```

`multiedges = FALSE` in the snapshot collapses repeated member sets, the
binary hypergraph, on which the same counts are smaller.

```{r gfcc-binary}
blocks_binary <- hypergraph_snapshot(blocks, mode = "cumulative",
                                     multiedges = FALSE)
blocks_binary
```

```{r gfcc-intersections-binary}
intersections_binary <- hg_edges(blocks_binary, s = 2:4,
                                 what = "distribution",
                                 measure = "n_incident_edges")
```

```{r gfcc-intersections-binary-plot, fig.height = 4}
plot(intersections_binary)
```

### Arbitrators and tribunals over time

`hg_measures()` with `what = "distribution"` gives the degree distributions
of the paper's Figure 5a. The number of tribunals an arbitrator sits on is
the `hyperdegree`; one arbitrator, Brigitte Stern, sits on 94 tribunals and
accounts for the tail.

```{r icsid-degrees}
tribunal_counts <- hg_measures(tribunals_all, what = "distribution",
                               measure = "hyperdegree")
tail(tribunal_counts, 3)
```

```{r icsid-degrees-plot, fig.height = 4}
plot(tribunal_counts)
```

For interval data `hg_growth()` reports the active counts and, in the
`_cumulative` columns, the counts of a static model that never retires a
tribunal.

```{r icsid-growth}
growth_tribunals <- hg_growth(tribunals)
tail(growth_tribunals, 3)
```

The paper's Figure 5b: cases grow about twice as fast as active cases, and
arbitrators about three times as fast as active arbitrators.

```{r icsid-growth-plot, fig.height = 4}
plot(growth_tribunals,
     columns = c("n_nodes", "n_edges", "n_nodes_cumulative", "n_edges_cumulative"),
     facets = FALSE)
```

`hg_edges()` with `what = "summary"` returns, per snapshot, the mean and
quartiles of a hyperedge measure. The number of arbitrators within one hop
of a tribunal is `n_neighbors`.

```{r icsid-neighbourhood}
neighbourhood <- hg_edges(tribunals, what = "summary", measure = "n_neighbors")
tail(neighbourhood, 3)
```

The paper's Figure 5c: the mean and median stay close together, unlike the
node degrees above.

```{r icsid-neighbourhood-plot, fig.height = 4}
plot(neighbourhood, columns = c("mean", "q25", "median", "q75"), facets = FALSE)
```

`components = TRUE` adds the connectivity of the active hypergraph at each
time: the number of components, the share of active arbitrators in the
largest, and its diameter. Since 2007 the largest component holds more than
90 percent of the active arbitrators and its diameter stays between 6 and 9,
the observation the paper makes in prose.

```{r icsid-components}
connectivity <- hg_growth(tribunals, components = TRUE)
tail(connectivity, 3)
```

### Repeated collaborations

`hg_null_test()` compares a statistic with its distribution under a null that
preserves every node's degree and every hyperedge's size. The
`repeated_edges` statistic counts tribunals whose member set already
occurred, the paper's repeated collaboration trios, and `repeated_pairs` the
pairs of arbitrators counted with multiplicity minus the distinct pairs, its
repeated collaboration pairs. `method = "assignment"` is the paper's null:
arbitrators in decreasing degree each choose their cases without
replacement among the cases with a free seat. The 357 repeated pairs and 20
repeated trios lie far above the null (the paper reports z-scores of 4.81
and 10.40 from 1,000 draws).

```{r icsid-repeats}
hg_null_test(tribunals_all, statistic = c("repeated_pairs", "repeated_edges"),
             method = "assignment", n = 999, seed = 1)
```

## Methods

### Hyperedge centralities

The centrality analysis of the paper runs on the tribunals active on the
last observed date. `hypergraph_snapshot()` with `mode = "active"` returns
them, and `multiedges = FALSE` collapses two tribunals with the same three
members into one hyperedge.

```{r icsid-active}
active <- hypergraph_snapshot(tribunals, at = observation_end,
                              mode = "active", multiedges = FALSE)
active
```

`hg_edge_centrality()` builds the s-line graph, in which two hyperedges are
adjacent when they share at least `s` nodes, and computes shortest-path
betweenness and closeness of the hyperedges on it with the NetworkX
normalisation the paper uses. `top = 10` keeps the ten most central cases.

```{r icsid-betweenness}
top_betweenness <- hg_edge_centrality(active, s = 1, measure = "betweenness",
                                      top = 10)
top_betweenness
```

`hg_subset()` takes that table and cuts the snapshot down to those cases.

```{r icsid-betweenness-subset}
central_tribunals <- hg_subset(active, edges = top_betweenness)
central_tribunals
```

`plot()` draws them coloured by economic sector: the paper's Figure 6a. The
names are the archive's.

```{r icsid-betweenness-plot, fig.width = 8, fig.height = 7}
plot(central_tribunals, color_by = "economic_sector",
     legend_title = "economic sector", label_size = 2.8)
```

Closeness ranks a different set of tribunals (Figure 6b).

```{r icsid-closeness}
top_closeness <- hg_edge_centrality(active, s = 1, measure = "closeness",
                                    top = 10)
top_closeness
```

```{r icsid-closeness-subset}
close_tribunals <- hg_subset(active, edges = top_closeness)
```

```{r icsid-closeness-plot, fig.width = 8, fig.height = 7}
plot(close_tribunals, color_by = "economic_sector",
     legend_title = "economic sector", label_size = 2.8)
```

The static aggregate ranks tribunals that never overlapped in time. As for
the active snapshot, `multiedges = FALSE` collapses the two tribunals with
the same three members into one hyperedge; that binary aggregate is the
representation of the paper's Figure 6d and of the reproduction runner.

```{r icsid-aggregate-binary}
tribunals_binary <- hypergraph_snapshot(tribunals, mode = "cumulative",
                                        multiedges = FALSE)
tribunals_binary
```

On its seven most central cases by closeness, `linetype_by` outlines the
concluded cases differently from the pending ones, the distinction the
paper draws with dashed and solid lines.

```{r icsid-aggregate}
top_aggregate <- hg_edge_centrality(tribunals_binary, s = 1,
                                    measure = "closeness", top = 7)
top_aggregate
```

```{r icsid-aggregate-subset}
aggregate_tribunals <- hg_subset(tribunals_binary, edges = top_aggregate)
```

```{r icsid-aggregate-plot, fig.width = 8, fig.height = 7}
plot(aggregate_tribunals, color_by = "economic_sector",
     linetype_by = "is_concluded",
     legend_title = "economic sector", label_size = 2.8)
```

The paper's Figure 6c ranks the edges of the temporal graph by ordinary
edge betweenness; that is a graph computation on the clique projection,
`hg_project(active, method = "clique")`, and is left to a graph engine.

### Motifs

`hg_motifs()` counts the three induced four-node motifs of a 3-uniform
hypergraph, Y (two of the four possible triples), T (three) and O (all
four), and draws configuration-model null hypergraphs by the pairwise
reshuffle of HypergraphX, preserving degrees and cardinalities. The result
holds the observed count, the null mean and standard deviation, a z-score
and an empirical p-value per motif. On the aggregate hypergraph the Y motif
occurs 478 times against a null mean near 290, and the T motif 7 times
where the null almost never produces one; the O motif never occurs. The
paper uses 1,000 draws; 300 are used here.

```{r icsid-motifs}
motifs_aggregate <- hg_motifs(tribunals_all, n = 300, seed = 1)
motifs_aggregate
```

`plot()` draws the null distribution with the observed count: the paper's
Figure 7c.

```{r icsid-motifs-plot, fig.height = 4}
plot(motifs_aggregate)
```

The last snapshot of the temporal hypergraph has far fewer Y motifs, 48,
and a smaller but still clear excess over its null (Figure 7b).

```{r icsid-motifs-active}
motifs_active <- hg_motifs(active, n = 300, seed = 1)
motifs_active
```

```{r icsid-motifs-active-plot, fig.height = 4}
plot(motifs_active)
```

### Communities across representations

The paper clusters the GFCC data eight ways. Four representations are
hypergraph-derived: the association graph of `hg_project()`, in which a
block of size |e| gives each of its member pairs the weight 1 / (|e| - 1),
on the binary (`bh`) or multi (`mh`) hypergraph, with (`bhs`, `mhs`) or
without self-association, the term that joins the citing decision to the
decisions it cites. Four are classic: the citation graph, binary (`bg`) or
multi (`mg`), directed or undirected (`bgu`, `mgu`). The clustering uses
backward citations only, those whose citing decision is later than the
cited one, so the temporal hypergraph is rebuilt from that subset of the
citation table.

```{r gfcc-backward-citations}
backward_citations <- subset(gfcc_citations, date_citing > date_cited)
head(backward_citations)
```

```{r gfcc-backward}
backward <- temporal_hypergraph(
  backward_citations, actor = "cited", group = "block",
  time = "date_citing", nodes = gfcc_decisions, sparse = TRUE
)
backward
```

```{r gfcc-backward-aggregate}
backward_all <- hypergraph_snapshot(backward, mode = "cumulative")
backward_all
```

`hg_communities()` runs Infomap repeatedly on the chosen projection,
compares every pair of partitions with adjusted mutual information and keeps
the run with the largest summed AMI as the medoid. The paper uses 50 seeds
with 100 trials each; five seeds with five trials are used here to keep the
document quick. The eight calls differ only in the projection they ask
for. The four classic citation graphs first: `method = "citation"` with
`edge_source` naming the citing decision, `duplicate_edges = "collapse"`
for the binary graph, `directed = TRUE` for the directed one.

```{r gfcc-communities-graphs, eval = has_igraph}
communities_bg <- hg_communities(backward_all, n_runs = 5, trials = 5,
                                 method = "citation", edge_source = "citing",
                                 duplicate_edges = "collapse", directed = TRUE)
communities_bgu <- hg_communities(backward_all, n_runs = 5, trials = 5,
                                  method = "citation", edge_source = "citing",
                                  duplicate_edges = "collapse")
communities_mg <- hg_communities(backward_all, n_runs = 5, trials = 5,
                                 method = "citation", edge_source = "citing",
                                 directed = TRUE)
communities_mgu <- hg_communities(backward_all, n_runs = 5, trials = 5,
                                  method = "citation", edge_source = "citing")
communities_bg
```

Then the four association graphs: the default projection, on the binary or
the multi-hypergraph, with or without `self_association`.

```{r gfcc-communities-hypergraphs, eval = has_igraph}
communities_bh <- hg_communities(backward_all, n_runs = 5, trials = 5,
                                 duplicate_edges = "collapse")
communities_bhs <- hg_communities(backward_all, n_runs = 5, trials = 5,
                                  duplicate_edges = "collapse",
                                  self_association = TRUE, edge_source = "citing")
communities_mh <- hg_communities(backward_all, n_runs = 5, trials = 5)
communities_mhs <- hg_communities(backward_all, n_runs = 5, trials = 5,
                                  self_association = TRUE, edge_source = "citing")
communities_mhs
```

`hg_compare_communities()` sets the fits side by side. With the hypergraph
passed as `hg`, every medoid is also scored on its own projection. The
summary counts the communities, singletons and the balance between the two
largest clusters of each medoid. The association graphs without
self-association leave hundreds of uncited decisions as singletons; with
self-association those decisions join the clusters of the decisions they
cite, and the largest clusters stay balanced.

```{r gfcc-compare, eval = has_igraph}
comparison <- hg_compare_communities(
  bg = communities_bg, bgu = communities_bgu, mg = communities_mg,
  mgu = communities_mgu, bh = communities_bh, bhs = communities_bhs,
  mh = communities_mh, mhs = communities_mhs, hg = backward_all,
  edge_source = "citing"
)
comparison
```

`plot()` draws the cluster-size distributions, the paper's Figure 8b.

```{r gfcc-compare-plot, eval = has_igraph, fig.height = 4}
plot(comparison)
```

The similarity table holds the AMI, ARI and NMI between every pair of
medoids.

```{r gfcc-similarity-table, eval = has_igraph}
similarity <- as.data.frame(comparison, what = "similarity")
head(similarity)
```

`plot(what = "similarity")` is the paper's Figure 8c with AMI below and ARI
above the diagonal. Clusterings of hypergraph-derived representations agree
more with each other than with those of the classic graphs.

```{r gfcc-similarity, eval = has_igraph, fig.width = 6, fig.height = 5}
plot(comparison, what = "similarity")
```

The quality table holds the graph-partition diagnostics of the paper for
each medoid on its own projection: coverage, weighted coverage, performance,
modularity, and the worst community conductance.

```{r gfcc-quality, eval = has_igraph}
quality <- as.data.frame(comparison, what = "quality")
quality
```

## What the reproduction covers

Every figure and table of the paper that concerns the hypergraphs is
produced above from the released data: Table 1 by `summary()`, Table 2 by
`hg_representations()`, Figure 3 by `hg_subset()` and `plot()`, Figures 4
and 5 by `hg_growth()`, `hg_edges()` and `hg_measures()`, Figure 6 by
`hg_edge_centrality()`, Figure 7 by `hg_motifs()`, Figure 8 by
`hg_communities()` and `hg_compare_communities()`, and the repeated
collaboration test by `hg_null_test()`. The full-data runner
`benchmarks/run_legal_hypergraphs_reproduction.R` checks the association
graphs, the archived medoids and the Figure 6 centralities against the
authors' archived results to numerical tolerance: the active betweenness
ranking is led by ARB/17/21 at 0.02654013 and the aggregate closeness
ranking by ARB/15/5 at 0.54805703, and the motif census of the aggregate is
Y = 478, T = 7, O = 0.

## References

Coupette, C., Hartung, D., and Katz, D. M. (2024). Legal hypergraphs.
Philosophical Transactions of the Royal Society A, 382(2270), 20230141.

Coupette, C., Hartung, D., and Katz, D. M. (2024). Legal hypergraphs
(Software). Zenodo, record 8081507.

Lotito, Q. F., Contisciani, M., De Bacco, C., Di Gaetano, L., Gallo, L.,
Montresor, A., Musciotto, F., Ruggeri, N., and Battiston, F. (2023).
Hypergraphx: a library for higher-order network analysis. Journal of
Complex Networks, 11(3), cnad019.

Rosvall, M., and Bergstrom, C. T. (2008). Maps of random walks on complex
networks reveal community structure. Proceedings of the National Academy of
Sciences, 105(4), 1118-1123.
