---
title: "Topic descriptions under several scores: Levebee mathematics feedback"
author: "honets"
date: "2026-09-06"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "", fig.width = 9, dpi = 110,
                      out.width = "100%")
library(honets)
has_torch <- requireNamespace("torch", quietly = TRUE) &&
  torch::torch_is_installed()
```

> The same corpus the **sbert** package uses in its "Semantic Topics in
> Levebee AI Mathematics Feedback" vignette, clustered as a hypergraph and
> described with `hg_keywords()` under several scores at once. Every number
> and bar is recomputed at knit time.

# The corpus and its embeddings

`sbert::feedback_translations` holds 8,757 AI-generated feedback messages
from the Levebee mathematics application, translated into English. The sbert
vignette deduplicates them and works with the first 600 distinct messages,
whose `all-MiniLM-L6-v2` embeddings ship precomputed with that package. Using
the same 600 messages and the same vectors keeps this document offline and
directly comparable to the sbert topics.

```{r corpus}
sample <- readRDS(system.file("extdata", "feedback_embeddings.rds",
                              package = "sbert"))
str(sample, max.level = 1)
```

`sample$text` is the 600 messages and `sample$embeddings` their 600 x 384
MiniLM matrix, one row per message in the same order. Both go straight into
the constructors; documents are numbered `doc_1` to `doc_600` on both sides.

# Two hypergraphs, one partition

The clustering uses the semantic geometry: each message plus its ten nearest
neighbours in embedding space is one hyperedge (`construction = "knn"`), and
`hg_cluster()` cuts the hypergraph Laplacian's spectral embedding into six
groups, the count the sbert vignette settles on by coherence.

```{r cluster}
hg_knn <- text_hypergraph(sample$text, construction = "knn", k = 10,
                          embeddings = sample$embeddings)
hg_knn
topics <- hg_cluster(hg_knn, k = 6, seed = 1)
table(topics$cluster)
```

The kNN hypergraph has no words in it, so the description step uses the
lexical view of the same 600 messages: the bag-of-words document hypergraph,
whose token layer carries the counts, weights and per-document vocabulary
that the keyword scores need. Any partition whose names are these documents
can be described with it.

```{r bag}
hg_bag <- text_hypergraph(sample$text, weight = "tfidf",
                          stop_words = stop_words_en(), min_count = 2L)
hg_bag
```

# Four keyword views of the same six topics

`hg_keywords()` ranks each cluster's words under the score named in `type`;
several types at once return one block of rows per score:

* **frequency** -- what the topic says most (raw counts);
* **ctfidf** -- what the topic alone says (Grootendorst's class-based
  tf-idf, the BERTopic representation);
* **centrality** -- which words sit at the centre of the topic's own
  word hypergraph (words as nodes, the topic's messages as hyperedges;
  clique eigenvector centrality);
* **attention** -- which words the HyperGAT classifier attends to when
  it learns to predict the topic label from the message: the first
  attention layer's node-level weights (the layer that attends over the
  word embeddings), summed over the topic's messages.

The first three are cheap and deterministic. The fourth needs
\pkg{torch}: `hg_hypergat(what = "attention")` trains the network on the
cluster labels and returns every message's per-word attention, which
`hg_keywords(type = "attention")` sums per topic.

```{r views}
views <- hg_keywords(hg_bag, topics, n = 6,
                     type = c("frequency", "ctfidf", "centrality"))
hg_keywords(hg_bag, topics, n = 6, collapse = TRUE,
            type = c("frequency", "ctfidf", "centrality"))
```

```{r attention, eval = has_torch}
attention <- hg_hypergat(sample$text, labels = topics, epochs = 10,
                         seed = 1, what = "attention")
head(attention)
hg_keywords(hg_bag, topics, n = 6, scores = attention, collapse = TRUE)
```

Two facts about this table are worth checking before trusting the column.
First, the network was trained to predict the cluster labels, so its
attention explains *this* partition; its agreement with the labels on all
600 messages is the number to look at. Second, `attention_2`, the second
layer's weights, is uniform within every one-sentence message by
construction (all of a document's words receive the same first-layer output
when the document has a single hyperedge), which is why the keyword score
uses the first layer. Within a message the first layer does discriminate
("red" gets more than "should" or "two" below), but summed over a topic the
score still rises with how often a word occurs, so its rank order is close
to frequency's. The Spearman correlation between the two, over all
(topic, word) pairs, is computed here rather than asserted:

```{r attention-check, eval = has_torch}
fit <- hg_hypergat(sample$text, labels = topics, epochs = 10, seed = 1)
mean(fit$predicted == fit$label)
subset(attention, node == "doc_2")
both <- hg_keywords(hg_bag, topics, n = Inf, scores = attention,
                    type = "frequency")
paired <- merge(subset(both, type == "frequency"),
                subset(both, type == "attention"),
                by = c("cluster", "word"), suffixes = c("_freq", "_att"))
cor(paired$score_freq, paired$score_att, method = "spearman")
```

# The figure: one panel per topic and score

`plot()` on the keyword table draws horizontal bars of the score, one row of
panels per topic and one column per score type, each panel with its own
word axis, fill colour by type. The three lexical scores side by side:

```{r plot-three, fig.height = 13}
plot(views)
```

With the attention view added as a fourth column:

```{r plot-four, fig.height = 13, eval = has_torch}
four <- hg_keywords(hg_bag, topics, n = 6, scores = attention,
                    type = c("frequency", "ctfidf", "centrality"))
plot(four)
```

# Reading the disagreement

Where the columns agree, the topic owns its vocabulary: read the
choose-a-picture panel ("them", "choose", "pick", "select") across the
row. Where they disagree, each score is telling you something different.
Frequency promotes words the whole corpus shares ("one", "you"); c-TF-IDF
demotes them and surfaces the topic's own terms ("middle", "top", "gray");
centrality picks the words that co-occur with the most other words inside
the topic, which favours the hub of a template ("picture", "pictures");
attention is whatever the classifier found diagnostic for the topic label,
and summed per topic it stays close to frequency (the correlation above),
so read it as a frequency count re-weighted by what the classifier used.
None is *the* keyword list. A topic is characterised by the set, and the figure is the
compact way to read the set.

These word lists resemble the ones printed in the sbert vignette for the
same 600 messages (its topic 1, for instance, is "picture", "pictures",
"choose"). That is a reading of the two documents side by side, not a
computed comparison: sbert's topic model is not run here, and no agreement
statistic between the two partitions is reported.

```{r session}
sessionInfo()
```
