---
title: "Memory networks: states that remember how you got there"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this document still knits (without figures)
# where cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

> Each section below is one verb, worked end to end on the same data. The
> sections run in order in a single R session, so later ones may reuse
> objects the earlier ones built.

## build_hon(): higher-order networks from sequences {#build_hon}

### The analytic question

A transition network answers "what follows what". It cannot answer "what
follows what, *given how you got here*" — because a first-order node is just
a state, with no memory of the path that reached it.

`build_hon()` builds a network whose nodes carry that memory. A node like
`Specify -> Command` is the state `Command` together with the fact that
`Specify` preceded it, and it keeps its own continuation distribution. The
network only creates such a node when the evidence justifies it: the
BuildHON+ rule of Xu, Wickramarathne & Chawla (2016) adds a higher-order
context **only if** its continuation distribution differs meaningfully from
the lower-order one it would otherwise inherit.

```{r build-hon-load}
library(honets)
```

```{r build-hon-cograph, eval = has_cograph}
library(cograph)
```

```{r build-hon-simpguard, include = FALSE, eval = has_cograph}
# grDevices::chull() aborts on a degenerate (collinear or single-point)
# pathway cloud. cograph >= 2.3.5 fixes this in .smooth_blob(); shadowing the
# function here keeps the render alive on CRAN cograph (2.1.1) by falling
# back to splot(). Scope is this vignette's knit environment only.
plot_simplicial <- function(...) {
  args <- list(...)
  tryCatch(do.call(cograph::plot_simplicial, args),
           error = function(e) {
             warning("plot_simplicial fell back to splot(): ",
                     conditionMessage(e), call. = FALSE)
             splot(args[[1]])
           })
}
```

The data are `human_long`: 10,796 coded human turns from 429 human-AI
pair-programming sessions, nine action codes.

### Building one

Long-format event data goes in directly — name the action, actor and time
columns and `build_hon()` does the splitting.

```{r build-hon-build}
hon <- build_hon(human_long,
                 action = "code", actor = "session_id", time = "timestamp",
                 max_order = 3, min_freq = 5)
hon
```

Nine states, but **233 nodes**. The extra 224 are memory contexts that
earned their place. `max_order = 3` is a ceiling, not a target: the rule
extraction stops promoting a context as soon as the higher order stops
paying.

`summary()` adds the order breakdown:

```{r build-hon-summary}
summary(hon)
```

### Reading the rules

`as.data.frame()` is the accessor. Each row is one extracted rule: a source
context, a target state, how often it occurred, and the conditional
probability.

```{r build-hon-rules}
as.data.frame(hon, sort_by = "count", top = 6)
```

Read the first row as: *of all the times `Specify` was followed by
something, 26.4% of the time it was `Command`* — 740 occurrences. That one
is first-order (`from_order = 1`), which is the baseline the higher-order
rules have to beat.

The genuinely higher-order rules are the ones with `from_order >= 2`:

```{r build-hon-higher}
as.data.frame(hon, order_min = 2, sort_by = "count", top = 6)
```

### Seeing it

**Do not draw the higher-order network itself.** It has 233 nodes, one per
memory context, and `splot()` on it is an unreadable hairball. The network
you draw is the **first-order** one — nine states — and the higher-order
structure goes on top of it as *pathways*.

```{r build-hon-net1, eval = has_cograph}
net1 <- build_hon(human_long, action = "code", actor = "session_id",
                  time = "timestamp", max_order = 1, min_freq = 5)
nodes_1 <- as.data.frame(net1, what = "nodes")
rules_1 <- as.data.frame(net1)
c(nodes = nrow(nodes_1), edges = nrow(rules_1))
```

`net_hon` is dual-classed `cograph_network`, so cograph draws it with no
conversion step.

```{r build-hon-splot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "First-order network of the nine action codes"}
splot(net1, minimum = 0.05,
      title = "Human actions in AI-assisted coding")
```

`plot_simplicial()` puts the extracted rules on that same nine-node layout:
each blob is one context and the state it predicts, sources in blue and the
target in red. Repeated states get distinct positions, so
`Command -> Command -> Correct` covers three visible nodes rather than
collapsing onto one. Pass the rules through `pathways =` — this is the same
table `as.data.frame()` returns, read as shape rather than as rows.

```{r build-hon-simplicial, eval = has_cograph, fig.width = 11, fig.height = 11, fig.alt = "Six strongest higher-order rules drawn as simplicial blobs"}
top_rules <- as.data.frame(hon, order_min = 2, sort_by = "count", top = 6)
plot_simplicial(net1,
                pathways = top_rules,
                max_pathways = 6,
                title = "Six strongest extracted rules")
```

### How much of the network is higher-order?

```{r build-hon-orders}
rules <- as.data.frame(hon)
table(rules$from_order)
```

81 first-order rules — the full 9×9 transition structure, since on data this
dense every pair occurs. Then 441 second-order and 297 third-order rules
were added on top. Roughly nine in ten rules in this network encode memory.

### `min_freq` is the consequential choice

A context is only considered if it occurs at least `min_freq` times. This is
the main lever on network size, and it is a statement about how much
evidence you require before believing a memory effect.

```{r build-hon-minfreq}
sizes <- do.call(rbind, lapply(c(5, 20, 50, 100), function(mf) {
  h <- build_hon(human_long, action = "code", actor = "session_id",
                 time = "timestamp", max_order = 3, min_freq = mf)
  h_nodes <- as.data.frame(h, what = "nodes")
  h_rules <- as.data.frame(h)
  data.frame(min_freq = mf, nodes = nrow(h_nodes), edges = nrow(h_rules),
             max_order_observed = max(h_rules$from_order))
}))
sizes
```

The network shrinks from 233 nodes to 15 as the evidence bar rises. All four
still observe order-3 contexts — the *strongest* memory effects survive even
a demanding threshold, which is reassuring. Report the threshold with any
result: a claim about a 233-node network and a claim about a 15-node network
are different claims.

### Accepted input

`build_hon()` takes sequences in whichever shape you have them.

```{r build-hon-inputs}
# a list of character vectors
build_hon(list(c("A", "B", "C", "A", "B"), c("B", "C", "A", "B")),
          max_order = 2, min_freq = 1)

# a wide data.frame, one sequence per row, trailing NAs stripped
build_hon(data.frame(V1 = c("A", "B"), V2 = c("B", "C"), V3 = c("C", NA)),
          max_order = 2, min_freq = 1)
```

It also accepts a fitted `tna` or `netobject` model directly, decoding
integer-coded states back to their labels.

### What to do next

- **[`markov_order_test()`](memory-networks.html#markov_order_test)** — the prior question.
  Is there memory in these sequences at all? Run it before reporting a
  higher-order network as evidence of anything.
- **[`path_dependence()`](memory-networks.html#path_dependence)** — *which* contexts carry
  the memory, one row per context, with the divergence that justifies it.
- **[`bootstrap_hon()`](memory-networks.html#hon_inference)** — do these rules survive
  resampling? A rule extracted once is not yet a finding.
- **[`hon_centrality()`](memory-networks.html#hon_centrality)** — rank the states using the
  higher-order topology.
- **[`build_honem()`](memory-networks.html#build_honem)** — embed the nodes in a vector
  space.
- **[`build_simplicial(type = "pathway")`](simplicial-complexes.html#build_simplicial)** — read
  the same fitted network as a simplicial complex.

### References

- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.
- Saebi, M., Xu, J., Kaplan, L. M., Ribeiro, B., & Chawla, N. V. (2020).
  Efficient modeling of higher-order dependencies in networks: from
  algorithm to application for anomaly detection. *EPJ Data Science*, 9, 15.


---

## markov_order_test(): does memory exist at all? {#markov_order_test}

### The analytic question

Every higher-order method in this package assumes the sequences have memory.
`markov_order_test()` is the question you should ask **before** any of them:
does the past predict the future beyond the present state, and if so, how far
back?

It fits Markov chains of increasing order, compares them by likelihood-ratio,
and — because the asymptotic chi-square reference is unreliable when the
transition table is sparse — calibrates the statistic against a **permutation
null** built by shuffling within sequences.

```{r markov-order-test-load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
wide <- as.data.frame(do.call(rbind, lapply(
  seqs, function(s) c(s, rep(NA, max(lengths(seqs)) - length(s))))),
  stringsAsFactors = FALSE)
```

### Running it

```{r markov-order-test-run}
mk <- markov_order_test(wide, max_order = 3, n_perm = 200, seed = 1)
mk
```

Three criteria, and they **disagree**: BIC selects order 1, AIC and the
permutation LRT both select order 3. That disagreement is the result, not a
problem to resolve away.

```{r markov-order-test-table}
as.data.frame(mk)
```

### Reading the disagreement

Look at the `df` column. Going from order 2 to order 3 costs **4,721 extra
parameters** on 10,792 observations. AIC charges 2 per parameter; BIC charges
`log(n)` ≈ 9.3. On data this size the two criteria are answering different
questions:

- **AIC** asks which order predicts best out of sample. It says 3.
- **BIC** asks which order is most likely to be the true generating model. It
  says 1, because the parameter cost of a third-order chain is enormous
  relative to the evidence.

Both are right about their own question. For *description* — which is what
`build_hon()` does — the AIC/LRT answer is the relevant one, and it is why
`max_order = 3` is a sensible ceiling for these data. For a claim that the
process **is** third-order Markov, BIC's scepticism should be reported.

### Seeing the test

`plot()` puts the two panels side by side: the information criteria by
order, and the observed statistic against its permutation null. The second
panel is the one to read — it shows *how far* outside the null the observed
value sits, which the p-value's floor cannot express.

```{r markov-order-test-mkplot, fig.width = 12, fig.height = 6, fig.alt = "Information criteria by order, and observed statistic against the permutation null"}
plot(mk)
```

### The permutation p-value has a floor

```{r markov-order-test-pfloor}
as.data.frame(mk)
```

Every permutation p-value is 0.004975 — that is `1 / (200 + 1)`, the
**smallest value 200 permutations can produce**. It does not mean the three
orders are equally significant; it means all three exceeded every one of
their 200 null replicates. Raise `n_perm` if you need to resolve below that
floor.

Note also how far the asymptotic p-values sit from it: 1.7e-129 at order 2.
That is the sparsity problem the permutation null exists to avoid — with 576
degrees of freedom over nine states, the chi-square approximation is not
trustworthy, and the permutation value is the one to report.

### The null distribution itself

`what = "null"` returns the replicates, so the calibration can be inspected
rather than trusted.

```{r markov-order-test-null}
nulls <- as.data.frame(mk, what = "null")
as.data.frame(mk, what = "null", top = 4)
aggregate(g2 ~ order, data = nulls,
          FUN = function(x) round(c(mean = mean(x), max = max(x)), 1))
```

Compare each order's observed `g2` in the table above against the maximum its
null ever reached. The observed statistics are far outside — which is what
"p at the floor" means, made concrete.

### Cost and reproducibility

Permutations dominate the runtime. `parallel = TRUE` uses
`parallel::mclapply` (skipped on Windows), and all randomness is drawn
serially before any parallel work, so **a parallel run reproduces the serial
result exactly under the same seed**.

```{r markov-order-test-parallel, eval = FALSE}
markov_order_test(wide, max_order = 3, n_perm = 2000,
                  parallel = TRUE, n_cores = 4, seed = 1)
```

Always pass `seed`. A permutation test reported without one is not
reproducible.

### What to do next

- **[`build_hon()`](memory-networks.html#build_hon)** — if there is memory, build the network
  that represents it. Use the selected order as the `max_order` ceiling.
- **[`path_dependence()`](memory-networks.html#path_dependence)** — this test is global. That
  one localises the memory to individual contexts.
- **[`build_mogen()`](memory-networks.html#build_mogen)** — the same model-selection question
  under a multi-order generative model, which mixes orders rather than
  choosing one.

### References

- Anderson, T. W., & Goodman, L. A. (1957). Statistical inference about
  Markov chains. *The Annals of Mathematical Statistics*, 28(1), 89-110.
- Good, P. (2005). *Permutation, Parametric and Bootstrap Tests of
  Hypotheses*. Springer.


---

## path_dependence(): which contexts carry the memory {#path_dependence}

### The analytic question

[`markov_order_test()`](memory-networks.html#markov_order_test) answers a global question:
*is there memory in these sequences?* It returns one verdict for the whole
corpus.

`path_dependence()` localises it. For every order-k context it asks: **how
different is what happens next, once you know the context, from what happens
next if you only know the current state?** The answer is a Kullback-Leibler
divergence in bits, one row per context, so memory stops being a property of
the dataset and becomes a property of specific situations.

```{r path-dependence-load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
wide <- as.data.frame(do.call(rbind, lapply(
  seqs, function(s) c(s, rep(NA, max(lengths(seqs)) - length(s))))),
  stringsAsFactors = FALSE)
```

### Running it

```{r path-dependence-run}
pd <- path_dependence(wide, order = 2)
pd
```

Two chain-level numbers summarise the corpus: knowing the previous action as
well as the current one buys **0.143 bits** on average, and cuts the
uncertainty about the next action by **0.109 bits**. Those are modest
averages hiding a very uneven distribution.

### The uneven part

```{r path-dependence-top}
as.data.frame(pd, sort_by = "KL", top = 8)
```

`Interrupt -> Interrupt` diverges by **1.065 bits** — seven times the corpus
average. Its entropy drops from 3.02 to 2.03 bits, meaning that once you know
an interruption followed an interruption, you have roughly halved your
uncertainty about what comes next.

### Seeing which contexts carry it

`plot()` is a dot plot of per-context divergence, sized by how often the
context occurs. The long right tail is the point: memory is concentrated in
a handful of situations, not spread evenly.

```{r path-dependence-pdplot, fig.width = 10.5, fig.height = 6, fig.alt = "Per-context KL divergence, sized by context count"}
plot(pd)
```

### `flips` is the column that matters most

`flips` is `TRUE` when the context changes the *modal* prediction — the
single most likely next action differs from what the first-order model would
have said.

```{r path-dependence-flips}
pd_all <- as.data.frame(pd)
table(flips = pd_all$flips)
```

35 of the 81 contexts flip the prediction. That is the practical statement of
what memory buys here: in **43% of situations, a first-order model would name
the wrong most-likely next action**.

```{r path-dependence-flipex}
as.data.frame(pd, sort_by = "KL", top = 5)
```

Read `Interrupt -> Interrupt`: a first-order model predicts `Command` next
(because `Command` is what usually follows `Interrupt`), but conditioned on
the repeat, the modal continuation is another `Interrupt`. Interruptions
cluster, and only the higher-order view sees it.

### KL and entropy drop are different questions

Both columns are reported because they answer different things.

```{r path-dependence-twoviews}
as.data.frame(pd, sort_by = "KL", top = 6)
```

`H_drop` is how much *uncertainty* the context removes. `KL` is how far the
context's distribution has *moved* from the first-order one. A context can
move a long way without becoming more predictable — `Frustrate -> Specify`
has a high KL (0.410) on a small entropy drop (0.154), meaning it reshuffles
which continuations are likely without concentrating them.

### `min_count` guards against noise

Divergence is easy to manufacture from a handful of observations. Contexts
below `min_count` are dropped.

```{r path-dependence-mincount}
do.call(rbind, lapply(c(5, 20, 50), function(mc) {
  p <- path_dependence(wide, order = 2, min_count = mc)
  d <- as.data.frame(p)
  data.frame(min_count = mc, contexts = nrow(d),
             mean_KL = round(mean(d$KL), 3),
             flips = sum(d$flips))
}))
```

The mean KL *falls* as the threshold rises — from 0.186 to 0.157 — which is
evidence that some of the apparent memory at `min_count = 5` is small-sample
noise. The effect is mild here: only two contexts fall below 20 occurrences,
and the flip count holds at 35 until the threshold reaches 50, where 20
sparse contexts drop out and 27 flips remain. That stability is itself
reassuring. Report the threshold regardless.

### Higher orders

```{r path-dependence-order3}
pd3 <- path_dependence(wide, order = 3, min_count = 20)
pd3
```

### What to do next

- **[`markov_order_test()`](memory-networks.html#markov_order_test)** — the global test this
  verb decomposes. Run it first.
- **[`build_hon()`](memory-networks.html#build_hon)** — the contexts with high KL are exactly
  the ones BuildHON+ promotes to higher-order nodes.
- **[`bootstrap_hon()`](memory-networks.html#hon_inference)** — before reporting a specific
  context as a finding, check that it survives resampling.

### References

- Cover, T. M., & Thomas, J. A. (2006). *Elements of Information Theory*
  (2nd ed.). Wiley.
- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.


---

## build_mogen(): choosing an order instead of assuming one {#build_mogen}

### The analytic question

[`build_hon()`](memory-networks.html#build_hon) decides *per context* whether memory is
warranted. `build_mogen()` asks the question at the level of the whole model:
fit Markov chains of every order from 0 up to `max_order`, score each one,
and report which order the data actually support.

The multi-order model of Scholtes (2017) treats the layers as a *nested
family* — each order's likelihood is computed on the paths it can explain,
with the degrees of freedom that order costs — so the comparison is a
like-for-like model selection rather than a race between separately-fitted
models.

```{r build-mogen-load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
```

```{r build-mogen-cograph, eval = has_cograph}
library(cograph)
```

```{r build-mogen-simpguard, include = FALSE, eval = has_cograph}
# grDevices::chull() aborts on a degenerate (collinear or single-point)
# pathway cloud. cograph >= 2.3.5 fixes this in .smooth_blob(); shadowing the
# function here keeps the render alive on CRAN cograph (2.1.1) by falling
# back to splot(). Scope is this vignette's knit environment only.
plot_simplicial <- function(...) {
  args <- list(...)
  tryCatch(do.call(cograph::plot_simplicial, args),
           error = function(e) {
             warning("plot_simplicial fell back to splot(): ",
                     conditionMessage(e), call. = FALSE)
             splot(args[[1]])
           })
}
```

### Fitting

```{r build-mogen-fit}
mo <- build_mogen(seqs, max_order = 4)
mo
```

The print method flags the minimum of each criterion, and here they
**disagree**: AIC selects order 4, BIC selects order 1. The asterisks make
the disagreement visible rather than hiding it behind a single "optimal
order".

### The selection table

```{r build-mogen-table}
as.data.frame(mo)
```

The `dof` column explains everything. Going from order 3 to order 4 adds
3,102 parameters (`layer_dof`) on 10,792 observations. The log-likelihood
improves substantially — from −17,163 to −11,366 — so AIC, which charges 2
per parameter, is delighted. BIC charges `log(10792)` ≈ 9.3 per parameter and
is appalled: its score rises from 42,402 at order 1 to 81,061 at order 4.

`optimal` marks the order selected under the criterion you asked for
(`criterion = "aic"` by default).

### Seeing the selection

`plot()` draws both information criteria against order, marking the
selected one. The AIC and BIC curves crossing in different places *is* the
disagreement the next section discusses.

```{r build-mogen-icplot, eval = has_cograph, fig.width = 8, fig.height = 5, fig.alt = "AIC and BIC by Markov order, with the selected order marked"}
plot(mo)
```

### Which criterion should you use?

It depends on what the model is for.

- **Describing the corpus** — AIC. You want the order that best captures the
  structure actually present, and over-parameterisation is a cost you are
  willing to pay for fidelity.
- **Claiming the process is order-k** — BIC. It is consistent for model
  identification, and its verdict of order 1 is a real warning that a
  fourth-order chain is not *identifiable* from 425 sessions.
- **Testing whether order k improves on k−1** — neither; use
  [`markov_order_test()`](memory-networks.html#markov_order_test), which calibrates the
  likelihood-ratio statistic against a permutation null rather than an
  information criterion.

```{r build-mogen-bic}
mo_bic <- build_mogen(seqs, max_order = 4, criterion = "bic")
as.data.frame(mo_bic)
```

Switching the criterion changes the selected order, and it should be reported
alongside the result.

### The fitted transitions

`mogen_transitions()` is the tidy accessor for the fitted layer.

```{r build-mogen-trans}
mogen_transitions(mo, top = 5)
```

At order 4, a "state" is a four-step history. The heaviest is
`Specify -> Command -> Request -> Specify` continuing to `Interrupt`, 127
times at probability 0.72 — a highly deterministic recurring routine, and the
kind of structure that only a high-order layer can express.

The same table is reachable through the standard accessor, and `order =`
selects a different layer:

```{r build-mogen-layers}
as.data.frame(mo, what = "transitions", order = 2, top = 4)
```

### The fitted layer as shape

`plot_simplicial()` takes a `mogen_transitions()` table directly through
`pathways =`, so the fitted second-order layer can be read as shape over the
network rather than as 702 rows. The network it draws on is the
**first-order** one — nine states. Because `mogen_transitions()` returns
rows already sorted by count, `top = N` picks the N heaviest.

```{r build-mogen-mogensimp, eval = has_cograph, fig.width = 11, fig.height = 11, fig.alt = "Six heaviest fitted second-order pathways as simplicial blobs"}
net1 <- build_hon(seqs, max_order = 1)
top_paths <- mogen_transitions(mo, order = 2, top = 6)
plot_simplicial(net1, pathways = top_paths,
                max_pathways = 6,
                title = "Six heaviest second-order pathways")
```

A single pathway, isolated:

```{r build-mogen-mogenone, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "The single heaviest fitted pathway"}
top_path <- mogen_transitions(mo, order = 2, top = 1)
plot_simplicial(net1, pathways = top_path,
                max_pathways = 1)
```

### Path counts without fitting a model

`path_counts()` is the descriptive counterpart: how often does each k-step
path occur, with no model behind it.

```{r build-mogen-paths}
path_counts(seqs, k = 3, top = 6)
```

Useful as a sanity check before fitting, and for reporting the raw evidence
under a modelled claim.

### A caution on `max_order`

Raising `max_order` is not free. The parameter count grows roughly as
`|S|^(k+1)`, and past a certain point the higher layers are fitting
individual sequences.

```{r build-mogen-cost}
do.call(rbind, lapply(2:4, function(k) {
  m <- build_mogen(seqs, max_order = k)
  d <- as.data.frame(m)
  selected <- subset(d, optimal)
  data.frame(max_order = k, selected_aic = selected$order,
             top_layer_dof = max(d$layer_dof))
}))
```

If AIC keeps selecting whatever ceiling you set, that is a sign the ceiling
is doing the choosing, not the data. Check BIC and the permutation test
before believing it.

### What to do next

- **[`markov_order_test()`](memory-networks.html#markov_order_test)** — the permutation-based
  version of this question, which does not rely on an information criterion.
- **[`build_hon()`](memory-networks.html#build_hon)** — per-context order selection instead
  of one order for the whole model.
- **[`build_hypa()`](memory-networks.html#build_hypa)** — which specific paths occur more or
  less often than a null model expects.
- **[`pathways()`](overview.html#pathways)** — hand the fitted model's paths to the
  simplicial and hypergraph families.

### References

- Scholtes, I. (2017). When is a network a network? Multi-order graphical
  model selection in pathways and temporal networks. *KDD 2017*, 1037-1046.
- Gote, C., & Scholtes, I. (2023). Predicting sequences of traversed nodes in
  graphs using network models with multiple higher orders. *Data Mining and
  Knowledge Discovery*.


---

## build_hypa(): which paths are anomalous {#build_hypa}

### The analytic question

A frequent path is not necessarily a surprising one. `Specify -> Command` is
common largely because `Specify` and `Command` are both common — the count
tells you about the states, not about the pairing.

`build_hypa()` (LaRock et al. 2020) separates the two. It asks: given how
often each state is entered and left, **how often would this path occur by
chance?** The null is a Wallenius non-central hypergeometric model, which
holds the marginal in- and out-degrees fixed and lets the paths fall where
they may. A path that beats its own expectation is carrying information the
degree structure does not explain.

```{r build-hypa-load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
```

```{r build-hypa-cograph, eval = has_cograph}
library(cograph)
```

```{r build-hypa-simpguard, include = FALSE, eval = has_cograph}
# grDevices::chull() aborts on a degenerate (collinear or single-point)
# pathway cloud. cograph >= 2.3.5 fixes this in .smooth_blob(); shadowing the
# function here keeps the render alive on CRAN cograph (2.1.1) by falling
# back to splot(). Scope is this vignette's knit environment only.
plot_simplicial <- function(...) {
  args <- list(...)
  tryCatch(do.call(cograph::plot_simplicial, args),
           error = function(e) {
             warning("plot_simplicial fell back to splot(): ",
                     conditionMessage(e), call. = FALSE)
             splot(args[[1]])
           })
}
```

### Running it

```{r build-hypa-run}
hp <- build_hypa(seqs, order = 2)
hp
```

Of the 702 second-order paths, 139 occur significantly more often than
expected and 46 significantly less, after Benjamini-Hochberg correction.

### Over-represented paths

```{r build-hypa-over}
as.data.frame(hp, what = "over", sort_by = "ratio", top = 6)
```

The top row occurs **10.3 times** more often than the degree structure
predicts. That is the signature of a routine — a sequence the sessions
actually follow, rather than one the marginals happen to produce.

### Under-represented paths are the more interesting half

```{r build-hypa-under}
as.data.frame(hp, what = "under", sort_by = "ratio", top = 4)
```

`Specify -> Specify -> Command` is observed 77 times against an expectation
of 126 — a ratio of 0.61. It is a *frequent* path (77 occurrences is not
rare) that is nonetheless **avoided**. A count-based analysis would list it
among the common transitions and miss entirely that it happens far less than
it should.

Absences are hard to see and easy to under-report. This is the column that
finds them.

### Seeing the anomalies

`plot_simplicial()` dispatches on `net_hypa` and ranks pathways by ratio, so
the most extreme deviations come first. `dismantled = TRUE` gives each path
its own panel, which is what makes overlapping routines readable.

```{r build-hypa-overplot, eval = has_cograph, fig.width = 14, fig.height = 14, fig.alt = "Nine most over-represented paths, one panel each"}
plot_simplicial(hp, anomaly = "over", max_pathways = 9,
                dismantled = TRUE, ncol = 3)
```

The under-represented half — the paths the sessions *avoid* — is the harder
one to see in a table and the easier one to see here:

```{r build-hypa-underplot, eval = has_cograph, fig.width = 14, fig.height = 7, fig.alt = "Six most under-represented paths, one panel each"}
plot_simplicial(hp, anomaly = "under", max_pathways = 6,
                dismantled = TRUE, ncol = 3)
```

### Ratio and p-value rank differently

```{r build-hypa-ranks}
as.data.frame(hp, what = "over", sort_by = "ratio", top = 3)
as.data.frame(hp, what = "over", sort_by = "observed", top = 3)
```

The most *surprising* paths and the most *frequent* surprising paths are
different sets. Sort by `ratio` to find structure; sort by `observed` to find
structure that carries weight. Report which you used.

### Multiplicity is handled, and directional

702 paths are tested, so raw p-values would be meaningless. `build_hypa()`
applies Benjamini-Hochberg by default and reports **separate adjusted values
for each direction** — `p_adjusted_over` and `p_adjusted_under` — because a
path cannot be simultaneously over- and under-represented and pooling them
would be wrong.

```{r build-hypa-padjust}
as.data.frame(hp, sort_by = "p", top = 4)
```

`p_value` is the two-sided value; the directional adjusted columns are the
ones to act on. `alpha` and `p_adjust` control the threshold and method:

```{r build-hypa-strict}
hp_strict <- build_hypa(seqs, order = 2, alpha = 0.001, p_adjust = "bonferroni")
hp_strict
```

Under Bonferroni at α = 0.001 the anomaly count falls sharply — the paths
that survive are the ones worth naming in a result.

### `min_count` and order

```{r build-hypa-params}
do.call(rbind, lapply(c(5, 20), function(mc) {
  h <- build_hypa(seqs, order = 2, min_count = mc)
  scores <- as.data.frame(h)
  over <- as.data.frame(h, what = "over")
  under <- as.data.frame(h, what = "under")
  data.frame(min_count = mc, scored = nrow(scores),
             over = nrow(over), under = nrow(under))
}))
```

Higher orders test longer paths, and the table grows quickly — check that
each path still has enough observations to support a claim.

### What to do next

- **[`build_hon()`](memory-networks.html#build_hon)** — HYPA scores paths against a null;
  BuildHON+ decides which contexts deserve their own node. They answer
  different questions about the same sequences.
- **[`bootstrap_hon()`](memory-networks.html#hon_inference)** — resampling-based stability
  for extracted rules.
- **[`path_dependence()`](memory-networks.html#path_dependence)** — per-context divergence
  rather than per-path anomaly.
- **[`pathways()`](overview.html#pathways)** — pass the scored paths on to the
  simplicial family.

### References

- LaRock, T., Nanumyan, V., Scholtes, I., Casiraghi, G., Eliassi-Rad, T., &
  Schweitzer, F. (2020). HYPA: Efficient detection of path anomalies in time
  series over networks. *Proceedings of the 2020 SIAM International
  Conference on Data Mining*, 460-468.
- Benjamini, Y., & Hochberg, Y. (1995). Controlling the false discovery rate.
  *Journal of the Royal Statistical Society B*, 57(1), 289-300.


---

## build_honem(): embedding a higher-order network {#build_honem}

### The analytic question

Network embeddings put nodes in a vector space so that structurally similar
nodes end up near each other, making the network usable by anything that
takes numeric features — clustering, regression, a downstream classifier.

Applied to a *first-order* network, an embedding can only encode who follows
whom. HONEM (Saebi, Ciampaglia, Kaplan & Chawla 2020) embeds the
**higher-order** network instead, so the vector for `Specify -> Command`
reflects that context's own continuation behaviour, not just `Command`'s.

The method builds higher-order neighbourhood matrices out to `max_power`
steps, sums them into a single deviation matrix, and takes a truncated SVD.

```{r build-honem-load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
```

```{r build-honem-cograph, eval = has_cograph}
library(cograph)
```

### Embedding

`build_honem()` consumes a fitted network, not raw sequences — build the
higher-order network first and embed that.

```{r build-honem-build}
hon <- build_hon(seqs, max_order = 3, min_freq = 5)
em <- build_honem(hon, dim = 8)
em
```

233 nodes in 8 dimensions.

### Reading the coordinates

```{r build-honem-coords}
as.data.frame(em, top = 6)
```

One row per higher-order node. Note what the rows are: `Command` is a
first-order node, `Command -> Command` a second-order context, and
`Command -> Command -> Correct` a third-order one. **All three are embedded
in the same space**, which is what makes the representation useful — a
context and the state it ends in can be compared directly.

### How many dimensions do you need?

```{r build-honem-variance}
as.data.frame(em, what = "variance")
```

The first dimension carries 24.3% of the squared singular value, the second
16.0%, and the tail flattens quickly — dimensions 4 through 8 each hold
8-10%. There is no sharp elbow, which is honest information: this network has
no small set of dominant structural axes, and an 8-dimensional embedding is
retaining genuinely distributed structure rather than noise.

```{r build-honem-scree}
em20 <- build_honem(hon, dim = 20)
as.data.frame(em20, what = "variance", top = 6)
var_20 <- as.data.frame(em20, what = "variance")
cumulative <- cumsum(var_20$proportion)
data.frame(dims = c(2, 5, 10, 20), cumulative = round(cumulative[c(2, 5, 10, 20)], 3))
```

**`proportion` is relative to the dimensions retained, not to the full
spectrum** — it sums to 1 by construction, so the same axis has a different
share depending on `dim`. At `dim = 20` the first two axes hold 21.9% and the
first ten 63.2%; the 24.3% reported for `dim1` above is its share of an
8-dimensional retention.

Compare like with like when choosing `dim`, and report the value you used.

### `max_power` sets how far the neighbourhood reaches

```{r build-honem-maxpower}
do.call(rbind, lapply(c(2, 5, 10), function(mp) {
  e <- build_honem(hon, dim = 4, max_power = mp)
  v <- as.data.frame(e, what = "variance")
  data.frame(max_power = mp, dim1_share = round(v$proportion[1], 3),
             top4_share = round(sum(v$proportion), 3))
}))
```

Larger `max_power` accumulates longer-range neighbourhoods. Because every
retained dimension's share is computed within the retained set, these numbers
describe how concentrated the structure is at each reach, not how much total
variance is captured.

### Plotting

```{r build-honem-plot, fig.width = 7, fig.height = 5}
plot(em, dims = c(1, 2))
```

Read this cautiously. A 2-D projection of an 8-D embedding shows the first
two axes only — and per the variance table, that is 40% of the structure.
Proximity in the plot is suggestive, not conclusive.

### The states the embedding sits over

The embedding has one row per higher-order *node* — 233 of them. That is
far too many to draw: `splot(hon)` is an unreadable hairball, and it is the
wrong picture anyway, because what the coordinates are relative to is the
nine underlying states. Draw those.

```{r build-honem-honnet, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "First-order network of the nine action codes"}
net1 <- build_hon(human_long, action = "code", actor = "session_id",
                  time = "timestamp", max_order = 1, min_freq = 5)
splot(net1, minimum = 0.05, title = "The nine states behind the embedding")
```

Every point in the scatter above is a *context over these nine states* —
`Command`, `Command -> Command`, `Command -> Command -> Correct` are three
separate rows of the embedding and three separate paths through this graph.

### What an embedding is not

An embedding is a *representation*, not a result. Nothing above is a
statistical claim: there are no p-values, no intervals, and the coordinates
are only identified up to sign and rotation. Two runs on the same input agree
exactly, but a change in `dim` or `max_power` produces a different space, not
a rescaled one.

Use it as input to something that does make claims — and if the question is
"which actions matter", [`hon_centrality()`](memory-networks.html#hon_centrality) answers it
directly and interpretably instead.

### What to do next

- **[`build_hon()`](memory-networks.html#build_hon)** — the network this verb consumes. The
  embedding inherits every choice made there, `min_freq` especially.
- **[`hon_centrality()`](memory-networks.html#hon_centrality)** — an interpretable ranking
  rather than a latent space.
- **[`hypergraph_cluster()`](hypergraphs.html#hypergraph_laplacian)** — if the goal is
  grouping rather than coordinates.

### References

- Saebi, M., Ciampaglia, G. L., Kaplan, L. M., & Chawla, N. V. (2020). HONEM:
  Learning embedding for higher order networks. *Big Data*, 8(4), 255-269.
- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.


---

## Higher-order centrality: ranking actions a first-order network cannot rank {#hon_centrality}

### The analytic question

Which actions are central to human–AI pair-programming sessions? The
usual instrument is a centrality measure on the transition network: build
the network of what-follows-what, run PageRank, read off the ranking.

That instrument fails on these data, and it fails in an instructive way.
This tutorial shows why, and what a higher-order network recovers.
`hon_centrality()` computes centralities on the higher-order topology —
where a node is not a state but *a state together with the memory of how
it was reached* — and projects them back onto the states, following
Scholtes, Wider & Garas (2016). The implementation follows pathpy 2.2.0,
that paper's reference implementation, generalized to the variable-order
networks `build_hon()` produces.

The data are `human_long`, bundled with honets: 10,796 coded human turns
from 429 sessions, nine action codes.

```{r hon-centrality-load}
library(honets)
```

```{r hon-centrality-cograph, eval = has_cograph}
library(cograph)
```

### The first-order network has nothing to say

```{r hon-centrality-firstorder}
fo <- build_hon(human_long, action = "code", actor = "session_id",
                time = "timestamp", max_order = 1, min_freq = 5)
fo
hon_centrality(fo)
```

Every measure returns the same value for all nine actions: PageRank
exactly 1/9, closeness exactly 8, betweenness exactly 0. This is not a
numerical accident. With 10,796 turns spread over nine codes, *every one
of the 81 possible transitions occurs at least five times*, so the
first-order network is the complete digraph on nine nodes. Every node has
in-degree nine, which forces PageRank to be uniform; every pair of states
is one step apart, which forces closeness to be uniform and betweenness
to be zero, because a one-step path has no intermediate node.

The conclusion is decisive and negative: on data this dense, the
first-order network is structurally incapable of ranking the actions. No
choice of centrality measure repairs it — the information is not in the
topology.

### The higher-order network ranks them

```{r hon-centrality-higherorder}
hon <- build_hon(human_long, action = "code", actor = "session_id",
                 time = "timestamp", max_order = 3, min_freq = 5)
hon
hon_centrality(hon, type = "pagerank")
```

The same nine actions, now ranked over a 5.6-fold range: Specify holds
23.4% of the flow, Command 18.4%, down to Verify at 4.2%. The topology
that produced this has 233 nodes rather than 9 — the extra 224 are
memory contexts such as `Specify -> Command`, each carrying its own
distinct continuation distribution.

The ranking exists because the *contexts* differ, even though the states
do not. Where the first-order network sees nine interchangeable hubs, the
higher-order network sees which of them recur inside the heavily
travelled contexts. Specify and Command lead because the flow keeps
returning to them along many distinct memory paths; Verify is visited
rarely and from few contexts.

### Seeing the ranking

Two pictures, and the choice between them is the whole point of this
vignette.

First the network that cannot rank. `fo` is the complete digraph on nine
states — every pair one step apart, every node the same degree. The figure
shows why no centrality measure can separate these actions: there is no
structure to separate them by.

```{r hon-centrality-foplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "The saturated first-order network: a complete digraph on nine states"}
splot(fo, minimum = 0.05, title = "First-order: saturated, unrankable")
```

Then the ranking itself. Note what is **not** plotted here: the
higher-order network has 233 nodes and drawing it is an unreadable
hairball. `hon_centrality()` already projects onto the nine states, so the
figure is of its tidy return — nine rows, one per action.

```{r hon-centrality-centplot, fig.width = 8, fig.height = 5, fig.alt = "PageRank of the nine actions under the higher-order topology"}
pr <- hon_centrality(hon, type = "pagerank", sort_by = "pagerank")
pr$state <- factor(pr$state, levels = rev(pr$state))
ggplot2::ggplot(pr, ggplot2::aes(x = pagerank, y = state)) +
  ggplot2::geom_segment(ggplot2::aes(x = 0, xend = pagerank,
                                     yend = state),
                        colour = "#999999", linewidth = 0.5) +
  ggplot2::geom_point(size = 3.2, colour = "#0072B2") +
  ggplot2::geom_vline(xintercept = 1 / nrow(pr), linetype = "dashed",
                      colour = "#D55E00") +
  ggplot2::annotate("text", x = 1 / nrow(pr), y = 0.7, hjust = -0.1,
                    label = "first-order: all nine sit here",
                    colour = "#D55E00", size = 3.4) +
  ggplot2::labs(x = "PageRank (projected to states)", y = NULL,
                title = "Higher-order topology ranks what first-order cannot") +
  ggplot2::theme_minimal(base_size = 12)
```

The dashed line is 1/9 — where the first-order network puts every action.
The higher-order reading spreads them over a 5.6-fold range.

### Where the ranking comes from

Setting `project = FALSE` reports the centralities of the higher-order
nodes themselves — that is, of the contexts rather than the states.

```{r hon-centrality-contexts}
hon_centrality(hon, type = "pagerank", project = FALSE,
               sort_by = "pagerank", top = 8)
```

The single most central object in these sessions is not an action but a
context: `Specify -> Command`, holding 3.5% of the flow on its own — more
than three times any individual first-order node. Below it sit
`Specify -> Specify` and `Command -> Command`, the two self-repetition
contexts, and a third-order context, `Inquire -> Command -> Command`,
already appears in the top eight. Specification followed by commanding is
the load-bearing pattern of this corpus, and it is invisible to any
first-order summary, which can only report that Specify and Command are
both frequent.

### Does the projection choice drive the ranking?

A higher-order node spans several states, so its PageRank has to be
distributed among them. The default, `"scaled"`, splits it equally
(preserving total mass); `"last"` assigns it entirely to the state
actually occupied.

```{r hon-centrality-projection}
scaled <- hon_centrality(hon, type = "pagerank", projection = "scaled")
last <- hon_centrality(hon, type = "pagerank", projection = "last")
data.frame(state = scaled$state, scaled = scaled$pagerank,
           last = last$pagerank)
```

The two agree to within about one percentage point on every action, and
the ordering is identical. The ranking is therefore a property of the
higher-order topology, not of the projection rule, and can be reported
without a projection caveat.

### Betweenness and closeness need a sparser network

Both measures were degenerate on the first-order network. They stay
degenerate on the full higher-order network, and for the same reason:

```{r hon-centrality-degenerate}
hon_centrality(hon, type = c("betweenness", "closeness"))
```

The projection maps every ordered pair of states to a one-step path,
because a second-order node exists for essentially every pair. With no
intermediate nodes on any shortest path, betweenness is zero everywhere
and closeness is a constant 8. Unlike PageRank, these two measures depend
only on the *reachability* structure, and reachability here is saturated.

Raising `min_freq` keeps only the well-evidenced transitions, which
restores a topology with genuine distances:

```{r hon-centrality-sparse}
sparse <- build_hon(human_long, action = "code", actor = "session_id",
                    time = "timestamp", max_order = 3, min_freq = 80)
sparse
hon_centrality(sparse, type = c("betweenness", "closeness"))
```

On the 17-node network of transitions seen at least 80 times, the
measures separate sharply. Specify brokers 22.3 shortest paths and
Command 12.4, while Interrupt, Refine and Verify broker none at all: the
well-evidenced routes between actions run through the two directive
codes, and the remaining actions sit at the ends of those routes rather
than along them. Closeness ranks the same two highest (7.5 and 7.0),
confirming that they are also the easiest actions to reach.

Note what this costs: the sparse network describes only the strongest
transitions, so it answers a narrower question than the PageRank analysis
above. Report the threshold with the result.

### When to use which

- **`hon_centrality(hon)`** — the question is which states carry flow,
  and the sequences have memory. PageRank is the measure to trust on
  dense data; betweenness and closeness require a topology with real
  distances, so check for saturation (all-zero betweenness, constant
  closeness) before interpreting them.
- **`project = FALSE`** — the question is which *contexts* dominate.
  Often the more useful table, and the only place where a pattern like
  `Specify -> Command` can appear at all.
- **`build_hon(max_order = 1)`** — the first-order baseline. Always run
  it: the gap between it and the higher-order ranking is the evidence
  that memory matters, and if there is no gap, the simpler model should
  be reported.
- **[`bootstrap_hon()`](memory-networks.html#hon_inference)** — before reporting a
  higher-order ranking, check that the rules producing it survive
  resampling.
- **`markov_order_test()`** — the prior question of whether memory exists
  at all, as a single global test.

### References

- Scholtes, I., Wider, N., & Garas, A. (2016). Higher-order aggregate
  networks in the analysis of temporal networks: path structures and
  centralities. *The European Physical Journal B*, 89, 61.
- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5),
  e1600028.
- Scholtes, I. (2017). When is a network a network? Multi-order graphical
  model selection in pathways and temporal networks. *KDD 2017*.


---

## Are the higher-order rules real? Bootstrap and permutation inference for HONs {#hon_inference}

### The analytic question

A higher-order network promotes a context to a rule — *after
Specify → Command, the next action follows a different distribution than
after Command alone* — whenever a divergence criterion is met in the
sample (Xu, Wickramarathne & Chawla 2016). The criterion is a point
decision on one dataset. Two questions have to be answered before such
rules are reported as findings:

1. **Stability.** Would the same rules be extracted from another sample
   of the same process? `bootstrap_hon()` resamples sequences with
   replacement and reports, for every rule, the fraction of replicates
   that re-extract it (*support*) and a percentile confidence interval
   for its conditional probability.
2. **Difference.** Do two cohorts follow different rules?
   `compare_hon()` tests every pooled rule edge with a permutation null
   and aggregates a global test, with Benjamini–Hochberg control over the
   per-edge multiplicity.

Both verbs precompute per-sequence counts once and rebuild every
replicate from reweighted counts, so their cost does not grow with
sequence length re-scans, and their results are identical in serial and
parallel runs under the same seed.

The data are `human_long`, bundled with honets: 10,796 coded human turns
from 429 human–AI pair-programming sessions; each session's ordered codes
form one categorical sequence.

```{r hon-inference-load}
library(honets)
```

### Which rules survive resampling?

Rules are extracted up to order 3, requiring each transition to occur at
least 5 times, and every rule is re-examined in 500 bootstrap
resamples of the sessions.

```{r hon-inference-bootstrap}
bs <- bootstrap_hon(human_long, action = "code", actor = "session_id",
                    time = "timestamp", n_boot = 500, max_order = 3,
                    min_freq = 5, seed = 11, parallel = TRUE, n_cores = 2)
bs
summary(bs)
```

The 425 usable sessions yield 829 rule edges: 81 first-order, 451
second-order, 297 third-order. The summary separates the layers sharply.
The second-order layer is solid — mean support 0.995, and even its least
stable rule is re-extracted in 89.8% of replicates. The third-order layer
is mixed: mean support 0.85, minimum 0.35, and its intervals are wide
(mean width 0.70 against 0.20 at order 2). The conclusion is directional:
the second-order structure of these sessions can be reported as a whole;
third-order claims need the per-rule filter below.

```{r hon-inference-stable}
as.data.frame(bs, order_min = 2, min_support = 0.9,
              sort_by = "count", top = 10)
```

604 of the 748 higher-order rules pass a 0.9 support floor; the table
shows the ten most frequent. The strongest patterns are the directive
loop seen from different depths: after `Command -> Request` the next
action is Specify with probability 0.77 (CI 0.72–0.84), and the
third-order context `Specify -> Command -> Request` sharpens this to 0.97
(CI 0.91–1.00, support 0.90). History genuinely narrows the prediction —
the deeper context nearly determines the next action, and the bootstrap
says that narrowing is not a sampling artifact.

```{r hon-inference-bootplot, fig.alt = "Forest plot of rule probabilities with bootstrap confidence intervals"}
plot(bs, top = 15)
```

The forest plot makes the layering visible: first- and second-order rules
carry tight intervals, third-order rules visibly wider ones.

### Did the rules change between the early and late period?

The sessions span 2026-02-05 to 2026-03-08. Splitting the events at the
median date gives two cohorts of comparable size (a session straddling
the cutoff contributes its early turns to the first cohort and its late
turns to the second).

```{r hon-inference-compare}
early <- subset(human_long, session_date < "2026-02-21")
late  <- subset(human_long, session_date >= "2026-02-21")
cmp <- compare_hon(early, late, action = "code", actor = "session_id",
                   time = "timestamp", n_perm = 500, max_order = 3,
                   min_freq = 5, seed = 12, parallel = TRUE, n_cores = 2,
                   names = c("early", "late"))
cmp
summary(cmp)
```

The two results disagree on purpose, and reading them together is the
point of the design. The global test is unambiguous: the
pooled-count-weighted mean absolute difference is 0.0875, and only 1 of
501 permutations reaches it (p = 0.002) — the rule structure did shift
between the periods. Yet not a single one of the 828 edges is
individually significant after Benjamini–Hochberg adjustment. The shift
is *diffuse*: many small, consistent differences spread across the rule
set, none large enough to survive 828-way multiplicity on its own. The
extreme per-edge differences (absolute difference 1.0 at orders 2–3) are
contexts observed in only one period with small counts — exactly the
edges a permutation null neutralizes.

The decisive report is therefore: the interaction rules changed over
time as a whole, and no single rule should be named as the locus of that
change. Naming one anyway — picking the largest difference and reporting
it despite its adjusted p-value — is the multiple-testing error this
verb's two-level design exists to prevent.

```{r hon-inference-compareplot, fig.alt = "Per-edge probability differences between early and late cohorts"}
plot(cmp, top = 15)
```

The difference plot confirms the reading: the largest differences all
plot in the not-significant encoding, and no edge separates from the
pack.

### When to use which

- **`bootstrap_hon()`** — one cohort, and the question is whether the
  extracted rules and their probabilities are stable enough to report.
  Filter with `as.data.frame(x, order_min = 2, min_support = 0.9)` before
  citing individual higher-order rules; report interval widths, not just
  point probabilities.
- **`compare_hon()`** — two cohorts, and the question is whether the rule
  structure differs. Read the global test first; per-edge findings only
  from the BH-adjusted column. A significant global with no significant
  edges is a finding, not a failure: the difference is distributed.
- **`markov_order_test()`** — the prior, coarser question: is there any
  memory beyond order 1 at all, as a single global decision.
- **`path_dependence()`** — a descriptive per-context diagnostic of where
  memory concentrates, without inference; useful for choosing contexts to
  examine before the bootstrap.

### References

- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5),
  e1600028.
- Efron, B., & Tibshirani, R. J. (1993). *An Introduction to the
  Bootstrap*. Chapman & Hall.
- Good, P. (2005). *Permutation, Parametric and Bootstrap Tests of
  Hypotheses* (3rd ed.). Springer.
- Benjamini, Y., & Hochberg, Y. (1995). Controlling the false discovery
  rate. *JRSS B*, 57(1), 289–300.

