---
title: "Higher-order networks: three readings of one dataset"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this document still knits (without figures)
# where cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

> Each section below is one verb, worked end to end on the same data. The
> sections run in order in a single R session, so later ones may reuse
> objects the earlier ones built.

## Three higher-order readings of one dataset {#three_families}

### The analytic question

"Higher-order network" names three different objects in the literature,
and they are usually treated as three literatures. A network is
higher-order when a relation reaches past a single pair of nodes at a
single moment — but there are two distinct ways to reach past it, and
three structures that do so:

- a relation can depend on **more than the current node** — a *memory
  network*;
- a relation can bind **more than two nodes at once**, either as a set
  whose every subset is also a relation — a *simplicial complex* — or as
  a set with no such requirement — a *hypergraph*.

This tutorial puts all three on the same data and asks a question that
only a single package can ask cleanly: **do they agree?** If they did,
the choice among them would be a matter of taste. They do not, and where
they disagree is the finding.

The data are `human_long`, bundled with honets: 10,796 coded human turns
from 429 human-AI pair-programming sessions, nine action codes.

```{r three-families-load}
library(honets)
```

```{r three-families-cograph, eval = has_cograph}
library(cograph)
```

```{r three-families-simpguard, include = FALSE, eval = has_cograph}
# grDevices::chull() aborts on a degenerate (collinear or single-point)
# pathway cloud. cograph >= 2.3.5 fixes this in .smooth_blob(); shadowing the
# function here keeps the render alive on CRAN cograph (2.1.1) by falling
# back to splot(). Scope is this vignette's knit environment only.
plot_simplicial <- function(...) {
  args <- list(...)
  tryCatch(do.call(cograph::plot_simplicial, args),
           error = function(e) {
             warning("plot_simplicial fell back to splot(): ",
                     conditionMessage(e), call. = FALSE)
             splot(args[[1]])
           })
}
```

### The baseline says nothing

Before reaching for any higher-order structure, run the first-order one.
It is the model that must be beaten.

```{r three-families-firstorder}
fo <- build_hon(human_long, action = "code", actor = "session_id",
                time = "timestamp", max_order = 1, min_freq = 5)
hon_centrality(fo)
```

Every action gets PageRank exactly 1/9, closeness exactly 8, betweenness
exactly 0. That is not a numerical coincidence. With 10,796 turns over
nine codes, every one of the 81 possible transitions occurs at least five
times, so the first-order network is the *complete digraph* on nine
nodes. Every node has the same in-degree, which forces PageRank uniform;
every pair is one step apart, which forces closeness constant and
betweenness zero.

The first-order network is structurally incapable of ranking these
actions. No choice of centrality repairs that — the information is not in
the topology. Everything below is an attempt to put it back.

### Reading 1 — memory: what follows what, given how you got here

A memory network makes a node a state *plus the memory of how it was
reached*, so `Specify -> Command` is a different object from
`Inquire -> Command`, with its own continuation distribution.

```{r three-families-memory}
hon <- build_hon(human_long, action = "code", actor = "session_id",
                 time = "timestamp", max_order = 3, min_freq = 5)
hon
hon_centrality(hon, type = "pagerank", sort_by = "pagerank")
```

Nine states, but 233 nodes: the extra 224 are memory contexts. And now
the actions separate over a 5.6-fold range — Specify holds 23.4% of the
flow, Command 18.4%, down to Verify at 4.2%.

This is a claim about **flow**: where a walker that respects the
sequences' memory spends its time.

The memory reading, drawn. The object to draw is `fo`, the **first-order**
network from the opening section — nine states. The higher-order network has
233 nodes and drawing it produces a hairball, so the memory structure goes on
top of the nine-node layout as pathways rather than becoming the layout.

```{r three-families-memplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "First-order network of the nine action codes"}
splot(fo, minimum = 0.05, title = "Reading 1 - the nine states")
```

```{r three-families-mempath, eval = has_cograph, fig.width = 11, fig.height = 11, fig.alt = "Six strongest memory contexts as simplicial blobs"}
contexts_top <- as.data.frame(hon, order_min = 2, sort_by = "count", top = 6)
plot_simplicial(fo,
                pathways = contexts_top,
                max_pathways = 6,
                title = "Reading 1 - memory, as pathways")
```

### Reading 2 — simplicial: which actions co-occur as coherent groups

The same fitted memory network can be read as a *pathway complex*: every
observed path becomes a simplex, and — this is what makes it simplicial —
so does every sub-path of it. That closure is what gives the object a
geometry.

```{r three-families-simplicial}
sc <- build_simplicial(hon, type = "pathway")
sc
as.data.frame(sc, what = "f_vector")
```

Nine nodes carry 157 simplices up to dimension 3: 36 edges, 79 triangles,
33 tetrahedra. Density 61.6% — most of the combinations that *could*
appear do.

Because it has a geometry, it has a topology:

```{r three-families-homology}
betti_numbers(sc)
euler_characteristic(sc)
```

`b0 = 1` says the complex is connected — no group of actions is isolated
from the rest. `b1 = 0` says there are no one-dimensional holes: any loop
of actions can be filled in by observed pathways. `b2 = 18` says there
are **18 independent two-dimensional voids** — configurations of four
actions where every triple was observed but the four-way combination
never was. Those are the near-misses of this corpus: patterns the
sessions came right up to the edge of and never completed.

Per-action, the simplicial reading is a count of memberships:

```{r three-families-simpdegree}
simplicial_degree(sc)
```

Specify sits in 68 simplices, Verify in 41. Note the shape of the table:
`d0`/`d1` are identical for every action (each is in one vertex and eight
edges — a consequence of the saturated first-order structure), so the
whole signal is in `d2` and `d3`. The discrimination lives entirely in
the three- and four-way combinations.

The simplicial reading, drawn. This one has its own `plot()` method, and it
is the right picture: face counts by dimension, the Betti numbers, and each
action's simplicial degree. The blob diagram above showed *which* pathways;
this shows the shape they add up to.

```{r three-families-simpplot, eval = has_cograph, fig.width = 13, fig.height = 9, fig.alt = "Face counts, Betti numbers and simplicial degree of the pathway complex"}
plot(sc)
```

### Reading 3 — hypergraph: which actions belong to the same session

A hypergraph drops the closure requirement: a set of nodes can be a
relation without any of its subsets being one. That fits observed group
membership exactly. Each session is one hyperedge over the codes it
contains.

```{r three-families-hypergraph}
hg <- group_hypergraph(human_long, actor = "code", group = "session_id")
hg
measures <- hypergraph_measures(hg)
as.data.frame(measures, what = "global")
```

429 hyperedges — one per session — averaging 6.6 actions each. The
distribution is heavily weighted toward large sessions: 80 of the 429
sessions use all nine codes.

```{r three-families-hgcent}
hypergraph_centrality(hg, type = "pagerank", sort_by = "pagerank")
```

This is a claim about **co-membership**: how much of a random walk that
hops action → session → action lands on each action.

The hypergraph reading, drawn. A hypergraph has no direct network picture,
so it goes through `clique_expansion()` — and the caption has to say so,
because the projection is a lossy one.

```{r three-families-hgplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "Session co-membership hypergraph, projected to a pairwise network"}
net <- clique_expansion(hg)
splot(net, minimum = 0.05,
      title = "Reading 3 - hypergraph, projected")
```

Three pictures of the same 10,796 turns. The first is directed and carries
memory; the second is undirected and carries closure; the third is
undirected, complete, and carries only co-membership weight. Nothing about
the data changed — only the question.

### Do the three agree?

Three tidy tables, joined in two steps. Nothing here reaches into a result
object.

```{r three-families-compare}
pr_memory <- hon_centrality(hon, type = "pagerank")
deg_simplicial <- simplicial_degree(sc)
pr_hypergraph <- hypergraph_centrality(hg, type = "pagerank")
cmp <- merge(pr_memory, deg_simplicial, by.x = "state", by.y = "node")
cmp <- merge(cmp, pr_hypergraph, by.x = "state", by.y = "node",
             suffixes = c("_memory", "_hypergraph"))
cmp
```

```{r three-families-spearman}
cmp_ranks <- subset(cmp, select = c(pagerank_memory, total, pagerank_hypergraph))
cor(cmp_ranks, method = "spearman")
```

The memory and simplicial readings agree strongly (Spearman 0.87) — both
are built from the same fitted pathways, so this is a consistency check
rather than independent evidence. The interesting number is **0.55**:
memory flow and hypergraph co-membership rank these nine actions
*differently*.

Two actions carry that disagreement, in opposite directions.

**Interrupt** is 8th of 9 by memory flow (0.064) and 3rd by
co-membership (0.135). It appears in 393 of the 429 sessions — third
highest hyperdegree — yet the transitions into and out of it are not on
heavily travelled paths. Interrupt is *ubiquitous but peripheral*: almost
every session contains one, and almost none is organised around it.

**Frustrate** runs the other way: 4th by memory flow (0.104), 7th by
co-membership (0.093). It shows up in fewer sessions (252) but when it
does, it sits on paths the walk keeps returning to. Frustration is
*localised but consequential*.

Neither reading is wrong. They answer different questions, and an
analysis that reports only one of them will describe Interrupt as either
a central action or a marginal one, with no way to see that both are
true.

### Crossing back

The families are entry points into one another. A hypergraph projects
back to an ordinary pairwise network, which any first-order tool accepts:

```{r three-families-expand}
net <- clique_expansion(hg)
summary(net)
```

Nine nodes, 36 edges — the complete undirected graph, weighted by how
often each pair shared a session. Compare that to the first-order network
this tutorial opened with: same saturation, but now the *weights* carry
the co-membership information, so the projection is not the degenerate
object the transition network was.

### Choosing a reading

- **Memory network** (`build_hon()`) — the question is about *order and
  history*: what follows what, and does the past matter. Check
  `markov_order_test()` first: if memory is not there, the simpler model
  should be reported.
- **Simplicial complex** (`build_simplicial()`) — the question is about
  *coherent groups and their geometry*: which combinations hold together,
  and what is conspicuously missing. `betti_numbers()` finds the absences;
  a first-order summary can only report presences.
- **Hypergraph** (`group_hypergraph()`, `window_hypergraph()`) — the
  question is about *co-membership*: which things belong to the same
  occasion, with no assumption that sub-groups are meaningful in
  themselves.
- **All three** — when an action's role is contested. The gap between the
  readings is the finding, as it is for Interrupt here.

Two cautions. The simplicial reading above was built *from* the memory
network, so its agreement with the memory ranking is partly structural;
build it from an independently estimated network if you need the two to
be independent evidence. And every number here is a point estimate — use
[`bootstrap_hon()`](memory-networks.html#hon_inference) before reporting a memory-based
ranking as a finding.

### See also

- [Higher-order centrality](memory-networks.html#hon_centrality) — the memory family in
  depth, including which *contexts* carry the flow.
- [Bootstrap and permutation inference](memory-networks.html#hon_inference) — how stable
  a rule is under resampling.
- [Windowed sequence hyperedges](hypergraphs.html#window_hypergraph) — the other
  route from sequences into the hypergraph family.
- [Hypergraph PageRank](hypergraphs.html#hypergraph_centrality) — edge-dependent vertex
  weights, and when a hypergraph collapses to a pairwise graph.

### References

- Battiston, F., Cencetti, G., Iacopini, I., Latora, V., Lucas, M.,
  Patania, A., Young, J.-G., & Petri, G. (2020). Networks beyond pairwise
  interactions: Structure and dynamics. *Physics Reports*, 874, 1-92.
- Bianconi, G. (2021). *Higher-Order Networks*. Cambridge University
  Press.
- Scholtes, I., Wider, N., & Garas, A. (2016). Higher-order aggregate
  networks in the analysis of temporal networks. *The European Physical
  Journal B*, 89, 61.
- Atkin, R. H. (1974). *Mathematical Structure in Human Affairs*.
  Heinemann.
- Chitra, U., & Raphael, B. J. (2019). Random walks on hypergraphs with
  edge-dependent vertex weights. *ICML 2019*, PMLR 97, 1172-1181.


---

## pathways(): the bridge between the three families {#pathways}

### The analytic question

The three structure families in honets are entry points into one another, not
islands. `pathways()` is the generic that carries a fitted **memory network**
across to the other two: it extracts the paths the model considers real and
returns them as plain strings, which
[`build_simplicial(type = "pathway")`](simplicial-complexes.html#build_simplicial) turns into a
complex.

It is a small verb doing one job, and it is the reason a HON can be handed to
a topological analysis without any manual reshaping.

```{r pathways-load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
```

```{r pathways-cograph, eval = has_cograph}
library(cograph)
```

```{r pathways-simpguard, include = FALSE, eval = has_cograph}
# grDevices::chull() aborts on a degenerate (collinear or single-point)
# pathway cloud. cograph >= 2.3.5 fixes this in .smooth_blob(); shadowing the
# function here keeps the render alive on CRAN cograph (2.1.1) by falling
# back to splot(). Scope is this vignette's knit environment only.
plot_simplicial <- function(...) {
  args <- list(...)
  tryCatch(do.call(cograph::plot_simplicial, args),
           error = function(e) {
             warning("plot_simplicial fell back to splot(): ",
                     conditionMessage(e), call. = FALSE)
             splot(args[[1]])
           })
}
```

### Three methods, one generic

`pathways()` dispatches on `net_hon`, `net_hypa` and `net_mogen`. Each
returns the paths that model regards as meaningful — which is a *different
set* in each case, because the three models mean different things by
"meaningful".

```{r pathways-hon}
hon <- build_hon(seqs, max_order = 2, min_freq = 50)
p_hon <- pathways(hon)
length(p_hon)
pathways(hon, top = 6)
```

From a HON: the extracted rules — contexts that earned their own node. 25 of
them at `min_freq = 50`.

```{r pathways-hypa}
hp <- build_hypa(seqs, order = 2)
p_hypa <- pathways(hp)
length(p_hypa)
pathways(hp, top = 4)
```

From HYPA: the **anomalous** paths — those significantly over- or
under-represented against the degree-preserving null. 185 of them. Note the
first entry, `Interrupt Interrupt -> Interrupt`: the same repeated-
interruption pattern that [`path_dependence()`](memory-networks.html#path_dependence) flags
as the highest-divergence context.

```{r pathways-mogen}
mo <- build_mogen(seqs, max_order = 2)
p_mogen <- pathways(mo)
length(p_mogen)
pathways(mo, top = 4)
```

From MOGen: every path in the fitted layer — 702, the complete second-order
transition set, because a multi-order model does not prune.

### The format

Each string is `"<context states, space separated> -> <next state>"`. The
context uses spaces so that the arrow separates history from continuation
unambiguously.

```{r pathways-format}
p_hon[1]
```

Read as: the context `Specify Command` continuing to `Specify`.

### Crossing to the simplicial family

```{r pathways-simplicial}
sc <- build_simplicial(hon, type = "pathway")
sc
```

`build_simplicial()` calls `pathways()` internally — you rarely need to call
it yourself. Each path becomes a simplex, and because a simplicial complex is
closed under subsets, so does every sub-path. The result has a geometry, and
therefore a topology: `b0 = 2` says the well-evidenced pathways fall into
**two disconnected groups** of actions, and `b1 = 1` says there is one
one-dimensional hole — a cycle of actions that is never closed by an observed
pathway.

Neither statement is available from the transition network the paths came
from.

### Seeing them

Whichever model produced them, pathways draw the same way — as blobs over
the **first-order** nine-node layout. The higher-order object supplies the
pathways; it is never the thing drawn. Compare the sources by eye: HON gives
the contexts that earned a node, HYPA the ones that deviate from the null,
MOGen every fitted path.

```{r pathways-pwplot, eval = has_cograph, fig.width = 11, fig.height = 11, fig.alt = "Pathways from the HON, drawn as simplicial blobs"}
net1 <- build_hon(seqs, max_order = 1)
contexts_top <- as.data.frame(hon, order_min = 2, sort_by = "count", top = 6)
plot_simplicial(net1,
                pathways = contexts_top,
                max_pathways = 6, title = "Pathways from build_hon()")
```

```{r pathways-pwhypa, eval = has_cograph, fig.width = 14, fig.height = 7, fig.alt = "Anomalous pathways from HYPA, one panel each"}
plot_simplicial(hp, anomaly = "over", max_pathways = 6,
                dismantled = TRUE, ncol = 3)
```

### The choice of source model matters

Because the three models select different paths, they build different
complexes:

```{r pathways-compare}
do.call(rbind, lapply(
  list(HON = hon, HYPA = hp, MOGen = mo),
  function(m) {
    p <- pathways(m)
    s <- build_simplicial(m, type = "pathway")
    f_vector <- as.data.frame(s, what = "f_vector")
    betti <- betti_numbers(s)
    data.frame(paths = length(p), simplices = sum(f_vector$count),
               dimension = max(f_vector$dim), b0 = betti[["b0"]])
  }))
```

MOGen's 702 unpruned paths produce a dense, connected complex; HON's 25
well-evidenced rules produce a sparse, disconnected one. **The topology you
report is a property of the model you chose**, not of the raw sequences — say
which model, and say why.

### Calling it directly

You would call `pathways()` yourself to inspect what a model considers real,
or to feed the paths somewhere else:

```{r pathways-direct}
# how long are the extracted paths?
table(lengths(strsplit(p_hon, " ", fixed = TRUE)))
```

### What to do next

- **[`build_simplicial()`](simplicial-complexes.html#build_simplicial)** — the main consumer, and
  where the topology gets computed.
- **[`build_hon()`](memory-networks.html#build_hon)**, **[`build_hypa()`](memory-networks.html#build_hypa)**,
  **[`build_mogen()`](memory-networks.html#build_mogen)** — the three producers.
- **[`window_hypergraph()`](hypergraphs.html#window_hypergraph)** — the other route out
  of sequence data, into the hypergraph family.

### References

- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.
- Battiston, F., et al. (2020). Networks beyond pairwise interactions.
  *Physics Reports*, 874, 1-92.

