---
title: "Simplicial complexes: the shape of an interaction space"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
```

> Each section below is one verb, worked end to end on the same data. The
> sections run in order in a single R session, so later ones may reuse
> objects the earlier ones built.

## build_simplicial(): from a network to a shape {#build_simplicial}

### The analytic question

A graph records which *pairs* of things are related. A simplicial complex
records which *groups* are related — and, crucially, it is **closed under
subsets**: if `{A, B, C}` is a relation, so are `{A, B}`, `{A, C}` and
`{B, C}`.

That closure is not bookkeeping. It is what gives the object a geometry, and
therefore a topology: connected pieces, loops, voids. Those are properties no
pairwise summary can express.

`build_simplicial()` builds one three ways, selected with `type =`.

```{r build-simplicial-load}
library(honets)
set.seed(1)
m <- matrix(runif(100), 10, 10)
m <- (m + t(m)) / 2
diag(m) <- 0
dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
```

### `type = "clique"` — from a network

Every clique in the thresholded graph becomes a simplex. A triangle of
mutually-connected nodes becomes a filled 2-simplex.

```{r build-simplicial-clique}
sc <- build_simplicial(m, type = "clique", threshold = 0.5)
sc
```

Ten nodes carry 51 simplices up to dimension 3: 24 edges, 15 triangles and 2
tetrahedra.

```{r build-simplicial-table}
as.data.frame(sc, top = 5)
as.data.frame(sc, what = "f_vector")
```

The **f-vector** counts faces by dimension and is the complex's basic
fingerprint. `dim = 0` are vertices, `1` edges, `2` triangles, `3`
tetrahedra.

```{r build-simplicial-bydim}
as.data.frame(sc, dim = 3, top = 3)
```

### Seeing the complex

`plot()` draws the face counts by dimension and the complex itself.

```{r build-simplicial-scplot, fig.width = 13, fig.height = 6, fig.alt = "Face counts by dimension for the clique complex"}
plot(sc)
```

### `threshold` decides the shape

This is the consequential parameter, and its effect is not monotone in any
simple way.

```{r build-simplicial-sweep}
do.call(rbind, lapply(c(0.3, 0.4, 0.5, 0.6, 0.7), function(t) {
  s <- build_simplicial(m, type = "clique", threshold = t)
  fv <- as.data.frame(s, what = "f_vector")
  b <- betti_numbers(s)
  e <- euler_characteristic(s)
  data.frame(threshold = t, simplices = sum(fv$count), dimension = max(fv$dim),
             euler = e,
             b0 = b[["b0"]], b1 = b[["b1"]])
}))
```

Watch `b1`, the number of one-dimensional holes. It is **0 at threshold 0.3,
rises to 4 at 0.6, then collapses to 0 at 0.7**. At a low threshold the graph
is dense enough that every loop is filled in; at a high one it is too sparse
to form loops at all. The interesting structure lives in between.

That instability is the whole argument for
[`persistent_homology()`](simplicial-complexes.html#persistent_homology), which sweeps the
threshold instead of committing to one. **Never report a Betti number from a
single arbitrary threshold** without showing the sweep.

### `type = "vr"` — from a distance matrix

The Vietoris-Rips complex connects points closer than a scale, then fills in
every resulting clique. It is the standard construction for point-cloud data.

```{r build-simplicial-vr}
set.seed(2)
pts <- matrix(rnorm(40), 20, 2)
d <- as.matrix(dist(pts))
sv <- build_simplicial(d, type = "vr", threshold = 1.5, max_dim = 3L)
sv
```

#### `max_dim` is not optional here

A VR complex grows combinatorially. On this 20-point cloud:

```{r build-simplicial-vrcost}
do.call(rbind, lapply(c(2L, 3L, 4L), function(md) {
  t0 <- proc.time()[["elapsed"]]
  s <- build_simplicial(d, type = "vr", threshold = 1.5, max_dim = md)
  fv <- as.data.frame(s, what = "f_vector")
  data.frame(max_dim = md, simplices = sum(fv$count),
             seconds = round(proc.time()[["elapsed"]] - t0, 2))
}))
```

Roughly a 3-4× increase per dimension. At the **default `max_dim = 10L`**
this same 20-point cloud produces **784,625 simplices** and takes minutes —
so for VR complexes, set `max_dim` deliberately. Dimensions above 3 rarely
carry interpretable structure, and `persistent_homology()` defaults to
`max_dim = 3L` for exactly this reason.

### `type = "pathway"` — from a memory network

The third route takes a fitted model from the memory family and turns its
extracted paths into simplices. See
[`pathways()`](overview.html#pathways) for what each model contributes.

```{r build-simplicial-pathway}
seqs <- split(human_long$code, human_long$session_id)
hon <- build_hon(seqs, max_order = 2, min_freq = 50)
build_simplicial(hon, type = "pathway")
```

Here `b0 = 2` says the well-evidenced pathways split into two disconnected
groups of actions — a statement about the corpus that the transition network
cannot make.

### Checking the construction

`verify_simplicial()` re-derives the f-vector independently, as a guard
against a mis-specified threshold or a malformed matrix.

```{r build-simplicial-verify}
verify_simplicial(m, threshold = 0.5)
```

It also confirms the **Euler-Poincaré identity** — the alternating sum of the
f-vector equals the alternating sum of the Betti numbers:

```{r build-simplicial-euler}
fv <- as.data.frame(sc, what = "f_vector")
sum(fv$count * (-1)^fv$dim)
euler_characteristic(sc)
```

Both give −1. If they ever disagree, the homology computation is wrong.

### What to do next

- **[`betti_numbers()`](simplicial-complexes.html#betti_numbers)** — what the holes mean, and how
  to read them.
- **[`persistent_homology()`](simplicial-complexes.html#persistent_homology)** — the answer to the
  threshold problem above.
- **[`q_analysis()`](simplicial-complexes.html#q_analysis)** — connectivity at each level of
  overlap, and per-node higher-order degree.
- **[`build_hypergraph()`](hypergraphs.html#build_hypergraph)** — the same cliques
  without the subset-closure requirement.

### References

- Edelsbrunner, H., & Harer, J. (2010). *Computational Topology: An
  Introduction*. American Mathematical Society.
- Zomorodian, A., & Carlsson, G. (2005). Computing persistent homology.
  *Discrete & Computational Geometry*, 33(2), 249-274.


---

## betti_numbers(): counting holes {#betti_numbers}

### The analytic question

Once a set of relations has been made into a
[simplicial complex](simplicial-complexes.html#build_simplicial), it has a shape — and a shape can
be counted. Betti numbers count the holes, one number per dimension:

- **b0** — connected components. How many separate pieces?
- **b1** — one-dimensional holes. Loops that are *not* filled in.
- **b2** — two-dimensional voids. Cavities enclosed by triangles.

The useful ones are `b1` and `b2`, because they detect **absence**. A loop in
`b1` is a cycle of relations where every consecutive pair is present but the
interior is not — a near-miss. Frequency-based summaries cannot see those;
they only report what happened.

```{r betti-numbers-load}
library(honets)
set.seed(1)
m <- matrix(runif(100), 10, 10)
m <- (m + t(m)) / 2
diag(m) <- 0
dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
sc <- build_simplicial(m, type = "clique", threshold = 0.5)
```

### Counting

```{r betti-numbers-betti}
betti_numbers(sc)
```

One connected component, two independent loops, no voids. The two loops are
the finding: pairs of nodes are connected all the way round a cycle, but the
cycle is never completed into filled triangles.

### The Euler characteristic

`euler_characteristic()` compresses the whole shape into one integer — the
alternating sum of face counts.

```{r betti-numbers-euler}
euler_characteristic(sc)
```

It can be computed two entirely different ways, and they must agree. That is
the **Euler-Poincaré identity**, and it is the best available check that the
homology computation is correct.

```{r betti-numbers-identity}
fv <- as.data.frame(sc, what = "f_vector")
fv

# from the faces: f0 - f1 + f2 - f3
sum(fv$count * (-1)^fv$dim)

# from the holes: b0 - b1 + b2 - b3
b <- betti_numbers(sc)
sum(b * (-1)^(seq_along(b) - 1))
```

All three give −1. If they diverge, something is wrong — this identity is
worth asserting in any pipeline that computes homology.

### Betti numbers are threshold-dependent

This is the caveat that matters most.

```{r betti-numbers-sweep}
do.call(rbind, lapply(seq(0.3, 0.75, by = 0.05), function(t) {
  s <- build_simplicial(m, type = "clique", threshold = t)
  fv <- as.data.frame(s, what = "f_vector")
  b <- betti_numbers(s)
  e <- euler_characteristic(s)
  data.frame(threshold = t, simplices = sum(fv$count),
             b0 = b[["b0"]], b1 = b[["b1"]],
             euler = e)
}))
```

`b1` is **0, then 2, then 4, then 0** across a modest range of thresholds. A
paper reporting "the complex has two loops" without the threshold is
reporting an artefact of an arbitrary choice.

Two honest ways to handle it:

1. **Show the sweep**, as above, and describe how the count behaves.
2. Use **[`persistent_homology()`](simplicial-complexes.html#persistent_homology)**, which sweeps
   the threshold automatically and reports how *long* each feature survives.
   A loop that persists across many scales is real; one that appears at a
   single threshold is noise.

### Seeing the sweep

The table above is a curve. Plotting it is the honest way to report a
threshold-dependent invariant: it shows the whole range rather than the one
threshold that happened to be chosen.

```{r betti-numbers-sweepplot, fig.width = 8, fig.height = 5, fig.alt = "Betti numbers b0 and b1 across the threshold sweep"}
long <- do.call(rbind, lapply(seq(0.3, 0.75, by = 0.05), function(t) {
  s <- build_simplicial(m, type = "clique", threshold = t)
  b <- betti_numbers(s)
  data.frame(threshold = t, dimension = c("b0", "b1"),
             betti = c(b[["b0"]], b[["b1"]]))
}))
ggplot2::ggplot(long, ggplot2::aes(x = threshold, y = betti,
                                   colour = dimension, shape = dimension)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_point(size = 2.2) +
  ggplot2::scale_colour_manual(values = c(b0 = "#0072B2", b1 = "#D55E00")) +
  ggplot2::labs(x = "threshold", y = "Betti number", colour = NULL,
                shape = NULL,
                title = "Betti numbers across the threshold sweep") +
  ggplot2::theme_minimal(base_size = 12)
```

### On real data

```{r betti-numbers-real}
seqs <- split(human_long$code, human_long$session_id)
hon <- build_hon(seqs, max_order = 3, min_freq = 5)
sc_hon <- build_simplicial(hon, type = "pathway")
sc_hon
betti_numbers(sc_hon)
```

`b0 = 1`: the pathway complex is connected — no group of actions is isolated.
`b1 = 0`: every loop of actions is filled in by some observed pathway.
`b2 = 18`: there are **18 independent two-dimensional voids** — configurations
of four actions where every triple was observed but the four-way combination
never was.

Those 18 are the corpus's near-misses: patterns the sessions came right up to
the edge of and never completed. That is a claim about what is *absent*,
which is why it needed a topological method to find.

### What to do next

- **[`persistent_homology()`](simplicial-complexes.html#persistent_homology)** — the threshold
  problem, solved properly.
- **[`q_analysis()`](simplicial-complexes.html#q_analysis)** — a different notion of
  connectivity: how strongly do the groups overlap?
- **[`build_simplicial()`](simplicial-complexes.html#build_simplicial)** — the constructions these
  numbers describe.

### References

- Edelsbrunner, H., & Harer, J. (2010). *Computational Topology: An
  Introduction*. American Mathematical Society.
- Ghrist, R. (2008). Barcodes: The persistent topology of data. *Bulletin of
  the American Mathematical Society*, 45(1), 61-75.


---

## persistent_homology(): which features are real {#persistent_homology}

### The analytic question

[`betti_numbers()`](simplicial-complexes.html#betti_numbers) counts holes at **one** threshold,
and the count is unstable: change the threshold slightly and loops appear and
vanish. Any single-threshold result is partly an artefact of the choice.

`persistent_homology()` removes the choice. It sweeps the threshold across
the whole range, tracks every feature from the scale where it is **born** to
the scale where it **dies**, and reports how long each survived. Features
that persist across many scales are structural; features that flicker in and
out at one scale are noise.

This turns "how many holes?" — which has no threshold-free answer — into
"which holes are robust?", which does.

```{r persistent-homology-load}
library(honets)
set.seed(1)
m <- matrix(runif(100), 10, 10)
m <- (m + t(m)) / 2
diag(m) <- 0
dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
```

### Running it

```{r persistent-homology-run}
ph <- persistent_homology(m, n_steps = 12, max_dim = 2)
ph
```

### The persistence diagram

One row per feature: the dimension it lives in, where it was born, where it
died, and the difference.

```{r persistent-homology-diagram}
as.data.frame(ph, sort_by = "persistence", top = 8)
```

**`persistence` is the column that matters.** The top feature — a
0-dimensional class with persistence 0.93 — is the connected component that
survives the entire filtration. Everything below it is short-lived by
comparison.

Note that `birth` exceeds `death` here. The filtration runs from a high
weight threshold down to a low one, so features are born at high weights and
die as the complex fills in. Persistence is the length of the interval either
way.

### Reading one dimension at a time

```{r persistent-homology-dim1}
as.data.frame(ph, dimension = 1, sort_by = "persistence", top = 5)
```

```{r persistent-homology-dim0}
as.data.frame(ph, dimension = 0, sort_by = "persistence", top = 5)
```

Dimension 0 is the merge history of connected components — each row is two
pieces joining. Dimension 1 is loops, dimension 2 voids.

### Betti curves

The same information viewed as counts against threshold, rather than as
intervals.

```{r persistent-homology-betti}
as.data.frame(ph, what = "betti", top = 8)
```

Read the `b0` column down: **9 components at the highest threshold, then 8,
7, and then 1** — the graph rapidly coalesces into a single piece as the
threshold falls. That transition point is where the network "connects", and
it is visible here without having to pick a threshold in advance.

### Plotting

```{r persistent-homology-plot, fig.width = 8, fig.height = 4}
if (requireNamespace("gridExtra", quietly = TRUE)) plot(ph)
```

The diagram plots death against birth; points far from the diagonal are the
persistent features. The Betti curves show the counts across the sweep.

### Choosing `n_steps` and `max_dim`

```{r persistent-homology-params}
do.call(rbind, lapply(c(6L, 12L, 24L), function(ns) {
  p <- persistent_homology(m, n_steps = ns, max_dim = 2)
  d <- as.data.frame(p)
  data.frame(n_steps = ns, features = nrow(d),
             max_persistence = round(max(d$persistence), 4))
}))
```

`n_steps` controls the resolution of the sweep — more steps locate births and
deaths more precisely, at linear cost. The **longest-lived feature is stable
across resolutions**, which is the reassurance you want: the headline finding
does not depend on the grid.

`max_dim` is the expensive one. It defaults to `3L` here (rather than the
`10L` that [`build_simplicial()`](simplicial-complexes.html#build_simplicial) uses) precisely
because complexes grow combinatorially with dimension. Raise it only if you
have a reason to expect structure above dimension 3.

### Essential classes

```{r persistent-homology-essential}
d <- as.data.frame(ph)
sum(!is.finite(d$death))
```

Features that never die are reported with `death = Inf`. There are none in
this filtration because it runs to completion. When comparing diagrams with
[`bottleneck_distance()`](simplicial-complexes.html#persistence_landscape), essential classes need
explicit handling — the reference implementations differ on whether to
include them.

### Validation

The Z/2 boundary-matrix reduction behind this verb is checked against
**ripser** (via `TDAstats`) in the package's equivalence suite: 30 random
point clouds, maximum discrepancy 4.44e-16 — machine precision.

### What to do next

- **[`persistence_landscape()`](simplicial-complexes.html#persistence_landscape)** — turn a
  diagram into something you can average, plot and compare statistically.
- **[`betti_numbers()`](simplicial-complexes.html#betti_numbers)** — the single-threshold version,
  and why it needs a sweep.
- **[`build_simplicial()`](simplicial-complexes.html#build_simplicial)** — the complexes this
  filters over.

### References

- Edelsbrunner, H., Letscher, D., & Zomorodian, A. (2002). Topological
  persistence and simplification. *Discrete & Computational Geometry*, 28,
  511-533.
- Zomorodian, A., & Carlsson, G. (2005). Computing persistent homology.
  *Discrete & Computational Geometry*, 33(2), 249-274.
- Ghrist, R. (2008). Barcodes: The persistent topology of data. *Bulletin of
  the American Mathematical Society*, 45(1), 61-75.


---

## Comparing persistence diagrams {#persistence_landscape}

### The analytic question

A [persistence diagram](simplicial-complexes.html#persistent_homology) is a *multiset of points* —
not a vector, not a function. You cannot average two of them, feed one to a
regression, or compute a standard deviation across a sample. That is
inconvenient the moment you want to compare two datasets rather than describe
one.

Two verbs solve it in different ways:

- **`bottleneck_distance()`** gives a *metric*: how far apart are these two
  diagrams? Good for a direct comparison, and it is the classical stability
  distance for persistence.
- **`persistence_landscape()`** gives a *function*: it re-encodes the diagram
  as a sequence of piecewise-linear functions living in a vector space, where
  averaging and statistics are ordinary operations.

```{r persistence-landscape-load}
library(honets)
mk <- function(seed) {
  set.seed(seed)
  m <- matrix(runif(100), 10, 10)
  m <- (m + t(m)) / 2
  diag(m) <- 0
  dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
  persistent_homology(m, n_steps = 12, max_dim = 2)
}
ph  <- mk(1)
ph2 <- mk(9)
```

### Bottleneck distance

```{r persistence-landscape-bottleneck}
bottleneck_distance(ph, ph2)
```

One value per dimension: components, loops, voids. These two random networks
differ most in dimension 2 (0.129) and least in dimension 1 (0.094).

Restrict to a single dimension with `dimension =`:

```{r persistence-landscape-dim}
bottleneck_distance(ph, ph2, dimension = 1)
```

#### It is a metric, and you should check that

```{r persistence-landscape-metric}
# identity of indiscernibles: d(x, x) = 0
bottleneck_distance(ph, ph)

# symmetry: d(x, y) = d(y, x)
d_12 <- bottleneck_distance(ph, ph2)
d_21 <- bottleneck_distance(ph2, ph)
identical(d_12, d_21)
```

Both hold exactly. The bottleneck distance is the smallest ε such that the
two diagrams can be matched within ε, allowing unmatched points to pair with
the diagonal — which is what makes it robust to the short-lived noise
features that clutter every diagram.

### Persistence landscapes

A landscape converts the diagram into `k_max` functions. The first, `k = 1`,
traces the most persistent feature at each scale; `k = 2` the second-most,
and so on.

```{r persistence-landscape-landscape}
pl <- persistence_landscape(ph, k_max = 3, dimension = 1)
pl
```

The **sup norms** summarise each level in one number: `k1 = 0.0987`,
`k2 = 0.0936`, `k3 = 0.0609`. The first two are close, meaning this
filtration has two comparably prominent loops rather than one dominant one —
consistent with `b1 = 2` from [`betti_numbers()`](simplicial-complexes.html#betti_numbers).

```{r persistence-landscape-ltable}
as.data.frame(pl, top = 4)
```

One row per (level, grid point). Restrict to one level with `k =`:

```{r persistence-landscape-k1}
as.data.frame(pl, k = 1, top = 3)
pl_1 <- as.data.frame(pl, k = 1)
nrow(pl_1)
```

200 grid points per level, which is `t_grid`'s default resolution.

### Plotting

```{r persistence-landscape-plot, fig.width = 7, fig.height = 4.5}
plot(pl)
```

### Why a landscape rather than a diagram

Because it lives in a vector space, a landscape supports operations a diagram
does not:

```{r persistence-landscape-average}
land_1 <- persistence_landscape(ph,  k_max = 2, dimension = 1)
land_2 <- persistence_landscape(ph2, k_max = 2, dimension = 1)
l1 <- as.data.frame(land_1, k = 1)
l2 <- as.data.frame(land_2, k = 1)

# the two landscapes are evaluated on different grids by default
range(l1$t)
range(l2$t)
```

Note the grids differ — each landscape is evaluated over its own diagram's
range. **To average or subtract landscapes, put them on a common grid** with
`t_grid`:

```{r persistence-landscape-commongrid}
g <- seq(0.45, 0.75, length.out = 100)
land_a <- persistence_landscape(ph,  k_max = 2, dimension = 1, t_grid = g)
land_b <- persistence_landscape(ph2, k_max = 2, dimension = 1, t_grid = g)
a <- as.data.frame(land_a, k = 1)
b <- as.data.frame(land_b, k = 1)

mean_landscape <- data.frame(t = g, value = (a$value + b$value) / 2)

# L2 distance between the two, on the shared grid
sqrt(sum((a$value - b$value)^2) * diff(g)[1])
```

A mean landscape is a landscape, so it plots like one:

```{r persistence-landscape-meanplot, fig.width = 8, fig.height = 4.5, fig.alt = "Two persistence landscapes and their mean on a shared grid"}
both <- rbind(
  data.frame(t = g, value = a$value,           landscape = "A"),
  data.frame(t = g, value = b$value,           landscape = "B"),
  data.frame(t = g, value = mean_landscape$value, landscape = "mean")
)
ggplot2::ggplot(both, ggplot2::aes(x = t, y = value,
                                   colour = landscape, linetype = landscape)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::scale_colour_manual(
    values = c(A = "#0072B2", B = "#D55E00", mean = "#009E73")) +
  ggplot2::labs(x = "filtration parameter t", y = expression(lambda^(1)),
                colour = NULL, linetype = NULL,
                title = "Two landscapes and their pointwise mean") +
  ggplot2::theme_minimal(base_size = 12)
```

That mean is a legitimate object — the average of two persistence landscapes
is itself a persistence landscape. The average of two persistence *diagrams*
is not well defined at all. This is the entire reason landscapes exist.

### Choosing between them

- **`bottleneck_distance()`** — comparing two datasets, or checking stability
  under perturbation. Returns one interpretable number per dimension. Slower
  on large diagrams, since it solves a matching problem.
- **`persistence_landscape()`** — many datasets, or any analysis needing
  means, variances or a regression. Fix `t_grid` across the sample, or the
  results are not comparable.

### What to do next

- **[`persistent_homology()`](simplicial-complexes.html#persistent_homology)** — where the
  diagrams come from.
- **[`betti_numbers()`](simplicial-complexes.html#betti_numbers)** — the single-threshold summary
  these methods generalise.

### References

- Bubenik, P. (2015). Statistical topological data analysis using persistence
  landscapes. *Journal of Machine Learning Research*, 16, 77-102.
- Cohen-Steiner, D., Edelsbrunner, H., & Harer, J. (2007). Stability of
  persistence diagrams. *Discrete & Computational Geometry*, 37(1), 103-120.
- Kerber, M., Morozov, D., & Nigmetov, A. (2017). Geometry helps to compare
  persistence diagrams. *ACM Journal of Experimental Algorithmics*, 22.


---

## q_analysis() and simplicial_degree(): connectivity by overlap {#q_analysis}

### The analytic question

In a graph, two nodes are connected or they are not. In a
[simplicial complex](simplicial-complexes.html#build_simplicial) there are *degrees* of
connection: two groups might share one member, or three, or six. Q-analysis
(Atkin 1974) takes that seriously.

Two simplices are **q-connected** when they share at least `q + 1` vertices,
possibly through a chain of intermediates. Sweep `q` from high to low and you
get a nested picture: at high `q` only heavily-overlapping groups are linked,
and the structure fragments; as `q` falls, the fragments merge.

Where the merging happens tells you how the system is organised — something
neither a degree distribution nor a Betti number reports.

```{r q-analysis-load}
library(honets)
set.seed(1)
m <- matrix(runif(100), 10, 10)
m <- (m + t(m)) / 2
diag(m) <- 0
dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
sc <- build_simplicial(m, type = "clique", threshold = 0.5)
```

### Running it

```{r q-analysis-run}
qa <- q_analysis(sc)
qa
```

### The q-vector

```{r q-analysis-qvec}
as.data.frame(qa)
```

Read it from the top down. At `q = 3` — groups sharing four or more members —
the complex is in **11 separate pieces**. At `q = 2` it is still 10. At
`q = 1` it drops to 2, and at `q = 0` everything is a single component.

The collapse happens between `q = 2` and `q = 1`: this structure is held
together by *pairwise* overlap, not by deep shared membership. Groups here
touch each other lightly and rarely share three or more members. A system
organised around tightly-interlocking teams would show the opposite profile —
few components even at high `q`.

### Seeing the q-vector

```{r q-analysis-qaplot, fig.width = 13, fig.height = 6, fig.alt = "Q-vector: connected components at each q-level"}
plot(qa)
```

### Per-node structure

```{r q-analysis-nodes}
as.data.frame(qa, what = "nodes")
```

`max_q` is the highest level at which each node still participates in a
connected group. Nodes A, C, D, E and I reach `q = 3`; the rest stop at 2.
Those five sit inside the most heavily-overlapping part of the complex.

### Simplicial degree

`simplicial_degree()` is the higher-order analogue of node degree: not "how
many neighbours" but "how many groups of each size am I in?"

```{r q-analysis-degree}
simplicial_degree(sc)
```

`d0` is always 1 — every node is its own vertex. `d1` is ordinary graph
degree. **`d2` and `d3` are where the higher-order information lives**: E is
in 7 triangles and 2 tetrahedra, F in only 2 triangles and no tetrahedra,
though their graph degrees (6 and 4) are not nearly so far apart.

The table is sorted by `total`, so the ranking is immediate.

#### Ordinary degree does not predict it

```{r q-analysis-vsdegree}
d <- simplicial_degree(sc)
data.frame(
  measure = c("d1 (graph degree) vs total", "d2 (triangles) vs total"),
  spearman = c(round(cor(d$d1, d$total, method = "spearman"), 3),
               round(cor(d$d2, d$total, method = "spearman"), 3))
)
```

Graph degree correlates with higher-order degree, but imperfectly — the
ranking by `total` is not the ranking by `d1`. A node can have many
neighbours who are not connected to *each other*, which produces edges
without triangles.

#### Normalising

```{r q-analysis-norm}
simplicial_degree(sc, normalized = TRUE, top = 4)
```

`normalized = TRUE` divides each count by the maximum possible at that
dimension, so values are comparable across dimensions and across complexes of
different sizes. Use it when comparing datasets; use the raw counts when
describing one.

### On real data

```{r q-analysis-real}
seqs <- split(human_long$code, human_long$session_id)
hon_3 <- build_hon(seqs, max_order = 3, min_freq = 5)
sc_hon <- build_simplicial(hon_3, type = "pathway")
q_hon <- q_analysis(sc_hon)
as.data.frame(q_hon)
simplicial_degree(sc_hon)
```

The pathway complex of the coding sessions behaves very differently from the
random network above: it stays in many components down to `q = 2` and only
consolidates at `q = 1`. And in the degree table, `d0` and `d1` are
**identical for every action** — a consequence of the saturated first-order
structure, where every pair of actions co-occurs. All the discrimination is
in `d2` and `d3`, where Specify (32 tetrahedra) separates sharply from Verify
(8).

When `d1` is constant, ordinary degree centrality is uninformative by
construction, and the higher-order degree is the only thing distinguishing
the nodes.

### What to do next

- **[`betti_numbers()`](simplicial-complexes.html#betti_numbers)** — holes rather than overlap;
  the other topological summary of the same complex.
- **[`build_simplicial()`](simplicial-complexes.html#build_simplicial)** — how the complex is
  built, and how sensitive it is to `threshold`.
- **[`hypergraph_measures()`](hypergraphs.html#hypergraph_measures)** — overlap
  statistics without the subset-closure assumption.

### References

- Atkin, R. H. (1974). *Mathematical Structure in Human Affairs*. Heinemann.
- Johnson, J. (2013). *Hypernetworks in the Science of Complex Systems*.
  Imperial College Press.
- Estrada, E., & Ross, G. J. (2018). Centralities in simplicial complexes.
  *Journal of Theoretical Biology*, 438, 46-60.

