---
title: "Plot review: plot.net_outcome()"
date: "2026-09-28"
output:
  html_document:
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                      comment = "", fig.width = 8, fig.height = 4.8,
                      out.width = "100%", dpi = 130)
library(hypernets)
```

## What is being compared with what

`hon_outcome()` gives every **actor** (here: a student) one number per
feature. The number is an average: take every higher-order state the student
passed through (for example "`debug -> code`", the state *code* reached from
*debug*), look up that state's centrality in the pooled network, and average
over the student's visits. A student whose trajectory keeps returning to
central states gets a high value; one who spends time in peripheral states
gets a low one.

The regression then compares **students with each other**: do students with
a higher average exposure to central states have a different outcome? Nothing
is compared within a student, and no single node or motif is being tested.

Each section shows two views of the same fit:

1. **Per-SD effects** (`plot(fit)`, the default): how much the outcome
   differs between two students one standard deviation apart on that feature,
   with the other features held fixed. All features sit on the same axis, so
   they can be compared directly.
2. **Contrasts** (`plot(fit, what = "contrasts")`): two hypothetical
   students per feature, one at the 25th and one at the 75th percentile of
   that feature, every other feature at its average; the plot shows the
   predicted outcome for each.

```{r describers, include = FALSE}
# Sentence builders for the inline prose. They read the tidy tables returned
# by the accessors; all indexing stays inside these function bodies.
fmt <- function(v, digits = 2L) format(signif(v, digits))
num <- function(v) formatC(v, format = "f", digits = 2L)

describe_per_sd <- function(fit, who, verb_gauss = "score",
                            odds_of = "the outcome") {
  d <- as.data.frame(fit)
  d <- d[order(d$p_adj, d$feature), , drop = FALSE]
  lab <- hypernets:::.hoo_feature_label(d$feature)
  lines <- vapply(seq_len(nrow(d)), function(i) {
    ci <- sprintf("95%% CI %s to %s; BH-adjusted p = %s",
                  num(d$per_sd_low[i]), num(d$per_sd_high[i]),
                  fmt(d$p_adj[i]))
    effect <- if (identical(fit$family, "gaussian")) {
      sprintf("%s %s %s", verb_gauss, num(abs(d$effect_per_sd[i])),
              if (d$effect_per_sd[i] < 0) "points lower" else "points higher")
    } else {
      sprintf("have %s times the %s", num(d$effect_per_sd[i]),
              if (identical(fit$family, "binomial"))
                paste("odds of", odds_of) else "expected count")
    }
    sprintf("%s one SD higher in %s (1 SD = %s) %s (%s).", who, lab[i],
            fmt(d$feature_sd[i]), effect, ci)
  }, character(1L))
  paste(lines, collapse = " ")
}

describe_contrasts <- function(fit, who, pct = FALSE) {
  d <- as.data.frame(fit, what = "contrasts")
  lab <- hypernets:::.hoo_feature_label(d$feature)
  show <- if (pct) function(v) sprintf("%s%%", num(100 * v)) else num
  lines <- vapply(seq_len(nrow(d)), function(i) {
    sprintf(paste0("For %s, a %s at the 25th percentile (%s) is predicted ",
                   "at %s (95%% CI %s to %s) and one at the 75th percentile ",
                   "(%s) at %s (%s to %s): a difference of %s (95%% CI %s ",
                   "to %s)."),
            lab[i], who, fmt(d$low_value[i], 3L),
            show(d$predicted_low[i]), show(d$predicted_low_ci_low[i]),
            show(d$predicted_low_ci_high[i]), fmt(d$high_value[i], 3L),
            show(d$predicted_high[i]), show(d$predicted_high_ci_low[i]),
            show(d$predicted_high_ci_high[i]),
            if (pct) sprintf("%s percentage points",
                             num(100 * d$difference[i]))
            else num(d$difference[i]),
            num(if (pct) 100 * d$conf_low[i] else d$conf_low[i]),
            num(if (pct) 100 * d$conf_high[i] else d$conf_high[i]))
  }, character(1L))
  paste(lines, collapse = " ")
}
```

```{r simulators}
simulate_students <- function(n_classes = 20L, per_class = 6L) {
  class_id <- rep(sprintf("class%02d", seq_len(n_classes)), each = per_class)
  n <- length(class_id)
  class_tilt <- rep(stats::rnorm(n_classes, sd = 0.8), each = per_class)
  class_effect <- rep(stats::rnorm(n_classes, sd = 3), each = per_class)
  debug_share <- stats::plogis(-0.5 + class_tilt + stats::rnorm(n, sd = 0.8))
  data.frame(
    student = sprintf("s%03d", seq_len(n)), class = class_id,
    debug_share = debug_share,
    score = 70 - 12 * debug_share + class_effect + stats::rnorm(n, sd = 3),
    completed = stats::rbinom(n, 1L, stats::plogis(2 - 4 * debug_share)),
    stringsAsFactors = FALSE)
}
simulate_logs <- function(students, n_motifs = 12L) {
  motifs <- list(plan = c("plan", "code", "test"),
                 debug = c("debug", "code", "debug"),
                 read = c("read", "plan", "code"),
                 review = c("test", "read", "plan"))
  rows <- lapply(seq_len(nrow(students)), \(i) {
    p <- students$debug_share[i]
    pick <- sample(names(motifs), n_motifs, replace = TRUE,
                   prob = c((1 - p) / 3, p, (1 - p) / 3, (1 - p) / 3))
    actions <- unlist(motifs[pick], use.names = FALSE)
    data.frame(student = students$student[i], step = seq_along(actions),
               action = actions, stringsAsFactors = FALSE)
  })
  do.call(rbind, rows)
}
set.seed(2026)
students <- simulate_students()
logs <- simulate_logs(students)
hon <- build_hon(logs, action = "action", actor = "student", time = "step",
                 max_order = 2L)
```

## 1. Gaussian score, two centralities, CR3 over 20 classes

```{r gaussian-two}
fit_two <- hon_outcome(hon, outcome = students, sequences = logs,
                       by = "student", outcome_col = "score",
                       action = "action", time = "step",
                       centrality_type = c("pagerank", "betweenness"),
                       nested_in = "class")
plot(fit_two)
```

**Reading it.** `r describe_per_sd(fit_two, "Students", "score")` The two
features are adjusted for each other, and the intervals are cluster-robust
because students in the same class are not independent.

```{r gaussian-two-contrasts}
plot(fit_two, what = "contrasts")
```

`r describe_contrasts(fit_two, "student")` The grey bar is the gap between the
two hypothetical students; when the two intervals overlap heavily the gap is
not distinguishable from zero.

```{r gaussian-two-table}
as.data.frame(fit_two)
as.data.frame(fit_two, what = "contrasts")
```

## 2. Same fit, model-based covariance (no nesting)

```{r gaussian-naive}
fit_naive <- hon_outcome(hon, outcome = students, sequences = logs,
                         by = "student", outcome_col = "score",
                         action = "action", time = "step",
                         centrality_type = c("pagerank", "betweenness"))
plot(fit_naive)
```

**Reading it.** `r describe_per_sd(fit_naive, "Students", "score")` The point
estimates are the same as in section 1; only the intervals change. These
treat the 120 students as independent, which the class structure makes
false, so they are too narrow.

```{r gaussian-naive-contrasts}
plot(fit_naive, what = "contrasts")
```

`r describe_contrasts(fit_naive, "student")`

## 3. Gaussian, all five features (near-collinear)

```{r gaussian-all}
fit_all <- hon_outcome(hon, outcome = students, sequences = logs,
                       by = "student", outcome_col = "score",
                       action = "action", time = "step",
                       features = c("centrality", "honem"),
                       nested_in = "class")
plot(fit_all, what = "per_sd")
```

**Reading it.** Five related features compete for the same signal, so each
per-SD effect is "holding the other four fixed" -- a comparison between
students who differ on one feature but not on the others, which barely exist
in these data. The wide intervals are that collinearity (scaled condition
index `r num(fit_all$condition_index)`), not an absence of signal.

```{r gaussian-all-contrasts, fig.height = 6}
plot(fit_all, what = "contrasts")
```

`r describe_contrasts(fit_all, "student")`

## 4. Binomial completion, CR3 over 20 classes

```{r binomial}
fit_completed <- hon_outcome(hon, outcome = students, sequences = logs,
                             by = "student", outcome_col = "completed",
                             action = "action", time = "step",
                             centrality_type = c("pagerank", "betweenness"),
                             nested_in = "class", family = "binomial")
plot(fit_completed)
```

**Reading it.** For a 0/1 outcome the effect is an odds ratio on a log
axis; 1 means no association. `r describe_per_sd(fit_completed, "Students", odds_of = "completing")`

```{r binomial-contrasts}
plot(fit_completed, what = "contrasts")
```

The contrast turns the odds ratio back into probabilities of completing.
`r describe_contrasts(fit_completed, "student", pct = TRUE)`

## 5. Poisson on real sessions, CR3 over 34 projects

```{r poisson}
session_messages <- aggregate(cbind(n_messages = message_id) ~
                                session_id + project,
                              data = human_long,
                              FUN = \(m) length(unique(m)))
session_hon <- build_hon(human_long, action = "code", actor = "session_id",
                         time = "order_in_session", max_order = 2L,
                         min_freq = 50L)
session_fit <- hon_outcome(session_hon, outcome = session_messages,
                           sequences = human_long, by = "session_id",
                           outcome_col = "n_messages", action = "code",
                           time = "order_in_session",
                           centrality_type = c("pagerank", "betweenness",
                                               "closeness"),
                           nested_in = "project", family = "poisson")
plot(session_fit)
```

**Reading it.** Here the actors are sessions and the outcome is the number
of human messages, so the effect is a rate ratio (log axis, 1 = no
association). `r describe_per_sd(session_fit, "Sessions")` This is a
mechanics demonstration: outcome and features come from the same sessions.

```{r poisson-contrasts}
plot(session_fit, what = "contrasts")
```

`r describe_contrasts(session_fit, "session")`

## 6. The old view, kept as `what = "raw"`

```{r raw}
plot(fit_two, what = "raw")
```
